### 2022-03-15 preregistration/details_comparison.Rmd
Initial commit

---
title: "Additional information to the preregistration for 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'"
author: Simon Nolte
output: 
  pdf_document:
    latex_engine: xelatex
mainfont: Arial
urlcolor: blue
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  include = TRUE
  )
```

# Part II: Experimental Comparison

This document includes the additional information accompanying the preregistration listed in the file `preregistration.pdf`. It includes example scripts and detailed workflows of the planed analysis steps.

This work is licensed under a [CC-BY 4.0](http://creativecommons.org/licenses/by/4.0/) license.

```{r packages}
# Packages used for the data workflow

library(purrr)
library(spiro)
library(tidyr)
library(ggplot2)
library(scales)
library(MetBrewer)
```

## Data Source

A total of n = 76 data sets will be used for analysis. The exercise tests were conducted during two studies and had the exact same methodology [@quittmann_unpubl_a; @quittmann_unpubl_b] . All individual data files will be saved in a `data` folder.

```{r data}
files <- list.files("data/", full.names = TRUE)
```

## Using different processing strategies on the data

The most common data processing strategies will be tested on the data for a wide range of parameter inputs. This will happen using the `spiro` package [@nolte2022].

I tested this exemplary code for two data files unrelated to the work presented here (two constant load tests) and for the exemplary strategies of breath- and time-based moving averages for the parameter range 5-60.

The resulting matrix (rows: data files, columns: [processing strategy_parameter]) will be saved as an `csv` file.

```{r}
vo2max <- function(parameter, method, data) {
  if (method == "ba") {
    parameter_exp <- paste0(parameter, "b")
  } else {
    parameter_exp <- parameter
  }
  max_values <- spiro_max(data, parameter_exp)
  out <- data.frame(
    name = max_values$VO2_rel
  )
  colnames(out) <- paste(method, parameter, sep = "_")
  out
}

# level 1: iterate over each data file
results <- purrr::map_dfr(
  .x = files,
  .f = function(file) {
    # level 2: iterate over each processing method
    data <- spiro(file)
    purrr::map_dfc(
      .x = c("ta", "ba"), 
      .f = function(method, data) {
        # level 3: iterate over each parameter range
        purrr::map_dfc(
          .x = seq.int(5,60),
          .f = vo2max,
          method = method,
          data = data
        )
      }, 
      data = data
    )
  }
)

write.csv(results, "results.csv")
```

## Normalize results

I now normalize the results to a reference procedure (the most common according to the review). For illustration and code testing, I here use the 10 second breath-averaging as the reference procedure.

```{r normalize}
results_norm <- results / results$ba_10
results_norm$id <- as.factor(seq_len(nrow(results_norm)))
```

## Plot data

I will plot the effect of data averaging strategy for varying parameter on V̇O~2max~ values.

### Turn data in tidy format

To plot the data using `ggplot` I have to bring it into tidy format first.

```{r tidy}
results_tidy <- pivot_longer(
  data = results_norm, 
  cols = !id,
  names_to = c("type", "parameter"),
  names_pattern = "(.*)_(.*)",
  names_transform = list(
    parameter = ~ as.numeric(.x),
    type = ~ factor(.x, 
      levels = c("ta", "ba"), 
      labels = c("time average", "breath average")
    )
  )
)

# calculate normalized results as percent difference
per_diff <- (results_tidy$value - 1) / 1
```

### Create plot

This is an exemplary plot to illustrate the supposed concept of visualisation. The final plot will rather display the median line with the 2.5 and 97.5 quantiles. Other processing strategies involving parameters that cannot be displayed on the given x-scale may be incorporated into the plot using appropriate techniques.

```{r}
ggplot(results_tidy, aes(x = parameter, y = per_diff, colour = type, fill = type)) +
  geom_hline(
    yintercept = 0,
    colour = "grey",
    alpha = 0.15
  ) +
  stat_summary(geom = "line", fun = "mean") +
  stat_summary(
    geom = "ribbon", 
    fun.min = "min", 
    fun.max = "max", 
    alpha = 0.15, 
    colour = NA
  ) +
  scale_x_continuous(
    name = "Parameter",
    breaks = c(15, 30, 45, 60)
  ) +
  scale_y_continuous(
    name = "% to Reference",
    labels = scales::percent,
    limits = c(-0.07,0.1)
  ) +
  scale_color_manual(
    values = met.brewer("OKeeffe1", 2)
  ) +
  scale_fill_manual(
    values = met.brewer("OKeeffe1", 2)
  ) +
  theme_classic() +
  theme(
    legend.position = c(0.8,0.8)
  )

ggsave("example_plot.png", width = 6, height = 4, dpi = 300, bg = "white")
```

![Proof of concept for a plot that illustrates differences between different data processing strategies](example_plot.png)

## References

### 2022-03-15 preregistration/details_review.Rmd
Initial commit

---
title: "Additional information to the preregistration for 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'"
author: Simon Nolte
output: 
  pdf_document:
    latex_engine: xelatex
mainfont: Arial
urlcolor: blue
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  include = TRUE
  )
```

# Part I: Review

This document includes the additional information accompanying the preregistration listed in the file `preregistration.pdf`. It includes example scripts and detailed workflows of the planed analysis steps.

This work is licensed under a [CC-BY 4.0](http://creativecommons.org/licenses/by/4.0/) license.

```{r packages}
# Packages used for the data workflow

library(readxl)
library(rentrez)
library(XML)
library(purrr)
library(ggplot2)
library(MetBrewer)
```

## Data retrieval

### Read data

Data will be searched from PubMed and Web of Science using the search terms indicated in the preregistration. Search results will be downloaded as a `csv` files (for PubMed entries) and as separate `xls` files (for Web of Science, which has a download limit of 1000 entries). The `xls` files will be manually merged. I developed the following data workflow on a small searching sample with search terms different to the ones of the present study.

```{r read}
# file name for PubMed results
pm_file <- NA

# file name for merged WoS results
wos_file <- NA

# read PubMed data
pm_data_pre <- read.csv(pm_file)
pm_data <- data.frame(
  pmid = pm_data_pre[, 1],
  title = pm_data_pre$Title,
  authors = pm_data_pre$Authors,
  journal = pm_data_pre$Journal.Book,
  year = pm_data_pre$Publication.Year,
  doi = pm_data_pre$DOI,
  abstract = NA
)
# read Web of Science data
wos_data_pre <- readxl::read_xls(wos_file)
wos_data <- data.frame(
  pmid = wos_data_pre$`Pubmed Id`,
  title = wos_data_pre$`Article Title`,
  authors = wos_data_pre$`Author Full Names`,
  journal = wos_data_pre$`Source Title`,
  year = wos_data_pre$`Publication Year`,
  doi = wos_data_pre$DOI,
  abstract = wos_data_pre$Abstract
)

# count PubMed results
nrow(pm_data)
# count Web of Science results
nrow(wos_data)
```

### Removal of Duplicates

Duplicates will be removed based on their DOI. Before, all entries without a DOI assigned will be excluded. This may remove literature fitting the scope of the review, but as later a random sample of the results is taken, we deem this as acceptable.

```{r duplicates}
# Merge data frame
merge_data <- rbind(wos_data, pm_data)

# Filter entries without doi
no_doi <- is.na(merge_data$doi)
merge_data <- merge_data[!no_doi, ]

# Count removal from missing doi
sum(no_doi)

# Find duplicated DOIs
dupl <- duplicated(merge_data$doi)
# Filter duplicates
merge_data <- merge_data[!dupl, ]

# Count removal of duplicates
sum(dupl)

```

### Automated title filtering

The titles of all non-duplicated articles will be screened to exclude those that indicate, that an article is not presenting original research (e.g. those with the term 'review' in the title). As with the DOI exclusion criteria, this will likely also exclude a minor number of possibly fitting articles, which is again not considered relevant, as later a sample is drawn.

```{r, filter}
# Filter titles by terms
terms <- paste0("review|comment|correction|retraction|meta-analysis",
                "|editorial|erratum|reply")
excl <- grepl(terms, merge_data$title, ignore.case = TRUE)
filtered_data <- merge_data[!excl, ]

# Count removal by title filtering
sum(excl)

```

## Random Sampling

Based on the included search results, a random sample of 500 articles will be drawn. According to the preregistration, the sample may be later increased in steps of 100 articles if a set threshold of finally included articles is not meet by the initial sample (see `preregistration.pdf` for more details). If the sample is later extended, this will happen by using the same random seed.

For all sampled articles, abstract data will be retrieved. For search results from Web of Science, this data is already included in the search results. For PubMed searches this is not the case, so the abstracts are automatically retrieved by using the article PMID on the PubMed API.

```{r sampling}
# Assign ids to articles left
filtered_data$id <- seq_len(nrow(filtered_data))

# randomly sample 500 articles
set.seed(4711)
smpl <- sample(filtered_data$id, size = 50)

sampled_data <- filtered_data[smpl, ]
# assign sample id
sampled_data$sid <- seq_len(nrow(sampled_data))

# get abstract from PubMed where missing

no_abstract <- is.na(sampled_data$abstract)

no_abstract_pmid <- sampled_data$pmid[no_abstract]

# function to scrape abstract from the PubMed API
get_abstract <- function(pmid) {
  xml_data <- rentrez::entrez_fetch(
    db = "pubmed", 
    id = pmid, 
    rettype = "xml", 
    parse = TRUE
  )
  text <- XML::xmlValue(XML::getNodeSet(xml_data, "//AbstractText"))
  out <- paste(text, collapse = "///")
  if (out == "") out <- NA
  out
}

# retrieve missing abstracts
sampled_data$abstract[no_abstract] <- purrr::map_chr(
  .x = no_abstract_pmid, 
  .f = get_abstract
)

```

## Screening

### Prepare Title-Abstract plots for screening

To simplify blinded title and abstract screening, I will create a plot for every sampled article, that displays title, abstract and sampling ids. The texts in the plots will have different (randomly assigned) colours, to increase the screener's focus during the work. If the abstract plot remains empty (i.e. no abstract data could be automatically retrieved), the abstract will be searched online, so in this case the screener will be unblinded.

```{r}
# Function for creating plots with blinded abstract data
create_abstract_plot <- function(row, data) {
  # get random colour
  # different colours in plots are used to stay concentrated when scanning
  # the abstracts
  clrs <- c(MetBrewer::met.brewer("Moreau")[-4])
  clr <- sample(clrs, 1)

  g <- ggplot(data = NULL, aes(x = 0, y = 1)) +
  ggtext::geom_textbox(
    width = 1, 
    box.colour = "transparent", 
    label =  gsub(">", "", data$abstract[row]),
    colour = clr
  ) +
  labs(
    title = paste0(strwrap(data$title[row], 80), collapse = "\n"), 
    subtitle = paste0("Sampling ID: ", sprintf("%03i", data$sid[row])),
    caption = paste0("ID: ", a$id, "; PMID = ", data$pmid[row])
    ) +
  theme_void()

  ggsave(
    filename = paste0("abstracts/", sprintf("%03i", data$sid[row]), ".png"), 
    plot = g, 
    dpi = 300, width = 7, height = 5, bg = "white"
  )
}

# Create blinded abstract plots
purrr::walk(
  .x = seq_len(nrow(sampled_data)), 
  .f = create_abstract_plot, 
  data = sampled_data
)

```

### Abstract Screening

Abstract screening results are collected in a `screening_extraction.csv` file. Abstract screening results are inserted in an 'abstract_exclusion' column. Based on the exclusion criteria the screener has to answer the following question using the title and abstract data:

1.  Does the abstract indicate, that the full-text is not in English language? (if yes, indicate 'l'; if no, continue)
2.  Is the article no original research? (if yes, indicate 'r'; if no, continue)
3.  Is the article not presenting research in humans? (if yes, indicate 'h'; if no, continue)
4.  Does the abstract indicate, that VO2 was not measured (e.g only estimated, approximated, predicted)? (if yes, indicate 'm'; if no indicate 'NA')

```{r abstract-screen}
# read screening results
incl_data <- read.csv("./screening_extraction.csv")
# assign exclusion criteria as factor levels
incl_data$abstract_exclusion <- factor(
  incl_data$abstract_exclusion, 
  levels = c("l", "r", "h", "m"),
  labels = c("no english full-text", "no original research", 
             "no humans", "VO2 not measured")
)
# count exclusions
summary(incl_data$abstract_exclusion)
```

### Full-text screening

For all articles not excluded during title/abstract scanning the full texts will be downloaded. Results of the full-text screening are inserted in the 'text_exclusion' column of the `screening_extraction.csv` file. Based on the exclusion criteria the screener has to answer the following question using the articles full text:

5.  Was no full-text available for the corresponding article? (if yes, indicate 't', if no continue)
6.  Apply step 1-4 from abstract screening (if all questions are answered with yes, continue)
7.  Was V̇O~2~ not continuously measured (e.g. only in the last seconds of the exercise; only post-exercise)? (if yes, indicate 'c'; if no, continue)
8.  Did the testing protocol not aspire a subjective exhaustion of the participants (e.g. involved only submaximal exercises)? (if yes, indicate 'e'; if no, continue)
9.  Did the testing protocol have a mean duration of more than 20 minutes until exhaustion (e.g. long graded incremental exercise protocols, long time-trials)? (if yes, indicate 'd'; if no, continue)
10. Are there other rationale to exclude the article (if yes, indicate 'o' and state the exact reason in the 'exclusion_note' column; if no indicate 'NA')

```{r text-screening}
# assign exclusion criteria as factor levels
incl_data$text_exclusion <- factor(
  incl_data$text_exclusion, 
  levels = c("t", "l", "r", "h", "m", "c", "e", "d", "o"),
  labels = c("no fulltext", "no english full-text", "no original research", 
             "no humans", "VO2 not measured", "VO2 not continously measured", 
             "no exhaustion", "long protocol", "other")
)
# count exclusions
summary(incl_data$text_exclusion[is.na(incl_data$abstract_exclusion)])
```

## Data Extraction

Data will be extracted from all articles not excluded in the previous steps. If one of the question cannot be answered based on the full-text, this will be indicated by 'NA' in the corresponding column. If an article references to other studies or supplementary material for further information on the methods, these additional materials will be assessed for data extraction.

The screener has to answer the following question during the data extraction process and has to fill the answer into the corresponding column of the `screening_extraction.csv` file:

'cart': Which metabolic cart did the authors use? State manufacturer and model.

'type': Which type of data was processed? Either 'bbb' for breath-by-breath data, 'mc' for mixed chamber, or others.

'pre': Which preprocessing algorithm did the researchers use to calculate the breath-by-breath data?

'software': Which software did the researchers use to analyse the gas exchange data? State the name and the version.

'interpolation': Which interpolation procedure did the researchers use to process the raw data? E.g. 'second' for interpolation to full seconds.

'ptype': Which type of data processing did the researcher use when analysing the gas exchange data? Either 'ta' for time-based average, 'ba' for breath-based average, 'bw' for Butterworth filter, or other

'averaging': Which type of averaging did the researchers use, if they used an averaging strategy for data processing? Either 'm' for moving, 'b' for binned, or others.

'parameters': Which parameters did the researchers use for the data processing. E.g. the length of the interval for calculating a breath- or time-based average, or the parameters (order, cut-off frequency) for a digital filter.

### 2022-03-15 preregistration/pregistration.Rmd
Initial commit

---
title: "Preregistration for 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'"
author: Simon Nolte
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Arial
urlcolor: blue
bibliography: ../references.bib
csl: ../apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
```

This document includes the preregistration of the aforementioned project. It follows the 'Inclusive Systematic Review Registration Form' [@akker2020]. For more information on the distinct parts of the planed work including analysis code, visit the accompanying file `detail_review.pdf` and `details_comparison.pdf`.

This work is licensed under a [CC-BY 4.0](http://creativecommons.org/licenses/by/4.0/) license.

## Meta-data

### Target discipline

Sport Science; Sports Medicine

### Title

'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'

### Author(s)/Contributor(s)

Simon Nolte^1^, Oliver Jan Quittmann^1^

^1^Institute of Movement and Neurosciences, German Sport University Cologne

### Tasks and Roles

SN: Conceptualization; Data curation; Formal Analysis; Software; Investigation; Methodology; Visualization; Writing

OJQ: Conceptualization, Investigation; Methodology

## Review methods

### Type of Review

Scoping review, following the PRISMA-ScR guidelines [@tricco2018]. Accompanied by an empirical analysis (see `details_comparison.pdf` for more information)

### Review Stages

Search, Automated Screening, Random Sampling, Manual Screening, Extraction

### Current review stage

Before Search; the project has not started yet.

### Start date

05 March 2022

### End date

30 April 2022 (for data extraction)

### Background

As the most relevant physiological determinant of endurance performance, the maximum oxygen uptake (V̇O~2max~) is determined by measuring gas exchange measures during cardiopulmonary exercise testing [@bassett2000]. Different processing routines of the collected raw data lead to different V̇O~2max~ values [@robergs2003].

@robergs2010 aimed to map the current state of data processing strategies in scientific cardiopulmonary exercise testing. They stated that there exist "two potential methods for this content: (i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible", while choosing the latter for their article [@robergs2010]. A systematic review of published research on the data processing methods for V̇O~2max~ determination has yet not been conducted.

Once the most common strategies have been identified in the literature, they can be compared in terms of their influence on determining the V̇O~2max~. @martin-rincon2019 compared an arbitrary chosen set of data processing strategies, but only on a group level in an assorted data set of mostly recreational or untrained individuals. Systematic comparisons on an individual level in well-trained athletes while controlling for external factors (protocol, measurement device, ...) are lacking.

### Primary research question

Which methods for processing of gas exchange data are currently used in the scientific literature when determining the maximum oxygen uptake?

### Secondary research question

How do the commonly used strategies for data processing influence the determined value for the maximum oxygen uptake?

### Expectations / hypotheses

--- No hypothesis tested due to the mapping character of the review ---

--- The empirical part of the project is purely descriptive ---

### Dependent variable(s) / outcome(s) / main variable(s)

Data processing strategy used

### Independent variable(s) / interventions(s) / treatment(s)

--- No independent variables used ---

### Additional variable(s) / covariate(s)

-   Metabolic cart used
-   Data preprocessing algorithm used

### Software

R Version 4.x [@rcoreteam2021]

### Funding

--- No funding was received for this project ---

### Conflicts of interest

--- All researchers involved state that there is no conflict of interest ---

### Overlapping authorship

--- No overlapping authorship is expected ---

## Search Strategy

### Databases

PubMed & Web of Science

### Interfaces

PubMed & Web of Science

### Grey literature

--- No grey literature will be searched ---

### Inclusion and exclusion criteria

Inclusion:

-   only research related to 'maximum oxygen uptake' (or similar term in title, abstract or keywords)
-   only research published in 2017 or later

### Query strings

PubMed: `((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`

Web of Science: `(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`

### Search validation

--- no search validation procedure is conducted, as we aim to only map current methodical practices on a sample of research articles. We do not believe that our search strategy incorporates a systematic bias towards this goal. ---

### Other search strategies

--- No further search strategies are used ---

### Procedures to contact authors

--- We will not contact any authors during the review ---

### Results of contacting authors

--- Not applicable as we do not contact authors ---

### Search expiration and repetition

We will conduct one single literature search in March 2022, which will not be repeated.

### Search strategy justification

We aim to map current methodical practices in a specific yet often employed setting (determining maximum oxygen uptake). As measurement devices and methods change with time, we decided to include only studies from the past five years (2017 to today). We want to keep the initial search quite broad to have a better control of article exclusion in later stages (e.g. by automated title screening instead of additional initial exclusion criteria in search). As we later only take a random sample of our search results, we do not need to exhaustively review the literature, which is why we refrain from searching additional databases and grey literature.

Based on @robergs2010 we use the term V̇O~2max~ to denote the highest oxygen uptake during a short exercise test until subjective exhaustion. Criteria to distinguish between V̇O~2max~ and V̇O~2peak~ are not universally accepted and heavily rely on the method used to process the gas exchange data (e.g. V̇O~2~ plateau as primary criteria or RER~max~ as secondary criteria). We do not use V̇O~2peak~ as a search term, as we expect such search to return many studies using exercise protocols that are generally not considered eligible for determining the maximum oxygen uptake.

We believe that our search strategy provides a reasonable subset of the current scientific literature determining the V̇O~2max~.

## Screening

### Screening stages

1.  Automated screening from Titles and meta data
2.  Automated removal of duplicates (based on DOI)
3.  Random sample of 500 articles
4.  Manual title and abstract screening
5.  Full-text retrieval and screening
6.  Data extraction
7.  *If less than 200 results remain after step 5, we will add an additional random sample of 100 article (step 3). This may be repeated until a minimum of 200 articles remains in step 5.*

### Screened fields / blinding

During manual title and abstract screening, all additional information will be blinded to the person screening. During full text screening no blinding applies. More information are available in the `details_review.pdf` file.

### Used exclusion criteria

During automated screening:

-   No DOI attached
-   Title contains at least one of the following phrases: 'review'; 'correction'; 'meta-analysis'; 'comment'; 'retraction'; 'editorial'; 'erratum'; 'reply'
-   article duplicated (based on DOI assigned)

During manual abstract screening:

-   text not in English language
-   no original research article
-   no research in humans
-   V̇O~2max~ not measured (e.g. only estimated)

During manual text screening:

-   no full text available
-   all of the above criteria (from manual abstract screening)
-   V̇O~2~ not continuously measured
-   testing protocol not to subjective exhaustion
-   testing protocol longer than 20 minutes (on average)

### Screener instructions

--- See document `details_review.pdf` for more details ---

### Screening reliability

All screening will be employed by two independent screeners (SN and OJQ).

### Screening reconciliation procedure

Conflicts during screening will be resolved by discussion between the two screeners.

### Sampling and sampling size

After the automated title and meta data screening and removing of duplicates, we will perform a random sample of 500 articles (see `detail_review.pdf` for technical details).

We believe a final sample size of 200 studies after all exclusion steps to be sufficient, assuming that about half of these lack appropriate reporting of methods, and half contain valuable information. We expect about 50% of the initially sampled articles to be excluded during abstract and full-text screening and therefore use the sampling size of 500.

If after all exclusion steps less than 200 articles remain, we will extend the random sampling by an additional 100 articles (using the same random seed), until the threshold of 200 articles after final exclusion is meet.

### Screening procedure justification

Due to economical and methodological reasons we perform no exhaustive literature search and investigate a sample of articles (see above). All exclusion criteria aim to leave the studies, that measured V̇O~2max~ using an appropriate testing procedure in humans.

### Data management and sharing

We will store the initial search results in a `csv` (for PubMed) and a `xls` (for Web of Science) file. The automated exclusion will be done in R [@rcoreteam2021] and documented in R Markdown. Results of manual screening and data extraction will be documented in a `csv` file. For more information see the file `details_review.pdf`.

All data and analysis code will be made publicly available on GitHub.

### Miscellaneous screening details

--- No further details ---

## Extraction

### Entities to extract

-   metabolic cart used

-   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)

-   preprocessing algorithm

-   data processing software

-   interpolation procedure

-   data processing/determination of V̇O~2max~:

    -   type (time average, breath average, digital filtering)
    -   alignment (rolling, binned, ...)
    -   interval (in sec/breath, parameters for filtering)

### Extraction stages

All extraction happens in a single manual step by one researcher (SN).

### Extractor instructions

See the document `details_review.pdf` for more details.

### Extractor blinding

No blinding is applied during extraction.

### Extraction reliability

Only a single researcher will extract the information

### Extraction reconciliation

--- Not applicable, as extraction will be done by a single researcher ---

### Extraction procedure justification

We aim to extract data on the data processing methods for V̇O~2max~ testing. We also collect related information (e.g. metabolic cart used) as stated above. If information are missing or incomplete we will document this, but not exclude the article.

### Data management and sharing

Extracted data will be saved in a `csv` file and publicly shared via GitHub.

## Synthesis and quality assessment

--- As this is a scoping review, we will not perform any quality assessment of the included research. Instead we will simply count the methodical characteristics from the included research. Note that this scoping review is accompanied by an empirical comparison of different data processing methods (see `details comparison.pdf` for more information). ---

## References

### 2022-03-15 thesis/discussion.Rmd
Initial commit

---
title: ''
output: pdf_document
bibliography: ../references.bib
---

# Discussion

### 2022-03-15 thesis/introduction.Rmd
Initial commit

---
title: ''
output: pdf_document
bibliography: ../references.bib
---

# Introduction

## Background

<!-- VO2max is important -->
<!-- physiological determinants -->
<!-- VO2max testing -->
<!-- Influence factors in testing -->

## Previous Research

<!-- First example in 90s: applied research -->
<!-- Since then: various example -->
<!-- conceptual work: Robergs -->
<!-- Comparison: martin-rincon -->
<!-- Problem: group + individual -->
<!-- Knowledge Gap -->

## Aim

<!-- Gap -> we need this: -->
<!-- overview what is actually done -->
<!-- and how does this impact VO2max assessment -->

<!-- Review: mapping current methodological (+ reporting) state  -->
<!-- Comparison: highlight advantage over previous -->

<!-- help compare results across studies + for individuals -->
<!-- advise for use of methods  -->
<!-- call for improved reporting -->

@bassett2000; @tricco2018

### 2022-03-15 thesis/methods.Rmd
Initial commit

---
title: ''
output: pdf_document
bibliography: ../references.bib
---

# Methods

The work presented in this thesis was preregistered before the start of the project on [OSF]().

<!-- deviations from preregistration -->

## Systematic Scoping Review

<!-- Detailed Aim of review -->

### Search & Screening

To assess the current state of data processing methods for determining V̇O~2max~ in exercise science I systematically screened recent literature. The article search was conducted on XX March 2022 using [PubMed]() and [Web of Science](). The search included articles articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword. Table 1 shows the exact search terms.

```{r terms, echo = FALSE}
out <- data.frame(
  source = c("PubMed", "Web of Science"),
  term = c('((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))', '(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)')
)

knitr::kable(out, 
  col.names = c("Source", "Search String"), 
  caption = "Search strings for the systematic scoping review.")
```

The search results from both data bases were joined at checked for the presence of a DOI. Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step. After the removal of duplicates I conducted an automated title scanning, to remove results that are likely no original research articles. Table 2 displays the exclusion terms during the automated title screening.

```{r titlescan, echo = FALSE}
out <- data.frame(
  term = c("review, correction, meta-analysis, comment, retraction, editorial, erratum, reply")
)

knitr::kable(out, 
  col.names = c("Search terms for exclusion by title"), 
  caption = "Exclusion terms during automated title screening after removal of duplicates. If an articles title matched least one of the given terms, it was excluded.")
```

In accordance with the preregistration a random sample was drawn from the search results. The goal of the random sample was to give an unbiased sample of the current state of scientific V̇O~2max~ testing. Based on the procedure described in the preregistration file (LINK), <!-- random sample size  -->

The abstracts from the articles included in the random sample were blinded for scanning. This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the articles (see Figure 2). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.

After the abstract screening I retrieved the full-texts for all articles remaining in the review. The full-texts were again scanned to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria).

All data exclusion steps are documented under XXXLINK.

### Data Extraction

I retrieved data from all articles remaining after the abstract and full-text screening. Extraction included the following data:

-   metabolic cart used

-   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)

-   preprocessing algorithm

-   data processing software

-   interpolation procedure

-   data processing/determination of V̇O~2max~:

    -   type (time average, breath average, digital filtering)
    -   alignment (rolling, binned, ...)
    -   interval (in sec/breath, parameters for filtering)

All extracted data is available in the Appendix A and under XXXLINK.

### Data Synthesis

## Experimental Comparison

To determine the influence the most common data processing strategies have on the determination V̇O~2max~, I compared them on a set of already collected gas exchange data from running.

### Data Source

A total of N = 76 exercise test were analysed for this study. The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times). The V̇O~2max~ testings were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol. Participants run on a treadmill (XXX) for ten minutes at a velocity of 2.8 m/s as a warm-up. After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increases of 0.15 m/s every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

Gas exchange data were recorded using a ZAN XXX device (XXX). The device was calibrated with a 3l-syringe pump (XXX) and a reference gas (XXX) before each measurement. The measured breath-by-breath data is available at XXXLINK.

### Data Processing

The spiro Package for R [@nolte2022a] processed the raw gas exchange data. It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data. Table 4 displays a brief summary exemplary function parameters used for the analysis. The full analysis script is available at XXXLINK.

#### Time based averaging

#### Breath based averaging

#### Butterworth filtering

### Comparison of methods

To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level I calculated the median and 2.5 and 97.5-quantiles of each processing strategy.

### 2022-03-15 thesis/results.Rmd
Initial commit

--- 2022-03-15 thesis/methods.Rmd
@@
 
-# Methods
-
-The work presented in this thesis was preregistered before the start of the project on [OSF]().
-
-<!-- deviations from preregistration -->
+# Results
 
@@
 
-<!-- Detailed Aim of review -->
-
-### Search & Screening
-
-To assess the current state of data processing methods for determining V̇O~2max~ in exercise science I systematically screened recent literature.
-The article search was conducted on XX March 2022 using [PubMed]() and [Web of Science]().
-The search included articles articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword.
-Table 1 shows the exact search terms.
-
-```{r terms, echo = FALSE}
-out <- data.frame(
-  source = c("PubMed", "Web of Science"),
-  term = c('((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))', '(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)')
-)
-
-knitr::kable(out, 
-  col.names = c("Source", "Search String"), 
-  caption = "Search strings for the systematic scoping review.")
-```
-
-The search results from both data bases were joined at checked for the presence of a DOI.
-Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step.
-After the removal of duplicates I conducted an automated title scanning, to remove results that are likely no original research articles.
-Table 2 displays the exclusion terms during the automated title screening.
-
-```{r titlescan, echo = FALSE}
-out <- data.frame(
-  term = c("review, correction, meta-analysis, comment, retraction, editorial, erratum, reply")
-)
-
-knitr::kable(out, 
-  col.names = c("Search terms for exclusion by title"), 
-  caption = "Exclusion terms during automated title screening after removal of duplicates. If an articles title matched least one of the given terms, it was excluded.")
-```
-
-In accordance with the preregistration a random sample was drawn from the search results.
-The goal of the random sample was to give an unbiased sample of the current state of scientific V̇O~2max~ testing.
-Based on the procedure described in the preregistration file (LINK), <!-- random sample size  -->
-
-The abstracts from the articles included in the random sample were blinded for scanning.
-This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the articles (see Figure 2).
-Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
-
-After the abstract screening I retrieved the full-texts for all articles remaining in the review.
-The full-texts were again scanned to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria).
-
-All data exclusion steps are documented under XXXLINK.
-
-### Data Extraction
-
-I retrieved data from all articles remaining after the abstract and full-text screening. Extraction included the following data:
-
--   metabolic cart used
-
--   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)
-
--   preprocessing algorithm
-
--   data processing software
-
--   interpolation procedure
-
--   data processing/determination of V̇O~2max~:
-
-    -   type (time average, breath average, digital filtering)
-    -   alignment (rolling, binned, ...)
-    -   interval (in sec/breath, parameters for filtering)
-
-All extracted data is available in the Appendix A and under XXXLINK.
-
-### Data Synthesis
-
 ## Experimental Comparison
 
-To determine the influence the most common data processing strategies have on the determination V̇O~2max~, I compared them on a set of already collected gas exchange data from running.
-
-### Data Source
-
-A total of N = 76 exercise test were analysed for this study.
-The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
-The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times).
-The V̇O~2max~ testings were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol.
-Participants run on a treadmill (XXX) for ten minutes at a velocity of 2.8 m/s as a warm-up.
-After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increases of 0.15 m/s every 30 seconds.
-The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.
-
-Gas exchange data were recorded using a ZAN XXX device (XXX).
-The device was calibrated with a 3l-syringe pump (XXX) and a reference gas (XXX) before each measurement.
-The measured breath-by-breath data is available at XXXLINK.
-
-### Data Processing
-
-The spiro Package for R [@nolte2022a] processed the raw gas exchange data.
-It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data.
-Table 4 displays a brief summary exemplary function parameters used for the analysis.
-The full analysis script is available at XXXLINK.
-
-#### Time based averaging
-
-#### Breath based averaging
-
-#### Butterworth filtering
-
-### Comparison of methods
-
-To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure.
-The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review.
-Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method.
-I calculated the data for all integer parameter values within the range of the values found in the literature during the review.
-On a group level I calculated the median and 2.5 and 97.5-quantiles of each processing strategy.
+![A plot](../preregistration/example_plot.png)

### 2022-03-17 scripts/review.Rmd
Start review

# Analysis Script for the Systematic Review

## Project: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'

This script closely follows the preregistration, which is uploaded on OSF.

```{r setup}
# Packages used for the data workflow

library(readxl)
library(rentrez)
library(XML)
library(purrr)
library(ggplot2)
library(MetBrewer)
```

#### Get search data

The search was conducted on the 16th March 2022 at around 10 pm CET.

The Pubmed results are saved in the file 'pubmed.csv'; the Web of Science results are saved in the file 'wos.xls'.

```{r read, warning=FALSE, message=FALSE}
# file name for PubMed results
pm_file <- "../data/search/pubmed.csv"

# file name for merged WoS results
wos_file <- "../data/search/wos.xls"

# read PubMed data
pm_data_pre <- read.csv(pm_file)
pm_data <- data.frame(
  pmid = pm_data_pre[, 1],
  title = pm_data_pre$Title,
  authors = pm_data_pre$Authors,
  journal = pm_data_pre$Journal.Book,
  year = pm_data_pre$Publication.Year,
  doi = pm_data_pre$DOI,
  abstract = NA
)
# read Web of Science data
wos_data_pre <- readxl::read_xls(wos_file)
wos_data <- data.frame(
  pmid = wos_data_pre$`Pubmed Id`,
  title = wos_data_pre$`Article Title`,
  authors = wos_data_pre$`Author Full Names`,
  journal = wos_data_pre$`Source Title`,
  year = wos_data_pre$`Publication Year`,
  doi = wos_data_pre$DOI,
  abstract = wos_data_pre$Abstract
)

# count PubMed results
nrow(pm_data)
# count Web of Science results
nrow(wos_data)
```

#### Removal of duplicates

```{r duplicates}
# Merge data frame
merge_data <- rbind(wos_data, pm_data)

# Filter entries without doi
no_doi <- is.na(merge_data$doi) | (merge_data$doi == "")
merge_data <- merge_data[!no_doi, ]

# Count removal from missing doi
sum(no_doi)

# Find duplicated DOIs
dupl <- duplicated(merge_data$doi)
# Filter duplicates
merge_data <- merge_data[!dupl, ]

# Count removal of duplicates
sum(dupl)
```

#### Automated title filtering

```{r filter}
# Filter titles by terms
terms <- paste0("review|comment|correction|retraction|meta-analysis",
                "|editorial|erratum|reply")
excl <- grepl(terms, merge_data$title, ignore.case = TRUE)
filtered_data <- merge_data[!excl, ]

# Count removal by title filtering
sum(excl)
# count remaining articles
nrow(filtered_data)
```

#### Random Sampling

```{r sampling, eval=FALSE}
# Assign ids to articles left
filtered_data$id <- seq_len(nrow(filtered_data))

# randomly sample 500 articles
set.seed(4711)
smpl <- sample(filtered_data$id, size = 500)

sampled_data <- filtered_data[smpl, ]
# assign sample id
sampled_data$sid <- seq_len(nrow(sampled_data))

# get abstract from PubMed where missing

no_abstract <- is.na(sampled_data$abstract)

no_abstract_pmid <- sampled_data$pmid[no_abstract]

# function to scrape abstract from the PubMed API
get_abstract <- function(pmid) {
  if (is.na(pmid))
    return(NA)
  xml_data <- rentrez::entrez_fetch(
    db = "pubmed", 
    id = pmid, 
    rettype = "xml", 
    parse = TRUE
  )
  text <- XML::xmlValue(XML::getNodeSet(xml_data, "//AbstractText"))
  out <- paste(text, collapse = "///")
  if (out == "") out <- NA
  out
}

# retrieve missing abstracts
sampled_data$abstract[no_abstract] <- purrr::map_chr(
  .x = no_abstract_pmid, 
  .f = get_abstract
)

# manually add abstract for articles, where automated abstract retrieval did not
# work. In these cases no blinding applies to the principal investigator
which(is.na(sampled_data$abstract))

sampled_data[500,]

sampled_data$abstract[50] <- readtext::readtext("../data/search/a50.txt")$text
sampled_data$abstract[238] <- readtext::readtext("../data/search/a238.txt")$text
sampled_data$abstract[288] <- readtext::readtext("../data/search/a288.txt")$text
sampled_data$abstract[416] <- readtext::readtext("../data/search/a416.txt")$text
sampled_data$abstract[488] <- readtext::readtext("../data/search/a488.txt")$text
sampled_data$abstract[490] <- readtext::readtext("../data/search/a490.txt")$text
sampled_data$abstract[500] <- readtext::readtext("../data/search/a500.txt")$text

# later plotting throws an error for some articles due to unrecognized html-tags
# in the abstracts. These abstract are replaced by their text equivalents
sampled_data$abstract[344] <- readtext::readtext("../data/search/a344.txt")$text
sampled_data$abstract[356] <- readtext::readtext("../data/search/a356.txt")$text

save(sampled_data, file = "../data/screen/sample.Rda")
```

#### Prepare Title-Abstract plots for screening

```{r abstractplots, eval=FALSE}
load("../data/screen/sample.Rda")

# Function for creating plots with blinded abstract data
create_abstract_plot <- function(row, data) {
  # get random colour
  # different colours in plots are used to stay concentrated when scanning
  # the abstracts
  clrs <- c(MetBrewer::met.brewer("Moreau")[-4])
  clr <- sample(clrs, 1)

  g <- ggplot(data = NULL, aes(x = 0, y = 1)) +
  ggtext::geom_textbox(
    width = 1, 
    box.colour = "transparent", 
    label =  gsub(">", "", data$abstract[row]),
    colour = clr
  ) +
  labs(
    title = paste0(strwrap(data$title[row], 80), collapse = "\n"), 
    subtitle = paste0("Sampling ID: ", sprintf("%03i", data$sid[row])),
    caption = paste0("ID: ", data$id[row], "; PMID = ", data$pmid[row])
    ) +
  theme_void()

  ggsave(
    filename = paste0("../plots/abstracts/", sprintf("%03i", data$sid[row]), ".png"), 
    plot = g, 
    dpi = 300, width = 7, height = 5, bg = "white"
  )
}

# Create blinded abstract plots
purrr::walk(
  .x = seq_len(nrow(sampled_data)),
  .f = create_abstract_plot, 
  data = sampled_data
)

```

### 2022-03-17 transparent_changes.Rmd
Add transparent changes document

---
title: "Transparent Changes"
author: Simon Nolte
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Arial
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
```

# Supplement to the project 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'

This document includes all deviation and modifications of code and methods in the final project compared to its [preregistration](https://osf.io/3am4s).

## Minor Changes

#### Code changes for automated article filtering and screening preparation

* advanced detection of missing DOIs: `is.na(merge_data$doi) | (merge_data$doi == "")` instead of `is.na(merge_data$doi)`.
* Improved function to retrieve missing PubMed abstracts: Handles case when input (PMID) is missing (`if (is.na(pmid)) return(NA)`).
* save/load of the sampling results as an .Rda file to reduce computation time when working on parts of the workflow.

#### Unblinding of single abstract

* Manual retrieval of abstracts for articles, as these were neither present in the search result data, nor could be automatically scraped. Abstracts were saved and imported as .txt files. Temporary unblinding only applied to the primary researcher. This concerns the abstract with the Sampling ID (sid): 50, 238, 288, 416, 488, 490, 500
* Manual retrieval of abstracts for articles, as the automatically collected abstract contained html-tags that could not be removed for later abstract plots. Abstracts were saved and imported as .txt files. Temporary unblinding only applied to the primary researcher. This concerns the abstract with the Sampling ID (sid): 344, 356
* Unblinding during screening to assess the implications of title given in squared brackets. This concerns the abstract with the Sampling ID (sid): 262, 303
* Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the Sampling ID (sid): 275

### 2022-04-02 scripts/review.Rmd
Analyse abstract screen, prepare fulltext screen

@@
 
-sampled_data[500,]
-
 sampled_data$abstract[50] <- readtext::readtext("../data/search/a50.txt")$text
@@
 ```
+
+#### Compare abstract screening results
+
+```{r abstract-screen}
+load("../data/screen/sample.Rda")
+# read screening results
+s_res1 <- read.csv("../data/screen/abstract_screening_i1.csv")[, 2]
+s_res2 <- read.csv("../data/screen/abstract_screening_i2.csv")[, 2]
+
+# compare results
+comp_res <- function(x1, x2) {
+  if (is.na(x1)) x1c <- 0 else x1c <- x1
+  if (is.na(x2)) x2c <- 0 else x2c <- x2
+  if (x1c == x2c) out <- x1 else out <- "XXX"
+  out
+}
+
+s_res <- purrr::map2_chr(s_res1, s_res2, comp_res)
+s_res_df <- data.frame(
+  id = 1:500,
+  i1 = s_res1,
+  i2 = s_res2,
+  result = s_res,
+  decision = rep.int("", 500),
+  doi = sampled_data$doi
+)
+# write.csv(s_res_df, "../data/screen/abstract_screening.csv")
+# Calculate relative agreement
+1 - (sum(s_res == "XXX", na.rm = TRUE) / length(s_res))
+
+# Get final decisions
+s_res_final <- read.csv("../data/screen/abstract_screening.csv")
+# number excluded 
+sum(!is.na(s_res_final$decision))
+# extract remaining data
+ft_screen <- sampled_data[is.na(s_res_final$decision), ]
+
+l <- list.files("../data/fulltext")
+get_ft_id <- function(n, files) {
+  name <- files[n]
+  as.numeric(strsplit(name, "\\.")[[1]][[1]])
+}
+fts <- purrr::map_dbl(seq_along(l), get_ft_id, files = l)
+# find missing full texts
+# ft_screen$sid[!ft_screen$sid %in% fts]
+
+# write file for full text screening
+ft_screen_df <- data.frame(
+  sid = ft_screen$sid,
+  fulltext_exclusion = rep.int("", nrow(ft_screen)),
+  exclusion_note = rep.int("", nrow(ft_screen)),
+  doi = ft_screen$doi
+)
+# write.csv(ft_screen_df, file = "../data/screen/fulltext_screening_i1.csv", row.names = FALSE)
+```

### 2022-04-15 scripts/review.Rmd
Improve review script

@@
 # count PubMed results
-nrow(pm_data)
+pm_count <- nrow(pm_data)
 # count Web of Science results
-nrow(wos_data)
+wos_count <- nrow(wos_data)
+overall_count <- pm_count + wos_count
 ```
+
+The search yielded `r overall_count` results (Pubmed: `r pm_count`; WOS: `r wos_count`)
 
@@
 # Count removal from missing doi
-sum(no_doi)
+no_doi_count <- sum(no_doi)
 
@@
 # Count removal of duplicates
-sum(dupl)
+dupl_count <- sum(dupl)
 ```
+
+Removed because of missing DOI: `r no_doi_count`
+Duplicates removed: `r dupl_count`
+
+Remaining: `r overall_count - no_doi_count - dupl_count`
 
@@
 # Count removal by title filtering
-sum(excl)
+title_excl_count <- sum(excl)
 # count remaining articles
-nrow(filtered_data)
+forsample_count <- nrow(filtered_data)
 ```
+
+Removed by automated title filtering: `r title_excl_count`
+Remaining articles for random sampling: `r forsample_count`
 
@@
 # number excluded 
-sum(!is.na(s_res_final$decision))
+abstr_excl_count <- sum(!is.na(s_res_final$decision))
 # extract remaining data
 ft_screen <- sampled_data[is.na(s_res_final$decision), ]
-
-l <- list.files("../data/fulltext")
-get_ft_id <- function(n, files) {
-  name <- files[n]
-  as.numeric(strsplit(name, "\\.")[[1]][[1]])
-}
-fts <- purrr::map_dbl(seq_along(l), get_ft_id, files = l)
-# find missing full texts
-# ft_screen$sid[!ft_screen$sid %in% fts]
 
@@
 ```
+
+Removed by abstract screening: `r abstr_excl_count`
+
+#### Compare full text screening results
+
+```{r fulltext-screening}
+# read screening results
+f_res_i2 <- read.csv("../data/screen/fulltext_screening_i2.csv")
+f_res_i1 <- read.csv("../data/screen/fulltext_screening_i1.csv", sep = ";")
+
+f_res1 <- f_res_i1$fulltext_exclusion
+f_res2 <- f_res_i2$fulltext_exclusion
+
+f_res <- purrr::map2_chr(f_res1, f_res2, comp_res)
+f_res_df <- data.frame(
+  sid = f_res_i2$sid,
+  i1 = f_res1,
+  i2 = f_res2,
+  result = f_res,
+  decision = rep.int("", length(f_res1)),
+  i1_note = f_res_i1$exclusion_note,
+  i2_note = f_res_i2$exclusion_note,
+  doi = f_res_i1$doi
+)
+# write.csv(f_res_df, "../data/screen/fulltext_screening.csv", row.names = FALSE)
+# Calculate relative agreement
+1 - (sum(f_res == "XXX", na.rm = TRUE) / length(f_res))
+
+1 - ((sum(is.na(f_res1) | is.na(f_res2)) - sum(is.na(f_res))) / length(f_res))
+
+# read final decision
+ft_res <- read.csv("../data/screen/fulltext_screening.csv")
+ft_res$decision <- factor(
+  ft_res$decision, 
+  levels = c("t", "l", "r", "h", "m", "c", "e", "d", "o"),
+  labels = c("no fulltext", "no english full-text", "no original research", 
+             "no humans", "VO2 not measured", "VO2 not continously measured", 
+             "no exhaustion", "long protocol", "other")
+)
+# count exclusions
+ft_excl_count <- sum(!is.na(ft_res$decision))
+
+# count remaining articles for data extraction
+final_count <- sum(is.na(ft_res$decision))
+
+```
+
+Excluded after full text screening: `r ft_excl_count`
+
+Remaining articles: `r final_count`
+
+#### Prepare data extraction
+
+```{r extraction, eval=FALSE}
+emp <- rep.int("", final_count)
+
+extr <- data.frame(
+  sid = ft_res$sid[is.na(ft_res$decision)],
+  cart = emp,
+  type = emp,
+  outcome = emp,
+  pre = emp,
+  software = emp,
+  interpolation = emp,
+  ptype = emp,
+  averaging = emp,
+  parameters = emp,
+  source = emp,
+  doi = ft_res$doi[is.na(ft_res$decision)]
+)
+
+write.csv(extr, "../data/extract/extraction.csv", row.names = FALSE)```

### 2022-04-15 transparent_changes.Rmd
Update transparent changes document

@@
 
+## Major Changes
+
+#### Addditional variables extracted from included research
+
+`outcome`: Which type of outcome VO2max is. Either `primary`, `secondary` or `other`
+`source`: Which source is provided for the data processing method used.
+
 ## Minor Changes
@@
 
-#### Unblinding of single abstract
+#### Unblinding of single abstracts
 
@@
 * Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the Sampling ID (sid): 275
+
+#### Minor Modification of exclusion criteria
+
+Changes are in italics:
+
+`'r'`: Is the article no original research *(i.e. no primary analysis of experimental data)* ? (if yes, indicate 'r'; if no, continue)
+
+`'t'`: Was no full-text ~~available~~ *accessible* for the corresponding article? (if yes, indicate 't', if no continue)

### 2022-04-15 thesis/methods.Rmd
Work on methods section of thesis

@@
 The work presented in this thesis was preregistered before the start of the project on [OSF]().
+Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix).
+All data and code can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).
 
-<!-- deviations from preregistration -->
+<!-- explicitly state deviations from preregistration -->
 
@@
 To assess the current state of data processing methods for determining V̇O~2max~ in exercise science I systematically screened recent literature.
-The article search was conducted on XX March 2022 using [PubMed]() and [Web of Science]().
-The search included articles articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword.
+The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science]().
+The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword.
 Table 1 shows the exact search terms.
@@
 Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step.
-After the removal of duplicates I conducted an automated title scanning, to remove results that are likely no original research articles.
+After the removal of duplicates I conducted an automated title scanning to remove results that are likely no original research articles.
 Table 2 displays the exclusion terms during the automated title screening.
@@
 The goal of the random sample was to give an unbiased sample of the current state of scientific V̇O~2max~ testing.
-Based on the procedure described in the preregistration file (LINK), <!-- random sample size  -->
+Based on the procedure described in the preregistration file (LINK), the sample included a total of 500 articles.
 
@@
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
+When the screeners disagreed in their assessments, conflicts were resolved by discussion.
 
 After the abstract screening I retrieved the full-texts for all articles remaining in the review.
-The full-texts were again scanned to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria).
+The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria).
+Conflicts were resolved by discussion.
 
@@
 -   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)
+
+-   type of outcome for V̇O~2max~ (primary, secondary, other)
 
@@
     -   interval (in sec/breath, parameters for filtering)
+    -   interpolation procedure
+
+-   reference for the used data processing strategy
 
@@
 ### Data Synthesis
+
+<!-- presentation of data -->
 
@@
 
-A total of N = 76 exercise test were analysed for this study.
+A total of N = 72 exercise test were analysed for this study.
 The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
 The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times).
-The V̇O~2max~ testings were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol.
-Participants run on a treadmill (XXX) for ten minutes at a velocity of 2.8 m/s as a warm-up.
-After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increases of 0.15 m/s every 30 seconds.
+The V̇O~2max~ tests were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol.
+Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up.
+After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increased by 0.15 m/s every 30 seconds.
 The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

### 2022-04-27 scripts/review.Rmd
Analyze extracted data

@@
 abstr_excl_count <- sum(!is.na(s_res_final$decision))
+# reasons for exclusion
+table(s_res_final$decision)
 # extract remaining data
@@
 
-write.csv(extr, "../data/extract/extraction.csv", row.names = FALSE)```
+write.csv(extr, "../data/extract/extraction.csv", row.names = FALSE)
+```
+
+#### Read and clean extracted data
+
+```{r clean, warning=FALSE}
+ext <- read.csv("../data/extract/extraction.csv")
+
+# Create column for processing type + averaging type combination
+ext$proc_combination <- paste(ext$ptype, ext$averaging, sep = "-")
+ext$proc_combination[ext$proc_combination == "NA-NA"] <- NA
+
+# Separate parameter column for multiple parameters (i.e. multiple binned averages)
+# and calculate total duration
+
+split_parameters <- function(n) {
+  value <- ext$parameters[n]
+  if (grepl("x", value)) {
+    value_split <- strsplit(value, "x")[[1]]
+    out <- data.frame(
+      param_times = as.numeric(value_split[1]), 
+      param_value = as.numeric(value_split[2])
+    )
+    out$param_duration <- out$param_times * out$param_value
+  } else {
+    out <- data.frame(
+      param_times = NA, 
+      param_value = as.numeric(value), 
+      param_duration = as.numeric(value)
+    )
+  }
+  out
+}
+
+ext <- cbind(ext, purrr::map_dfr(seq_len(nrow(ext)), split_parameters))
+
+# save data
+# save(ext, file = "../data/review.Rda")
+
+# Show metabolic carts used
+table(ext$type, useNA = "ifany")
+```
+
+#### Analyze extracted data
+
+```{r percentages}
+load("../data/review.Rda")
+# Remove Douglas bag measurements
+ext <- ext[ext$type != "douglas", ]
+
+n_total <- nrow(ext)
+ext_bbb <- ext[ext$type == "bbb" & !is.na(ext$type),]
+n_bbb <- nrow(ext_bbb)
+
+# Type of metabolic cart
+p_cart <- 1 - (sum(grepl("NA", ext$cart) | is.na(ext$cart)) / n_total)
+
+# Preprocessing
+p_pre <- sum(!is.na(ext_bbb$pre)) / n_bbb
+
+# Software
+p_software <- sum(!is.na(ext_bbb$software)) / n_bbb
+
+# Processing strategy
+p_proc <- sum(!is.na(ext$ptype) & !is.na(ext$averaging) & !is.na(ext$parameters)) / n_total
+
+# Source
+p_source <- sum(!is.na(ext$source)) / n_total
+
+reporting <- data.frame(
+  "cart" = p_cart,
+  "pre" = p_pre,
+  "software" = p_software,
+  "processing" = p_proc,
+  "source" = p_source
+)
+
+save(reporting, file = "../data/reporting.Rda")
+
+knitr::kable(purrr::map_dfr(reporting, scales::percent, accuracy = 0.1))
+
+```
+
+```{r analysis}
+# Show data processing strategies overall
+table(ext$ptype, useNA = "ifany")
+
+# Show data processing type for breath-by-breath measurements only
+table(ext$proc_combination[ext$type == "bbb"], useNA = "ifany")
+
+# Plot for total calculation interval lengths
+ext[ext$ptype == "time" & !is.na(ext$ptype), ] |>
+  ggplot(aes(y = param_duration)) +
+    geom_bar() +
+  geom_text(aes(label = ..count..), stat = "count", hjust = 1.2, colour = "white") +
+    scale_y_reverse(
+      breaks = c(5, 10, 15, 20, 30, 45, 60),
+      minor_breaks = NULL,
+      name = "Duration (s)"
+    ) +
+    scale_x_continuous(
+      minor_breaks = NULL,
+      name = "Count",
+      breaks = c(0, 10, 20, 30, 40),
+      labels = NULL,
+      expand = c(0, 0),
+      position = "top"
+    ) +
+    theme_minimal(13) +
+   theme(
+     panel.grid.major.y = element_blank(),
+     axis.ticks.y = element_line(colour = "grey", size = 0.5)
+    )
+
+ggsave("../plots/duration_count.png", width = 5, height = 3, dpi = 600, bg = "white")
+```
+
+```{r}
+library(treemapify)
+library(dplyr)
+
+ext_count <- ext |> 
+  filter(type == "bbb") |>
+  select(c(proc_combination, ptype)) |>
+  filter(!is.na(proc_combination)) |>
+  count(proc_combination, ptype)
+
+ext_count$proc_combination <- factor(
+  ext_count$proc_combination, 
+  levels = c("breath-binned", "time-moving", "breath-moving", "time-binned", "time-mbinned"),
+  labels = c("Binned Breath","Moving Time", "Moving Breath", "Binned Time", "Multiple Binned Time"))
+
+percs <- scales::percent(ext_count$n / sum(ext_count$n))
+ext_count$label <- paste0(ext_count$proc_combination, " (", percs, ")")
+ext_count$subgroup <- c(1, 1, 2, 3, 4)
+
+ggplot(ext_count, aes(area = n, fill = proc_combination, label = label, subgroup = subgroup)) +
+  geom_treemap(show.legend = FALSE, start = "topleft") +
+  geom_treemap_text(reflow = TRUE, place = "centre", start = "topleft")
+
+ggsave("../plots/strategies.png", width = 5, height = 3, dpi = 600, bg = "white")
+```

### 2022-04-28 scripts/comparison.Rmd
Analyze experimental data

# Analysis Script for the Experimental Comparison

## Project: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'

This script closely follows the preregistration, which is uploaded on OSF.

```{r setup}
# Packages used for the data workflow

library(purrr)
library(spiro)
library(tidyr)
library(ggplot2)
library(scales)
library(MetBrewer)

# Get data files

files <- list.files("../data/ramptests", full.names = TRUE)
```

# Functions to calculate VO2max using different strategies

```{r}
# Moving time average
max_time_moving <- function(data, parameter) {
  spiro::spiro_max(data = data, smooth = parameter)$VO2_rel
}

# Moving breath average
max_breath_moving <- function(data, parameter) {
  p <- paste0(parameter, "b")
  spiro::spiro_max(data = data, smooth = p)$VO2_rel
}

# Binned time average
max_time_binned <- function(data, parameter) {
  nbin <- ceiling(nrow(data) / parameter)
  ibins <- seq(from = 1, by = parameter, length.out = nbin)
  binmean <- function(i, length, data) {
    values <- data$VO2_rel[i:(i + length - 1)]
    values[is.na(values)] <- 0
    mean(values)
  }
  binmeans <- purrr::map_dbl(.x = ibins, .f = binmean, length = parameter, data = data)
  max(binmeans)
}

vo2max <- function(parameter, method, data) {
  if (method == "bm") {
    result <- max_breath_moving(data = data, parameter = parameter)
  } else if (method == "tm") {
    result <- max_time_moving(data = data, parameter = parameter)
  } else { # time binned
    result <- max_time_binned(data = data, parameter = parameter)
  }
  out <- data.frame(
    name = result
  )
  colnames(out) <- paste(method, parameter, sep = "_")
  out
}
```

```{r calc}
# ZAN raw data only saves imprecise body mass data in the raw data file.
info <- read.csv("../data/participants.csv")

# level 1: iterate over each data file
results <- purrr::map_dfr(
  .x = seq_along(files),
  .f = function(n) {
    # level 2: iterate over each processing method
    data <- spiro(files[n], weight = info$bodymass[n])
    purrr::map_dfc(
      .x = c("bm", "tm", "tb"), 
      .f = function(method, data) {
        # level 3: iterate over each parameter range
        purrr::map_dfc(
          .x = seq.int(5,60),
          .f = vo2max,
          method = method,
          data = data
        )
      }, 
      data = data
    )
  }
)

write.csv(results, "../data/results.csv", row.names = FALSE)
```

## Normalize results

```{r}
res <- read.csv("../data/results.csv")

# Reference method: 30 seconds binned average
res_norm <- res / res$tb_30
res_norm$id <- as.factor(seq_len(nrow(res_norm)))
```

## Turn into tidy format

```{r}
res_tidy <- pivot_longer(
  data = res_norm, 
  cols = !id,
  names_to = c("type", "parameter"),
  names_pattern = "(.*)_(.*)",
  names_transform = list(
    parameter = ~ as.numeric(.x),
    type = ~ factor(.x, 
      levels = c("bm", "tm", "tb"), 
      labels = c("moving breath", "moving time", "binned time")
    )
  )
)

# calculate normalized results as percent difference
res_tidy$perdiff <- (res_tidy$value - 1) / 1
```

## Plot data

```{r}
quant <- function(data, lower, upper) {
  qs <- quantile(data, probs = c(lower, upper), names = FALSE)
  data.frame(
    ymin = qs[[1]],
    ymax = qs[[2]]
  )
}

ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
  geom_hline(
    yintercept = 0,
    colour = "grey",
    alpha = 0.3
  ) +
  stat_summary(geom = "line", fun = "median") +
  stat_summary(
    geom = "ribbon", 
    fun.data = "quant", 
    fun.args = list(lower = 0.1, upper = 0.9),
    alpha = 0.1, 
    colour = NA
  ) +
  scale_x_continuous(
    name = "Parameter",
    breaks = c(15, 30, 45, 60)
  ) +
  scale_y_continuous(
    name = "% to Reference",
    labels = scales::percent,
    limits = c(-0.053,0.1)
  ) +
  scale_color_manual(
    values = met.brewer("Java", 3)
  ) +
  scale_fill_manual(
    values = met.brewer("Java", 3)
  ) +
  theme_classic() +
  theme(
    legend.position = c(0.8,0.8)
  )

ggsave("../plots/comparison.png", width = 6, height = 4, dpi = 600, bg = "white")
```

## Respiratory Rate over time

```{r}
read_rr <- function(file) {
  dat <- spiro(file)
  # Zero-lag Butterworth filter
  rr <- spiro_smooth(dat, smooth = "fz", columns = "RR")[, 1]
  time <- attr(dat, "raw")$time
  load <- attr(dat, "raw")$load

  filename_split <- strsplit(file, "/")[[1]]
  id <- filename_split[length(filename_split)]

  rr_df <- data.frame(
    id = id,
    time = time,
    value = rr,
    load = load,
    is_end = FALSE
  )

  # remove pre and post exercise data
  out <- rr_df[rr_df$load != 0, ]

  # mark last 60 seconds
  out$is_end[out$time >= (max(out$time) - 60)] <- TRUE

  out
}

# read all respiratory rate data
rr_all <- purrr::map_dfr(files, read_rr)

# Line plots with large data are very slow in ggplot2
# using a scatter plot instead
p_rr <- ggplot(rr_all, aes(x = time, y = value, colour = is_end, alpha = is_end)) +
  geom_point(size = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = c("black", met.brewer("Peru2")[2])) +
  scale_alpha_manual(values = c(0.05, 0.3)) +
  scale_x_continuous(
    name = "Time (min)",
    limits = c(60,950),
    breaks = c(180, 360, 540, 720, 900),
    labels = c(2, 5, 8, 12, 15),
    expand = c(0,0)
  ) +
  scale_y_continuous(
    name = expression(paste("Respiratory Rate ", (min^-1))),
    breaks = c(20,40,60,80)
  ) +
  theme_classic()

ggsave(p_rr, filename = "../plots/rr.png", width = 6, height = 4, dpi = 600, bg = "white")
```

### 2022-04-28 thesis/discussion.Rmd
Continue writing on thesis

@@
 # Discussion
+
+<!-- if report - diverse strategies -->
+<!-- most (multiple) binned time averages: necessary for mixing chamber, but bbb has more information! -->
+<!-- current practice is in contrast to recommendations (Robergs)! -->
+
+<!-- variety of calculation intervals - different results (example)-->
+<!-- comparing results for strategies -->
+
+<!-- recommendation for choosing a strategy: ... -->
+
+<!-- guidelines for reporting: ... -->

### 2022-04-28 thesis/methods.Rmd
Continue writing on thesis

@@
 
-The work presented in this thesis was preregistered before the start of the project on [OSF]().
+The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework]().
 Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix).
-All data and code can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).
+Major deviations will also explicitly name be named within the methods section.
+All data and code of this research project can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).
 
@@
 
-<!-- Detailed Aim of review -->
+The aim of the Scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature.
+Since the V̇O~2max~ determination is too common to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords.
+Data on processing strategies was extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
+
+The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A contains the corresponding reporting checklist.
 
@@
 
-To assess the current state of data processing methods for determining V̇O~2max~ in exercise science I systematically screened recent literature.
 The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science]().
 The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword.
-Table 1 shows the exact search terms.
+Table 1 shows the exact search terms used.
 
@@
   col.names = c("Source", "Search String"), 
-  caption = "Search strings for the systematic scoping review.")
+  caption = "Search strings for the systematic scoping review."
+)
 ```
 
-The search results from both data bases were joined at checked for the presence of a DOI.
+The search results from both data bases were joined and checked for the presence of a DOI.
 Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step.
-After the removal of duplicates I conducted an automated title scanning to remove results that are likely no original research articles.
+After the removal of duplicates I conducted an automated title scanning to remove results that were likely no original research articles.
 Table 2 displays the exclusion terms during the automated title screening.
@@
 The abstracts from the articles included in the random sample were blinded for scanning.
-This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the articles (see Figure 2).
+This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix XXX for an example).
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
@@
 
--   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)
+-   type of metabolic cart/data basis (breath-by-breath, mixing chamber, ...)
 
@@
 
-All extracted data is available in the Appendix A and under XXXLINK.
+All extracted data is available in the Appendix XXX and under XXXLINK.
 
@@
 
-<!-- presentation of data -->
+The extracted data is presented in a purely descriptive way. I calculated relative and absolute proportions of the different extracted variables.
 
@@
 Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up.
-After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increased by 0.15 m/s every 30 seconds.
+After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds.
 The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.
@@
 
-The spiro Package for R [@nolte2022a] processed the raw gas exchange data.
+The spiro Package Version XXX for R [@nolte2022a] processed the raw gas exchange data.
 It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data.
-Table 4 displays a brief summary exemplary function parameters used for the analysis.
+Table 4 displays a brief summary of exemplary functions and parameters used for the analysis.
 The full analysis script is available at XXXLINK.
+
+```{r spiro-function, echo = FALSE}
+spiro_functions <- data.frame(
+  proc_type = c("Moving Time Average", "Binned Time Average", "Multiple Binned Time Average", "Moving Breath Average", "Digital Filtering"),
+  func = c("`spiro_max()`", "`spiro_summary()`", "`spiro_summary(); filter()`", "`spiro_max()`", "`spiro_max()`"),
+  parameter = c("`smooth = 30`", "XXX", "XXX", '`smooth = "10b"`', '`smooth = "XXX"')
+)
+
+knitr::kable(spiro_functions)
+```
 

### 2022-04-28 thesis/results.Rmd
Continue writing on thesis

@@
 
+Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in Figure 1).
+Out of the random sample (n = 500), 244 articles were included in the final analysis.
+
+![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png)
+
+Reporting practices of the methodology of gas exchange measures differed widely within the literature (see Table 4).
+More than half (51.8%) of the articles did not report any information on their data processing strategy.
+Only one in twenty articles (5.4%) provided a rationale for their used strategy.
+
+```{r reporting-table, echo=FALSE}
+load("../data/reporting.Rda")
+t <- purrr::map_dfr(reporting, scales::percent, accuracy = 0.1)
+
+t[["pre"]] <- paste0(t[["pre"]], "*")
+t[["software"]] <- paste0(t[["software"]], "*")
+
+knitr::kable(
+  t,
+  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
+  caption = "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing.
+*only examined within the subgroup of studies using breath-by-breath measurements",
+  align = "r"
+)
+```
+
+Out of the authors that provided information and collected breath-by-breath measurements most (79.5%) utilized binned averages to determine V̇O~2max~.
+Moving time averages, or breath-based averages were uncommon (see Figure 2).
+No study used digital filtering methods to determine V̇O~2max~.
+
+![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png)
+
+The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3).
+30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20s) and longer (60s) periods.
+
+![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png)
+
 ## Experimental Comparison
 
-![A plot](../preregistration/example_plot.png)
+![VO2max varies by data processing strategy.
+Values are expressed relative to the VO2max from a 30-second binned average --- the most common strategy as determined by the review.
+Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
+Using moving average leads to systematically higher VO2max values compared to binned time averages.
+Changing the averaging interval (in seconds or breaths) can lead to changes in VO2max as large as 5%.](../plots/comparison.png)
+
+![Respiratory Rates peak around 60min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72). .](../plots/rr.png)

### 2022-05-02 transparent_changes.Rmd
Add more transparent changes

@@
 urlcolor: blue
+bibliography: references.bib
 ---
@@
 
-#### Addditional variables extracted from included research
+### Number of exercise tests for comparison
+
+Due to a miscalculation, the preregistration provided an incorrect number of exercise test (n = 76).
+The correct number of exercise tests is n = 72, with 44 from @quittmann_unpubl_a (one test only partly included in the original work due to missing other data) and 28 from @quittmann_unpubl_b (three test only partly included in the original work due to missing other data).
+
+#### Additional variables extracted from included research
 
 `outcome`: Which type of outcome VO2max is. Either `primary`, `secondary` or `other`
+
 `source`: Which source is provided for the data processing method used.
@@
 
-* advanced detection of missing DOIs: `is.na(merge_data$doi) | (merge_data$doi == "")` instead of `is.na(merge_data$doi)`.
-* Improved function to retrieve missing PubMed abstracts: Handles case when input (PMID) is missing (`if (is.na(pmid)) return(NA)`).
-* save/load of the sampling results as an .Rda file to reduce computation time when working on parts of the workflow.
+-   advanced detection of missing DOIs: `is.na(merge_data$doi) | (merge_data$doi == "")` instead of `is.na(merge_data$doi)`.
+-   Improved function to retrieve missing PubMed abstracts: Handles case when input (PMID) is missing (`if (is.na(pmid)) return(NA)`).
+-   save/load of the sampling results as an .Rda file to reduce computation time when working on parts of the workflow.
 
@@
 
-* Manual retrieval of abstracts for articles, as these were neither present in the search result data, nor could be automatically scraped.
+-   Manual retrieval of abstracts for articles, as these were neither present in the search result data, nor could be automatically scraped.
 Abstracts were saved and imported as .txt files.
@@
 This concerns the abstract with the Sampling ID (sid): 50, 238, 288, 416, 488, 490, 500
-* Manual retrieval of abstracts for articles, as the automatically collected abstract contained html-tags that could not be removed for later abstract plots.
+-   Manual retrieval of abstracts for articles, as the automatically collected abstract contained html-tags that could not be removed for later abstract plots.
 Abstracts were saved and imported as .txt files.
@@
 This concerns the abstract with the Sampling ID (sid): 344, 356
-* Unblinding during screening to assess the implications of title given in squared brackets. This concerns the abstract with the Sampling ID (sid): 262, 303
-* Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the Sampling ID (sid): 275
+-   Unblinding during screening to assess the implications of title given in squared brackets. This concerns the abstract with the Sampling ID (sid): 262, 303
+-   Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the Sampling ID (sid): 275
 
@@
 `'t'`: Was no full-text ~~available~~ *accessible* for the corresponding article? (if yes, indicate 't', if no continue)
+
+#### Minor screening error for two articles
+
+For two articles (sampling ids: 194, 282) I only realized during data extraction that they matched the exclusion criteria (\`c\`: no continuous measurement of oxygen uptake).
+In agreement of both screeners, the screening data was retrospectively changed.

### 2022-05-02 scripts/review.Rmd
Update review script

@@
 
-write.csv(extr, "../data/extract/extraction.csv", row.names = FALSE)
+# write.csv(extr, "../data/extract/extraction.csv", row.names = FALSE)
 ```
@@
 load("../data/review.Rda")
-# Remove Douglas bag measurements
-ext <- ext[ext$type != "douglas", ]
 
@@
 knitr::kable(purrr::map_dfr(reporting, scales::percent, accuracy = 0.1))
-
 ```
@@
 library(dplyr)
+library(colorspace)
 
@@
   geom_treemap(show.legend = FALSE, start = "topleft") +
-  geom_treemap_text(reflow = TRUE, place = "centre", start = "topleft")
+  geom_treemap_text(
+    color = "black",
+    reflow = TRUE, 
+    place = "centre", 
+    start = "topleft",
+    padding.y = grid::unit(1.5, "mm")) +
+  scale_fill_manual(
+    values = colorspace::lighten(c("#663171", "#ea7428", "#653151", "#0c7156", "#0c7182"), amount = 0.2)
+  )
 

### 2022-05-02 scripts/comparison.Rmd
Update comparison script

@@
 
-# Functions to calculate VO2max using different strategies
+## Functions to calculate VO2max using different strategies
 
@@
 # Binned time average
-max_time_binned <- function(data, parameter) {
+max_time_binned <- function(data, parameter, max = TRUE) {
   nbin <- ceiling(nrow(data) / parameter)
@@
   binmeans <- purrr::map_dbl(.x = ibins, .f = binmean, length = parameter, data = data)
-  max(binmeans)
+  if (max) max(binmeans) else binmeans
 }
@@
 ```
+
+## Iterate over all exercise tests
 
@@
 
-## Plot data
+## Plot data: Comparison of processing strategies
 
 ```{r}
+# Function for calculating quantils
 quant <- function(data, lower, upper) {
@@
 
-## Respiratory Rate over time
+## Plot: Respiratory Rate over time
 
@@
   dat <- spiro(file)
-  # Zero-lag Butterworth filter
+  # Zero-lag Butterworth filter with default parameters
   rr <- spiro_smooth(dat, smooth = "fz", columns = "RR")[, 1]
@@
 
+  # file id
   filename_split <- strsplit(file, "/")[[1]]
@@
 ```
+
+# Plot: Example of different strategies
+
+```{r}
+id <- 46
+id_info <- read.csv("../data/participants.csv")
+bodymass <- id_info$bodymass[id_info$p == id]
+
+exm_data <- spiro(files[id])
+raw_vo2 <- attr(exm_data, "raw")$VO2
+raw_time <- attr(exm_data, "raw")$time
+fz <- spiro_smooth(exm_data, "fz", "VO2")[,1]
+mb <- spiro_smooth(exm_data, "30b", "VO2")[,1]
+mt <- spiro_smooth(exm_data, 30, "VO2")[,1]
+tb <- max_time_binned(exm_data, 30, max = FALSE) * attr(exm_data, "info")[["weight"]]
+rb_time <- seq(from = 15, by = 30, length.out = length(tb))
+
+exm_tidy <- data.frame(
+  method = c(rep.int("raw", length(raw_vo2)), rep.int("fz", length(fz)), rep.int("mb", length(mb)), rep.int("mt", length(mt)), rep.int("tb", length(tb))),
+  value = c(raw_vo2, fz, mb, mt, tb),
+  time = c(rep.int(raw_time, 3), seq_along(mt), rb_time)
+)
+
+exm_tidy$method <- factor(
+  exm_tidy$method, 
+  levels = c("raw", "mb", "mt", "tb", "fz"), 
+  labels = c("raw", "Moving Breath", "Moving Time", "Binned Time", "Digital Filter")
+)
+exm_tidy$value <- exm_tidy$value / bodymass
+
+ggplot(exm_tidy, aes(x = time)) +
+  geom_point(data = exm_tidy[exm_tidy$method == "raw", ], aes(y = value), alpha = 0.1) +
+  geom_line(data = exm_tidy[exm_tidy$method %in% c("Moving Time", "Moving Breath", "Digital Filter"), ], aes(y = value, color = method), size = 0.7) +
+  geom_point(data = exm_tidy[exm_tidy$method == "Binned Time", ], aes(y = value, fill = method), color = met.brewer("Java", 3)[3], size = 1.2) +
+  scale_x_continuous(
+    name = "Time (min)",
+    limits = c(500,850),
+    breaks = c(540, 630, 720),
+    labels = c(8, 10, 12),
+    expand = c(0,0)
+  ) +
+  scale_y_continuous(
+    name = "VO2 (ml/min/kg)",
+    limits = c(55,75)
+  ) +
+  scale_color_manual(
+    name = "",
+    values = c(met.brewer("Java", 3)[1:2], "blue")
+  ) +
+   scale_fill_manual(
+    name = "",
+    values = "white"
+  ) +
+  theme_classic() +
+  theme(
+    legend.position = c(0.9, 0.55),
+  )
+
+ggsave(filename = "../plots/example.png", width = 6, height = 4, dpi = 1200, bg = "white")
+```

### 2022-05-02 thesis/introduction.Rmd
Continue writing thesis

@@
 
-<!-- VO2max is important -->
-<!-- physiological determinants -->
-<!-- VO2max testing -->
-<!-- Influence factors in testing -->
+### Maximum oxygen uptake: Definition and Relevance
+
+Performance in endurance sports is limited by the human physiology.
+Athletes need to supply energy to locomotion, a process that happens mainly via the oxidative phosphorylation.
+A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster.
+On a whole-body level the highest activity of oxidative phosphorylation can be approximated by measuring the maximum oxygen uptake (V̇O~2max~).
+
+V̇O~2max~ is defined as the highest rate a body can consume oxygen.
+Determined by measuring gas exchange data, it is one of the most common measures investigated in sports science and medicine.
+V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups, and thus can be regarded the most relevant physiological parameter for endurance performance prediction.
+Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~
+
+### Measuring the VO~2max~
+
+Researchers measure V̇O~2max~ during exercise tests to exhaustion.
+In a laboratory setting such exercise commonly takes place on treadmills or cycling Prior fatigue results in underestimating the true V̇O~2max~.
+Therefore most exercise protocol consist of a continuous or stepwise, fast increase in load.
+As the oxygen kinetics themselves comprise a time lag, XXX suggested an optimal duration of 8-12 minutes for testing V̇O~2max~.
+Despite protocol type and duration, factors such as nutrition, XXX bias the V̇O~2max~ determination.
+
+Early research in V̇O~2max~ used bags --- the so called Douglas bag method --- to collect expired air for later analysis.
+Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chamber (which sample data at fixed time intervals, e.g., 15s) or breath-by-breath methods.
+While mixing chambers are regarded as more reliable, measuring breath-by-breath generates more data with a higher temporal resolution.
+<!--Which is more common?--> The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~.
+Another important, yet often overlooked factor influencing the V̇O~2max~ is the data processing of the raw data.
+
+### VO~2max~ vs VO~2peak~
+
+Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades.
+To identify a 'true' maximum researchers evaluate a set of parameters measured in a ramp test --- the V̇O~2max~ criteria [@howley1995].
+If these criteria a (partly) not fulfilled, researchers should speak of the peak oxygen uptake instead of V̇O~2max~.
+In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ themselves (e.g.
+the primary criterium of a plateau in oxygen uptake; the secondary criterium of the maximum respiratory quotient) heavily rely on the data processing strategy used [@astorino2009].
+I thus speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.
 
@@
 
-<!-- First example in 90s: applied research -->
+Measured breath-by-breath data is noisy (see grey points in Figure 1).
+Both the biological variability of breathing patterns (as well as irregular breaths such as coughs and swallowing) and the measurement error lead to highly fluctuating raw oxygen uptake data.
+For interpretation the raw data requires some form of processing.
+
+![Raw breath by breath data and different processing strategies during the final minutes of a ramp test to exhaustion. Grey points mark single breaths.](../plots/example.png)
+
+Different data processing strategies influence measured parameters of gas exchange.
+As soon as when the first automated systems for gas exchange measure were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
+Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998a; @astorino2009a; @sell2021].
+@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that consistency of strategies is key when comparing outcomes.
+Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear -log equations to compare mean V̇O~2max~ values derived from different processing strategies.
+There is inconclusive evidence whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
+
+Differences in V̇O~2max~ due to data processing can cause serious implications in practice.
+They hinder the comparability of data from studies within meta-analysis or the assessment of longitudinal data from athletes that used different labs [@martin-rincon2019].
+Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterium of V̇O~2max~ [@astorino2009].
+Crucially, in situation where individuals are classified by their V̇O~2max~ --- for example when describing the trainings status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassification [@johnson1998].
+
+Calls to standardize data processing strategies of gas exchange data are frequent (XXX; ...).
+@howley1995 argued to use longer calculation intervals (60 s), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~.
+Based on synthetic and experimental data @robergs2003a opposed these conclusion, stating that shorter calculation intervals are less erroneous.
+They argued to use breath-based moving averages as a strategy, as XXX.
+In a classic review, @robergs2010 argued for using digital filtering (i.e.
+a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004 .
+However, @robergs2010 acknowledge the lack of software implementing such procedures.
+
+In absence of one commonly accepted method, data processing varies among the literature.
+In light of the influence on outcome variables, @myers1990 highlighted the need to to report processing strategies in research articles.
+@midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath data in selected journals.
+They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively.
+One third of the studies did not describe their processing method at all.
+While providing interesting first insights into reporting and processing practices, the search by @midgley2007 was not systematic and its methods were not described in detail.
+@robergs2010 wrote that to investigate the current state of data processing strategies two possible approaches are "(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible".
+They provides the latter one with a total of 75 respondents, which reported a great variety in data processing strategies.
+Most researchers reported the use of binned time averages over 30 or 60 seconds.
+Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning.
+Practices in reporting and processing may have changed more than a decade after publication of such numbers.
+
+Astorini 2008: Processing strategy (bbb, 15/30/60 tb) influences VO2max and plateau incidence
+
+Johnson 1998: Processing strategy (bbb, 15/30/60 tb, 8bm) influences VO2max in cardiac patients for heart transplantation evalution
+
+Midgley 2007:
+
+-   processing strategy (tb 10/15/20/30/60) influences VO2max
+
+-   processing strategies vary within the literature (binned time far most common, non-systematic review)
+
+-   processing strategy does not influence reliability of VO2max -\> but consistency is important
+
+Robergs & Burnett 2003: comparing methods (tb, bb, bm) on synthetic and experimental data: Recommendation to use 8 breath moving average.
+Potential use of filtering methods.
+Shorter intervals - less error (opposing Howley!)
+
+Howley 1995: classic for VO2max criteria; longer intervals preferable (i.e. 60s), shorter introduce systematic bias towards extreme values (sure???)
+
+Hill 2003: Effect of processing (bm) on VO2max interacts with exercise protocol (= duration)
+
+Schaedler 2017: No interaction between protocol and processing methods (opposing Hill, bit different protocols)
+
+TO READ:
+
+Weir JP, Koerner S, Mack B, Masek J, Vanderhoff J, Heiderscheit BC. VO 2\
+plateau detection in cycle ergometry. J Exerc Physiol (2004); 7: 55--62. -\> first with Butterworth (even zero-lag!)
+
+ATS/ACCP Guidelines
+
+Macfarlane DJ. Automated metabolic gas analysis systems: a review. Sports\
+Med (2001); 31: 841--861.
+
+Roecker K, Prettin S, Sorichter S. Gas exchange measurements with high\
+temporal resolution: the breath-by-breath approach. Int J Sports Med\
+(2005); 26(Suppl. 1): S11--S1
+
 <!-- Since then: various example -->
+
 <!-- conceptual work: Robergs -->
+
 <!-- Comparison: martin-rincon -->
+
 <!-- Problem: group + individual -->
+
 <!-- Knowledge Gap -->
@@
 <!-- Gap -> we need this: -->
+
 <!-- overview what is actually done -->
+
 <!-- and how does this impact VO2max assessment -->
@@
 <!-- Review: mapping current methodological (+ reporting) state  -->
+
 <!-- Comparison: highlight advantage over previous -->
@@
 <!-- help compare results across studies + for individuals -->
+
 <!-- advise for use of methods  -->
+
 <!-- call for improved reporting -->

### 2022-05-02 thesis/methods.Rmd
Continue writing thesis

@@
 
-The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework]().
-Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix).
-Major deviations will also explicitly name be named within the methods section.
-All data and code of this research project can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).
-
-<!-- explicitly state deviations from preregistration -->
+The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s), following the 'Inclusive Systematic Review Registration Form' [@akker2020].
+Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A).
+Major deviations will also be explicitly stated within the methods section.
+All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).
 
@@
 
-The aim of the Scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature.
-Since the V̇O~2max~ determination is too common to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords.
-Data on processing strategies was extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
+The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature.
+Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords.
+Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
 
-The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A contains the corresponding reporting checklist.
+The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix B contains the corresponding reporting checklist.
 
@@
   source = c("PubMed", "Web of Science"),
-  term = c('((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))', '(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)')
+  term = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
 )
@@
 
-The search results from both data bases were joined and checked for the presence of a DOI.
+The search results from both data bases were joined and checked for the presence of a Digital Object Identifier (DOI).
 Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step.
-After the removal of duplicates I conducted an automated title scanning to remove results that were likely no original research articles.
+After the removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
 Table 2 displays the exclusion terms during the automated title screening.
@@
 
-In accordance with the preregistration a random sample was drawn from the search results.
-The goal of the random sample was to give an unbiased sample of the current state of scientific V̇O~2max~ testing.
-Based on the procedure described in the preregistration file (LINK), the sample included a total of 500 articles.
+In accordance with the preregistration I drew a random sample from the search results.
+The goal of this process was to give an unbiased sample of the current state of scientific V̇O~2max~ testing.
+Based on the procedure described in the preregistration, the sample included a total of 500 articles.
 
 The abstracts from the articles included in the random sample were blinded for scanning.
-This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix XXX for an example).
+This meant the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix C for an example).
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
-When the screeners disagreed in their assessments, conflicts were resolved by discussion.
+When the screeners disagreed in their assessments, they resolved conflicts by discussion.
 
-After the abstract screening I retrieved the full-texts for all articles remaining in the review.
-The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria).
+```{r exclusion, echo = FALSE}
+exclusion_criteria <- data.frame(
+  id = c("A*", "B*", "C*", "D*", "E", "F"),
+  criteria = c("not in English", "no primary research", "research not in humans", "VO2max only estimated", "no appropriate test protocol", "no information regarding the exclusion criteria"),
+  details = c("Full text only available in non-English language", "research was no original investigation or only a reanalysis of data", "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", "Protocol for VO2max testing did either not include exercise to voluntial exhaustion or was to long (>20 min) for a reliable estimate", "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing")
+)
+
+knitr::kable(
+  exclusion_criteria,
+  col.names = c("", "Criterium", "Details"),
+  caption = "Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated, during full text screening all were assessed."
+)
+```
+
+After the abstract screening I retrieved the full texts for all articles remaining in the review.
+The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the detailed full-text exclusion criteria).
 Conflicts were resolved by discussion.
 
-All data exclusion steps are documented under XXXLINK.
+All data exclusion steps are documented in an Markdown script on GitHub (LINKXXX).
 
@@
 
-All extracted data is available in the Appendix XXX and under XXXLINK.
+The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstract scanning, thus not stated in the preregistration.
+All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).
 
@@
 
-The extracted data is presented in a purely descriptive way. I calculated relative and absolute proportions of the different extracted variables.
+The extracted data is presented in a purely descriptive way.
+I calculated the relative and absolute frequency for the reporting of the extracted items.
+Similarly, I brook down the different data strategies use for processing data in all articles that reported the use of breath-by-breath measurements.
+Total interval duration of averaging procedures were derived from the reported parameters.
 
@@
 
-To determine the influence the most common data processing strategies have on the determination V̇O~2max~, I compared them on a set of already collected gas exchange data from running.
+To determine the influence the most common data processing strategies have on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from running.
 
@@
 A total of N = 72 exercise test were analysed for this study.
+Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
 The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
-The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times).
-The V̇O~2max~ tests were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol.
-Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up.
-After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds.
+The tested individuals were experienced distance runners (15 female, 55 male; three of the males participated in both studies).
+The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using exactly the same exercise protocol.
+Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up.
+After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds.
 The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.
 
-Gas exchange data were recorded using a ZAN XXX device (XXX).
-The device was calibrated with a 3l-syringe pump (XXX) and a reference gas (XXX) before each measurement.
-The measured breath-by-breath data is available at XXXLINK.
+Gas exchange data were recorded using a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America).
+The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement.
+The measured breath-by-breath data is available on GitHub (LINKXXX).
 
@@
 I calculated the data for all integer parameter values within the range of the values found in the literature during the review.
-On a group level I calculated the median and 2.5 and 97.5-quantiles of each processing strategy.
+On a group level I calculated the median and 10%- and 90%-quantiles of each processing strategy.
+
+In an additional exploratory I investigated the respiratory rate (number of breathes per minute) during the ramp tests.
+This may help to understand how breath-based and time-based data processing methods relate to each other.

### 2022-05-02 thesis/results.Rmd
Continue writing thesis

@@
 
-Reporting practices of the methodology of gas exchange measures differed widely within the literature (see Table 4).
+Reporting practices of the methodology of gas exchange measures differed widely across the literature (see Table 4).
 More than half (51.8%) of the articles did not report any information on their data processing strategy.
-Only one in twenty articles (5.4%) provided a rationale for their used strategy.
+One in twenty articles (5.4%) provided a rationale for their used strategy.
+Only a single article [@maturana2021] reported information regarding all the investigated criteria.
 
@@
 
+Data preprocessing
+
 The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3).
@@
 
-![VO2max varies by data processing strategy.
-Values are expressed relative to the VO2max from a 30-second binned average --- the most common strategy as determined by the review.
+The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 (mean ± standard deviation).
+Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see Figure 4).
+
+![V̇O~2max~ varies by data processing strategy.
+Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review.
 Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
-Using moving average leads to systematically higher VO2max values compared to binned time averages.
-Changing the averaging interval (in seconds or breaths) can lead to changes in VO2max as large as 5%.](../plots/comparison.png)
+Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
+Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png)
 
-![Respiratory Rates peak around 60min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72). .](../plots/rr.png)
+Binned time averages systematically lead to lower V̇O~2max~ values compared to moving averages.
+Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~-values.
+Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
+
+Moving time and moving breath averages yielded nearly identical values for V̇O~2max~ over all calculation intervals.
+This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see Figure 4).
+
+![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png)

### 2022-05-02 scripts/review.Rmd
Update flow diagram

@@
+---
+title: ''
+output: github_document
+bibliography: ../references.bib
+---
+
 # Analysis Script for the Systematic Review
@@
 
-This script closely follows the preregistration, which is uploaded on OSF.
+This script closely follows the preregistration, which is uploaded on the [OSF](https://osf.io/3am4s).
 
-```{r setup}
+```{r setup, warning = FALSE, message = FALSE}
+knitr::opts_chunk$set(out.width = "70%", fig.align = "center")
+
 # Packages used for the data workflow
-
 library(readxl)
@@
 library(purrr)
+library(dplyr)
 library(ggplot2)
 library(MetBrewer)
+library(treemapify)
+library(colorspace)
 ```
@@
 Removed because of missing DOI: `r no_doi_count`
+
 Duplicates removed: `r dupl_count`
@@
 Removed by automated title filtering: `r title_excl_count`
+
 Remaining articles for random sampling: `r forsample_count`
@@
 
-save(sampled_data, file = "../data/screen/sample.Rda")
+# save(sampled_data, file = "../data/screen/sample.Rda")
 ```
@@
 # Create blinded abstract plots
-purrr::walk(
-  .x = seq_len(nrow(sampled_data)),
-  .f = create_abstract_plot, 
-  data = sampled_data
-)
+# purrr::walk(
+#   .x = seq_len(nrow(sampled_data)),
+#   .f = create_abstract_plot, 
+#   data = sampled_data
+# )
 
@@
 # write.csv(s_res_df, "../data/screen/abstract_screening.csv")
+
 # Calculate relative agreement
@@
 )
+
 # write.csv(ft_screen_df, file = "../data/screen/fulltext_screening_i1.csv", row.names = FALSE)
@@
 )
+
 # write.csv(f_res_df, "../data/screen/fulltext_screening.csv", row.names = FALSE)
+
 # Calculate relative agreement
@@
 )
+
 # count exclusions
@@
 final_count <- sum(is.na(ft_res$decision))
+
+table(ft_res$decision, useNA = "ifany")
 
@@
 
-save(reporting, file = "../data/reporting.Rda")
+# save(reporting, file = "../data/reporting.Rda")
 
@@
 
-```{r analysis}
+#### Plot: Calculation intervals
+
+```{r interval, warning = FALSE}
 # Show data processing strategies overall
@@
 # Plot for total calculation interval lengths
-ext[ext$ptype == "time" & !is.na(ext$ptype), ] |>
+p_int <- ext[ext$ptype == "time" & !is.na(ext$ptype), ] |>
   ggplot(aes(y = param_duration)) +
@@
 
-ggsave("../plots/duration_count.png", width = 5, height = 3, dpi = 600, bg = "white")
+# ggsave(p_int, "../plots/duration_count.png", width = 5, height = 3, dpi = 600, bg = "white")
+
+knitr::include_graphics("../plots/duration_count.png")
 ```
 
-```{r}
-library(treemapify)
-library(dplyr)
-library(colorspace)
+#### Plot: Processing strategies
+
+```{r strategies}
 
@@
 
-ggplot(ext_count, aes(area = n, fill = proc_combination, label = label, subgroup = subgroup)) +
+p_strat <- ggplot(ext_count, aes(area = n, fill = proc_combination, label = label, subgroup = subgroup)) +
   geom_treemap(show.legend = FALSE, start = "topleft") +
@@
 
-ggsave("../plots/strategies.png", width = 5, height = 3, dpi = 600, bg = "white")
+# ggsave(p_strat, "../plots/strategies.png", width = 5, height = 3, dpi = 600, bg = "white")
+
+knitr::include_graphics("../plots/strategies.png")
 ```

### 2022-05-02 scripts/comparison.Rmd
Update plots

@@
+---
+title: ''
+output: github_document
+bibliography: ../references.bib
+---
+
 # Analysis Script for the Experimental Comparison
@@
 
-This script closely follows the preregistration, which is uploaded on OSF.
+This script closely follows the preregistration, which is uploaded on the [OSF](https://osf.io/3am4s).
 
-```{r setup}
+```{r setup, warning = FALSE, message = FALSE}
+knitr::opts_chunk$set(out.width = "70%", fig.align = "center")
+
 # Packages used for the data workflow
@@
 
-```{r}
+```{r fuctions}
 # Moving time average
@@
 
-```{r calc}
+```{r calc, eval = FALSE}
 # ZAN raw data only saves imprecise body mass data in the raw data file.
@@
 
-write.csv(results, "../data/results.csv", row.names = FALSE)
+# write.csv(results, "../data/results.csv", row.names = FALSE)
 ```
@@
 
-```{r}
+```{r norm}
 res <- read.csv("../data/results.csv")
@@
 
-```{r}
+```{r wrangle}
 res_tidy <- pivot_longer(
@@
       levels = c("bm", "tm", "tb"), 
-      labels = c("moving breath", "moving time", "binned time")
+      labels = c("Moving Breath", "Moving Time", "Binned Time")
     )
@@
 
-## Plot data: Comparison of processing strategies
+## Plot: Comparison of processing strategies
 
-```{r}
-# Function for calculating quantils
+```{r comparison, warning = FALSE}
+# Function for calculating quantiles
 quant <- function(data, lower, upper) {
@@
 
-ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
+p_comp <- ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
   geom_hline(
@@
   scale_color_manual(
+    name = "Processing Strategy",
     values = met.brewer("Java", 3)
@@
   scale_fill_manual(
+    name = "Processing Strategy",
     values = met.brewer("Java", 3)
@@
 
-ggsave("../plots/comparison.png", width = 6, height = 4, dpi = 600, bg = "white")
+# ggsave("../plots/comparison.png", plot = p_comp, width = 6, height = 4, dpi = 600, bg = "white")
+
+knitr::include_graphics("../plots/comparison.png")
 ```
@@
 
-```{r}
+```{r rr}
 read_rr <- function(file) {
@@
 
-ggsave(p_rr, filename = "../plots/rr.png", width = 6, height = 4, dpi = 600, bg = "white")
+# ggsave(filename = "../plots/rr.png", plot = p_rr, width = 6, height = 4, dpi = 600, bg = "white")
+
+knitr::include_graphics("../plots/rr.png")
 ```
 
-# Plot: Example of different strategies
+## Plot: Example of different strategies
 
-```{r}
+```{r example, warning = FALSE}
 id <- 46
@@
   levels = c("raw", "mb", "mt", "tb", "fz"), 
-  labels = c("raw", "Moving Breath", "Moving Time", "Binned Time", "Digital Filter")
+  labels = c("raw","Moving Breath","Moving Time","Binned Time", "Digital Filter"),
+  ordered = TRUE
 )
@@
 
-ggplot(exm_tidy, aes(x = time)) +
+p_ex <- ggplot(exm_tidy, aes(x = time)) +
   geom_point(data = exm_tidy[exm_tidy$method == "raw", ], aes(y = value), alpha = 0.1) +
-  geom_line(data = exm_tidy[exm_tidy$method %in% c("Moving Time", "Moving Breath", "Digital Filter"), ], aes(y = value, color = method), size = 0.7) +
-  geom_point(data = exm_tidy[exm_tidy$method == "Binned Time", ], aes(y = value, fill = method), color = met.brewer("Java", 3)[3], size = 1.2) +
+  geom_line(data = exm_tidy[exm_tidy$method %in% c("Moving Breath", "Moving Time", "Digital Filter"), ], aes(y = value, color = method), size = 0.7) +
+  geom_point(data = exm_tidy[exm_tidy$method == "Binned Time", ], aes(y = value, fill = method, colour = "Binned Time"), size = 1.4) +
   scale_x_continuous(
     name = "Time (min)",
-    limits = c(500,850),
+    limits = c(500,860),
     breaks = c(540, 630, 720),
@@
   scale_color_manual(
-    name = "",
-    values = c(met.brewer("Java", 3)[1:2], "blue")
+    name = "Processing Strategy",
+    values = c(met.brewer("Java", 3)[3], "blue", met.brewer("Java", 3)[c(1,2)]),
+    labels = c("Binned Time", "Digital Filter","Moving Time","Moving Breath"),
+    guide = guide_legend(
+      reverse = TRUE,
+      override.aes = list(
+        linetype = c(rep.int("solid", 3), "blank"),
+        shape = c(rep.int(NA, 3), 19),
+        color = c(met.brewer("Java", 3)[c(1,2)], "blue", met.brewer("Java", 3)[3])
+      )
+    )
   ) +
    scale_fill_manual(
-    name = "",
-    values = "white"
+    values = "white",
+    guide = "none"
   ) +
@@
   theme(
-    legend.position = c(0.9, 0.55),
+    legend.position = c(0.88, 0.6),
   )
 
-ggsave(filename = "../plots/example.png", width = 6, height = 4, dpi = 1200, bg = "white")
+# ggsave(filename = "../plots/example.png", plot = p_ex, width = 6, height = 4, dpi = 1200, bg = "white")
+
+knitr::include_graphics("../plots/example.png")
 ```

### 2022-05-02 scripts/review.Rmd
Update plots

@@
 
-# ggsave(p_int, "../plots/duration_count.png", width = 5, height = 3, dpi = 600, bg = "white")
+# ggsave("../plots/duration_count.png", plot = p_int, width = 5, height = 3, dpi = 600, bg = "white")
 
@@
 
-# ggsave(p_strat, "../plots/strategies.png", width = 5, height = 3, dpi = 600, bg = "white")
+# ggsave("../plots/strategies.png", plot = p_strat, width = 5, height = 3, dpi = 600, bg = "white")
 

### 2022-05-03 thesis/introduction.Rmd
Improve thesis methods and results

@@
 As soon as when the first automated systems for gas exchange measure were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
-Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998a; @astorino2009a; @sell2021].
+Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998; @astorino2009; @sell2021].
 @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that consistency of strategies is key when comparing outcomes.

### 2022-05-03 thesis/methods.Rmd
Improve thesis methods and results

@@
 
-The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s), following the 'Inclusive Systematic Review Registration Form' [@akker2020].
+The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020].
 Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A).
@@
 The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science]().
-The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword.
+The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords.
 Table 1 shows the exact search terms used.
@@
 
-The search results from both data bases were joined and checked for the presence of a Digital Object Identifier (DOI).
-Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step.
-After the removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
-Table 2 displays the exclusion terms during the automated title screening.
+The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI).
+Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step.
+After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
+Table 2 displays the exclusion terms for the automated title screening.
 
@@
   col.names = c("Search terms for exclusion by title"), 
-  caption = "Exclusion terms during automated title screening after removal of duplicates. If an articles title matched least one of the given terms, it was excluded.")
+  caption = "Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded.")
 ```
@@
 In accordance with the preregistration I drew a random sample from the search results.
-The goal of this process was to give an unbiased sample of the current state of scientific V̇O~2max~ testing.
+The goal of this process was to give an unbiased estimate of the current state of scientific V̇O~2max~ testing.
 Based on the procedure described in the preregistration, the sample included a total of 500 articles.
@@
 The abstracts from the articles included in the random sample were blinded for scanning.
-This meant the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix C for an example).
+This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix C for an example).
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
-When the screeners disagreed in their assessments, they resolved conflicts by discussion.
+When the screeners disagreed in their assessments, they resolved the conflict by discussion.
 
@@
   col.names = c("", "Criterium", "Details"),
-  caption = "Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated, during full text screening all were assessed."
+  caption = "Exclusion criteria for the screening process.
+During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed."
 )
@@
 
-All data exclusion steps are documented in an Markdown script on GitHub (LINKXXX).
+All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).
 
@@
 -   metabolic cart used
-
--   type of metabolic cart/data basis (breath-by-breath, mixing chamber, ...)
-
+-   measurement type (breath-by-breath, mixing chamber, ...)
 -   type of outcome for V̇O~2max~ (primary, secondary, other)
-
--   preprocessing algorithm
-
+-   data preprocessing (e.g., filtering)
 -   data processing software
-
 -   interpolation procedure
-
 -   data processing/determination of V̇O~2max~:
-
     -   type (time average, breath average, digital filtering)
     -   alignment (rolling, binned, ...)
-    -   interval (in sec/breath, parameters for filtering)
-    -   interpolation procedure
-
+    -   interval (in seconds or breaths, parameters for filtering)
 -   reference for the used data processing strategy
 
-The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstract scanning, thus not stated in the preregistration.
+The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.
 All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).
@@
 I calculated the relative and absolute frequency for the reporting of the extracted items.
-Similarly, I brook down the different data strategies use for processing data in all articles that reported the use of breath-by-breath measurements.
+Similarly, I counted the use of different data strategies for processing data in all articles that reported measuring breath-by-breath.
 Total interval duration of averaging procedures were derived from the reported parameters.
@@
 
-To determine the influence the most common data processing strategies have on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from running.
+To determine the influence the most common data processing strategies have on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from ramp tests in running.
 
@@
 
-A total of N = 72 exercise test were analysed for this study.
+A total of N = 72 exercise tests were analysed for this study.
 Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
 The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
-The tested individuals were experienced distance runners (15 female, 55 male; three of the males participated in both studies).
-The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using exactly the same exercise protocol.
-Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up.
-After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds.
+The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies).
+The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using an identical exercise protocol.
+Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for ten minutes at a velocity of 2.8 m·s^-1^ as a warm-up.
+After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds.
 The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.
@@
 The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement.
-The measured breath-by-breath data is available on GitHub (LINKXXX).
+The measured breath-by-breath data is available on [GitHub](https://github.com/smnnlt/vo2max-processing/tree/master/data/ramptests).
 
@@
 
-The spiro Package Version XXX for R [@nolte2022a] processed the raw gas exchange data.
+The spiro Package Version XXX for R [@nolte2022] processed the raw gas exchange data.
 It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data.
-Table 4 displays a brief summary of exemplary functions and parameters used for the analysis.
-The full analysis script is available at XXXLINK.
-
-```{r spiro-function, echo = FALSE}
-spiro_functions <- data.frame(
-  proc_type = c("Moving Time Average", "Binned Time Average", "Multiple Binned Time Average", "Moving Breath Average", "Digital Filtering"),
-  func = c("`spiro_max()`", "`spiro_summary()`", "`spiro_summary(); filter()`", "`spiro_max()`", "`spiro_max()`"),
-  parameter = c("`smooth = 30`", "XXX", "XXX", '`smooth = "10b"`', '`smooth = "XXX"')
-)
-
-knitr::kable(spiro_functions)
-```
+The full analysis script is available on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md).
 
@@
 
+Moving time based averages were calculated by initially linearly interpolating the breath-by-breath data to seconds.
+Subsequently a (center aligned) moving average was calculated over a defined timespan.
+These processing steps are implemented in the spiro package [@nolte2022].
+For calculating the moving average over 30 seconds for example, I used the functions `spiro(data) |> spiro_max(30)`.
+
+Binned time averages were calculated using a custom function (available in the [analysis script](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md) on GitHub).
+Breath-by-breath data was initially interpolated to full seconds and then binned into consecutive interval of a constant length.
+The average of each interval was aligned to its center.
+Incomplete intervals (i.e.
+the last seconds of measurement) were not considered in the analysis.
+Note that some authors use a different procedure for determining their bins, starting by the end point of the measurement.
+However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software.
+
 #### Breath based averaging
 
+Breath based moving averages were calculated on the raw data.
+As this functionality is implemented in the spiro package, I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.
+
 #### Butterworth filtering
+
+Butterworth filters are a class of recursive digital filters.
+@robergs2010 and @weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data.
+However @robergs2010 missed to account for the time lag introduced by Butterworth filters.
+Therefore I applied a zero-phase Butterworth filter by means of a forward-backward filtering.
+While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it can non be used online (i.e.
+in real time).
+However, for the present application, an offline filter is sufficient.
+The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996].
+I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipi.signal.filtfilt()` function [@virtanen2020]).
+I used 3 as the order parameter and 0.04 as the cut-off frequency for each filter [@robergs2010].
+Note that due to the forward-backward approach both the overall order and overall cut-off frequency are different from these values.
+The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.
 
@@
 
-In an additional exploratory I investigated the respiratory rate (number of breathes per minute) during the ramp tests.
+In an additional exploratory analysis I investigated the respiratory rate (number of breaths per minute) during the ramp tests.
 This may help to understand how breath-based and time-based data processing methods relate to each other.

### 2022-05-03 thesis/results.Rmd
Improve thesis methods and results

@@
 Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in Figure 1).
-Out of the random sample (n = 500), 244 articles were included in the final analysis.
+Out of the random sample (n = 500), 242 articles were included in the final analysis.
 
@@
 Reporting practices of the methodology of gas exchange measures differed widely across the literature (see Table 4).
-More than half (51.8%) of the articles did not report any information on their data processing strategy.
+More than half (51.8%) of the articles did not report any information regarding their data processing strategy.
 One in twenty articles (5.4%) provided a rationale for their used strategy.
@@
 
-Out of the authors that provided information and collected breath-by-breath measurements most (79.5%) utilized binned averages to determine V̇O~2max~.
+Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~.
 Moving time averages, or breath-based averages were uncommon (see Figure 2).
-No study used digital filtering methods to determine V̇O~2max~.
+No study used methods of digital filtering to determine V̇O~2max~.
 
@@
 
-Data preprocessing
+For preprocessing, some authors reported the use of a (linear) interpolation for the breath-by-breath data to seconds (n = 7; 4.3%).
+A minority of researchers reported the use of data filtering strategies to remove outliers.
+This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5).
+For the automated outlier detection authors removed single data points differing from a local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval.
+When reported, the software used for data processing varied among studies showing a total of more than 15 reported programs (for 30 studies that reported this parameter).
 
 The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3).
-30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20s) and longer (60s) periods.
+30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
 
@@
 
-The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 (mean ± standard deviation).
+The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
 Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see Figure 4).
@@
 
-Binned time averages systematically lead to lower V̇O~2max~ values compared to moving averages.
-Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~-values.
+Binned time averages systematically generate V̇O~2max~ values compared to moving averages.
+Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.
 Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
 
-Moving time and moving breath averages yielded nearly identical values for V̇O~2max~ over all calculation intervals.
+Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals.
 This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see Figure 4).

### 2022-05-03 thesis/methods.Rmd
Add appendices

@@
 The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020].
-Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A).
+Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A.1).
 Major deviations will also be explicitly stated within the methods section.
 All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).
+
+I conducted all analyses using R Version 4.1.2 [@rcoreteam2022].
+The thesis was entirely written in R Markdown Version 1.1 [@allaire2022].
+The attached packages and default settings are documented in Appendix A.4.
 
@@
 
-The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix B contains the corresponding reporting checklist.
+The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A.2 contains the corresponding reporting checklist.
 
@@
 The abstracts from the articles included in the random sample were blinded for scanning.
-This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix C for an example).
+This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example).
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.

### 2022-05-03 transparent_changes.Rmd
Add appendices

@@
 ---
-title: "Transparent Changes"
-author: Simon Nolte
-output: 
-  pdf_document: 
-    latex_engine: xelatex
-mainfont: Arial
-urlcolor: blue
+title: ''
+output: pdf_document
 bibliography: references.bib
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-options(tinytex.verbose = TRUE)
-```
-
-# Supplement to the project 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
-
 This document includes all deviation and modifications of code and methods in the final project compared to its [preregistration](https://osf.io/3am4s).
 
-## Major Changes
+### Major Changes
 
-### Number of exercise tests for comparison
+#### Number of exercise tests for comparison
 
@@
 
-## Minor Changes
+### Minor Changes
 

### 2022-05-10 thesis/introduction.Rmd
Complete introduction draft

--- 2022-03-15 thesis/introduction.Rmd
@@
 
-<!-- VO2max is important -->
-<!-- physiological determinants -->
-<!-- VO2max testing -->
-<!-- Influence factors in testing -->
+### Maximum Oxygen Uptake: Definition and Relevance
+
+Performance in endurance sports is limited by the human physiology.
+Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation.
+A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984].
+On a whole-body level the highest activity of oxidative phosphorylation can be approximated by measuring the maximum oxygen uptake (V̇O~2max~).
+
+V̇O~2max~ is defined as the highest rate a body can consume oxygen.
+Determined by measuring gas exchange data, it is one of the most common assessed exercise parameters in sports science and medicine.
+V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological parameter for predicting endurance performance [@bassett2000].
+Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~
+
+### Measuring the Maximum Oxygen Uptake
+
+Researchers measure V̇O~2max~ during exercise tests to exhaustion.
+In a laboratory setting such tests commonly take place on treadmills or cycling ergometer.
+Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value.
+Therefore most exercise protocols consist of a continuous or stepwise, fast increase in load.
+@buchfuhrer1983 and @yoon2007 suggested an optimal duration of 8-12 or 8-10 minutes for testing V̇O~2max~.
+While this narrow limits have been challenged, considerably shorter or longer protocols are advised against [@midgley2008].
+Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.
+
+Early research in V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017].
+Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001].
+While mixing chambers are regarded as more reliable [@beijst2013], measuring breath-by-breath generates data with a higher temporal resolution [@roecker2005].
+The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994].
+But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently.
+
+### Peak vs. Maximum Oxygen Uptake
+
+Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018].
+To identify a 'true' maximum, researchers commonly evaluate a set of parameters measured during the ramp test --- the so called 'V̇O~2max~ criteria' [@howley1995].
+If these criteria are (partly) not fulfilled, researchers are advised to speak of a peak oxygen uptake instead of V̇O~2max~ [@howley1995].
+In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ (e.g.
+the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do by themselves heavily rely on the data processing strategy used [@astorino2009].
+I thus speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.
 
@@
 
-<!-- First example in 90s: applied research -->
-<!-- Since then: various example -->
-<!-- conceptual work: Robergs -->
-<!-- Comparison: martin-rincon -->
-<!-- Problem: group + individual -->
-<!-- Knowledge Gap -->
+Measured breath-by-breath data is noisy (see grey points in Figure 1).
+Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake.
+For interpretation the raw data requires some form of processing.
+
+Different data processing strategies influence measured parameters of gas exchange.
+As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
+Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998; @astorino2009; @sell2021].
+@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that no optimal strategy exists in terms of reliability, but that consistency of strategies is key when comparing outcomes.
+Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
+There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
+
+![Raw breath by breath data processed by different strategies during the final minutes of a ramp test to exhaustion.
+Grey points display oxygen uptake from the single breaths.
+The moving averages were calculated over 30 breaths and seconds, respectively.
+The binned average was calculated over 30-second intervals with the mean values aligned to the center of each interval.
+For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied.
+For more details on the used data processing strategies, see the methods section.](../plots/example.png)
+
+Differences in V̇O~2max~ due to data processing can cause serious implications in practice.
+They hinder the comparability of data from studies within meta-analysis [@martin-rincon2019] or the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing.
+Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion of V̇O~2max~ [@astorino2009].
+Crucially, in situation where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].
+
+Calls to standardize data processing strategies of gas exchange data are frequent.
+@howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~.
+Based on synthetic and experimental data, @robergs2003a opposed these conclusion, stating that shorter calculation intervals are less erroneous.
+They further argued to use breath-based moving averages instead of time averages.
+In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003].
+In a classic opinion piece, @robergs2010 argued for using digital filtering (i.e.
+a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004.
+However, @robergs2010 acknowledged the lack of accessible software implementing such procedures.
+
+In absence of one commonly accepted method, data processing varies among the literature.
+In light of the influence on outcome variables, @myers1990 highlighted the need to report processing strategies in research articles.
+@midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.
+They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively.
+One third of the studies did not describe their processing method at all.
+While providing interesting first insights into reporting and processing practices, the search by @midgley2007 was not systematic and its methods were not described in detail.
+@robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*.
+They provided the latter one with a total of 75 respondents, which reported a great variety in data processing strategies.
+Most researchers reported the use of binned time averages over 30 or 60 seconds.
+Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning.
+Practices in reporting and processing may have changed more than a decade after the publication of such numbers.
 
@@
 
-<!-- Gap -> we need this: -->
-<!-- overview what is actually done -->
-<!-- and how does this impact VO2max assessment -->
+In this thesis I will review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~.
 
-<!-- Review: mapping current methodological (+ reporting) state  -->
-<!-- Comparison: highlight advantage over previous -->
+To date, the practices of data processing to determine V̇O~2max~ in the actual scientific work remain largely unknown.
+Previous research on this topic has only performed unsystematic searches [@midgley2007] or non-representative surveys [@robergs2010].
+Moreover, recent recommendation [@robergs2010] as well as advances in measurement devices and analysis software may have changed processing practices in the past 15 years.
 
-<!-- help compare results across studies + for individuals -->
-<!-- advise for use of methods  -->
-<!-- call for improved reporting -->
+Selected data processing strategies have been extensively compared in the literature.
+Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare.
+Many studies extensively compared different averaging intervals, but not averaging types (e.g.
+moving breath vs.
+binned time) [@astorino2009; @midgley2007].
+@martin-rincon2019 based their comparisons on a data set of sedentary or recreational trained athletes using two different metabolic carts.
+Therefore motivation and measurement devices may have interacted with the influence of processing strategies.
+Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies.
+No research has yet compared a variety of systematically derived strategies among a group of well-trained individuals using a standardized measurement set-up.
 
-@bassett2000; @tricco2018
+This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature.
+Secondly, I will compare different data processing strategies in relation to the most common applied on a set of 72 standardized treadmill tests in well-trained individuals.
+The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals.
+The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting.
+The results build a basis for providing new recommendations for data processing and its reporting in determining the V̇O~2max.~

### 2022-05-10 thesis/discussion.Rmd
Update thesis pdf

--- 2022-03-15 thesis/discussion.Rmd
@@
 # Discussion
+
+The collected data shows, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained.
+Many articles only incompletely report on their methods and use outdated strategies.
+These practices hinder the validity and reproducibility of V̇O~2max~ measurement.
+
+## Current state of data processing
+
+Despite the call to use moving averages [@robergs2010], binned time averages remain the most commonly used in the reviewed literature.
+<!-- compare with midgley? differences? why? highlight systematic aspect? -->
+<!-- ...different journals, not self-assessed -->
+<!-- explain why binned leads to lower vo2max - see fig 1 -->
+<!-- Why do researchers use binned? Not aware of problems? -->
+<!-- binned common for mixing (but not necessary!). Removes major advantages of bbb -->
+<!-- not reporting at all - compare number -->
+
+<!-- length of calculation interval: very important (example), but often differing -->
+<!-- may be due to opposing recommendations -->
+<!-- rational for both short and long, reporting even more important -->
+
+<!-- project allows to compare data among strategies -->
+<!-- but: very individual (and most research only on group mean) -->
+<!-- : for comparison of processing strategies in individuals simply reporting is not enough -->
+
+<!-- recommendation for choosing a strategy: digital filter! -->
+<!-- but zero phase (not as robergs) -->
+<!-- practical issue: requires advanced software -->
+
+<!-- guidelines for reporting: report! -->
+<!-- checklist -->
+
+<!-- even better: share data and code (reporducible analysis) -->
+<!-- data is easy to anonymize -->
+<!-- tools have been recently developed (whippr, spiro) -->

### 2022-05-10 thesis/conclusion.qmd
Move thesis from R Markdown to Quarto

---
title: ''
format: pdf
bibliography: ../references.bib
---

# Conclusion

### 2022-05-10 thesis/introduction.qmd
Move thesis from R Markdown to Quarto

--- 2022-05-10 thesis/introduction.Rmd
@@
 title: ''
-output: pdf_document
+format: pdf
 bibliography: ../references.bib
@@
 
-Measured breath-by-breath data is noisy (see grey points in Figure 1).
+Measured breath-by-breath data is noisy (see grey points in @fig-example).
 Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake.
@@
 For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied.
-For more details on the used data processing strategies, see the methods section.](../plots/example.png)
+For more details on the used data processing strategies, see the methods section.](../plots/example.png){#fig-example}
 

### 2022-05-10 thesis/methods.qmd
Move thesis from R Markdown to Quarto

--- 2022-05-03 thesis/methods.Rmd
@@
 
-I conducted all analyses using R Version 4.1.2 [@rcoreteam2022].
-The thesis was entirely written in R Markdown Version 1.1 [@allaire2022].
+I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022].
+The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a] .
 The attached packages and default settings are documented in Appendix A.4.
@@
 The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords.
-Table 1 shows the exact search terms used.
+@tbl-search shows the exact search terms used.
 
-```{r terms, echo = FALSE}
-out <- data.frame(
-  source = c("PubMed", "Web of Science"),
-  term = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
-)
+| Source         | Search String                                                                                                                                                                                                            |
+|----------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
+| PubMed         | `((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))` |
+| Web of Science | `(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`                              |
 
-knitr::kable(out, 
-  col.names = c("Source", "Search String"), 
-  caption = "Search strings for the systematic scoping review."
-)
-```
+: Search strings for the systematic scoping review. {#tbl-search}
 
@@
 After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
-Table 2 displays the exclusion terms for the automated title screening.
+@tbl-titlescan displays the exclusion terms for the automated title screening.
 
-```{r titlescan, echo = FALSE}
-out <- data.frame(
-  term = c("review, correction, meta-analysis, comment, retraction, editorial, erratum, reply")
-)
+| Search terms for exclusion by title                                               |
+|-----------------------------------------------------------------------------------|
+| review, correction, meta-analysis, comment, retraction, editorial, erratum, reply |
 
-knitr::kable(out, 
-  col.names = c("Search terms for exclusion by title"), 
-  caption = "Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded.")
-```
+: Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded. {#tbl-titlescan}
 
@@
 This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example).
-Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3.
+Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion.
 When the screeners disagreed in their assessments, they resolved the conflict by discussion.
 
-```{r exclusion, echo = FALSE}
-exclusion_criteria <- data.frame(
-  id = c("A*", "B*", "C*", "D*", "E", "F"),
-  criteria = c("not in English", "no primary research", "research not in humans", "VO2max only estimated", "no appropriate test protocol", "no information regarding the exclusion criteria"),
-  details = c("Full text only available in non-English language", "research was no original investigation or only a reanalysis of data", "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", "Protocol for VO2max testing did either not include exercise to voluntial exhaustion or was to long (>20 min) for a reliable estimate", "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing")
-)
+|     | Criterium                                       | Details                                                                                                                                 |
+|-----|-------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|
+| A\* | not in English                                  | Full text only available in non-English language                                                                                        |
+| B\* | no primary research                             | research was no original investigation or only a reanalysis of data                                                                     |
+| C\* | research not in humans                          | research was conducted in animals                                                                                                       |
+| D\* | V̇O~2max~ only estimated                         | V̇O~2max~ was only approximated by means of a predictive equation                                                                        |
+| E   | no appropriate test protocol                    | Protocol for V̇O~2max~ testing did either not include exercise to voluntial exhaustion or was to long (\>20 min) for a reliable estimate |
+| F   | no information regarding the exclusion criteria | crucial information on V̇O~2max~ testing that allowed the evaluation of the other exclusion criteria were missing                        |
 
-knitr::kable(
-  exclusion_criteria,
-  col.names = c("", "Criterium", "Details"),
-  caption = "Exclusion criteria for the screening process.
-During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed."
-)
-```
+: Exclusion criteria for the screening process.
+During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed.
+{#tbl-exclusion}
 
 After the abstract screening I retrieved the full texts for all articles remaining in the review.
-The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the detailed full-text exclusion criteria).
+The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria).
 Conflicts were resolved by discussion.

### 2022-05-10 thesis/results.qmd
Move thesis from R Markdown to Quarto

--- 2022-05-03 thesis/results.Rmd
@@
 
-Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in Figure 1).
+Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in @fig-flow).
 Out of the random sample (n = 500), 242 articles were included in the final analysis.
 
-![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png)
+![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png){#fig-flow}
 
-Reporting practices of the methodology of gas exchange measures differed widely across the literature (see Table 4).
+Reporting practices of the methodology of gas exchange measures differed widely across the literature (see @tbl-reporting).
 More than half (51.8%) of the articles did not report any information regarding their data processing strategy.
@@
 
-```{r reporting-table, echo=FALSE}
+```{r}
+#| echo: false
+#| label: tbl-reporting
+#| tbl-cap: "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing.
+*only examined within the subgroup of studies using breath-by-breath measurements"
 load("../data/reporting.Rda")
@@
   col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
-  caption = "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing.
-*only examined within the subgroup of studies using breath-by-breath measurements",
   align = "r"
@@
 Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~.
-Moving time averages, or breath-based averages were uncommon (see Figure 2).
+Moving time averages, or breath-based averages were uncommon (see @fig-strategies).
 No study used methods of digital filtering to determine V̇O~2max~.
 
-![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png)
+![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png){#fig-strategies}
 
@@
 
-The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3).
+The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
 
-![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png)
-
-## Experimental Comparison
-
-The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
-Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see Figure 4).
-
-![V̇O~2max~ varies by data processing strategy.
-Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review.
-Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
-Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
-Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png)
-
-Binned time averages systematically generate V̇O~2max~ values compared to moving averages.
-Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.
-Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
-
-Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals.
-This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see Figure 4).
-
-![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png)
+![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png){#fig-duration}

### 2022-05-10 thesis/thesis.qmd
Move thesis from R Markdown to Quarto

---
title: " "
format: pdf
number-sections: true
top-level-division: section
mainfont: Arial
fontsize: 11pt
urlcolor: blue
linkcolor: black
link-citations: yes
bibliography: ../references.bib
csl: '`r here::here("apa.csl")`'
geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
params:
  title: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
  name: Simon Nolte
  university: German Sport University Cologne
  year: 2022
---

\begin{centering}

\vspace*{2cm}

\LARGE
{`r params$title`}

\vspace{5cm}

\LARGE
{Bachelor thesis}

\LARGE
{from}

\vspace{2cm}

\LARGE
{`r params$name`}

\vspace{6cm}

\LARGE
{`r params$university`}

\Large
{`r paste0("Cologne ", params$year)`}

\end{centering}

\pagenumbering{gobble}

\newpage

Thesis supervisor:

Dr. Oliver Jan Quittmann

Institute of Movement and Neurosciences

\newpage

Affirmation in lieu of an oath

Herewith I affirm in lieu of an oath that I have authored this Master thesis independently and did not use any other sources and tools than indicated. All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named. The same is true for tables and figures. I did not submit this piece of work in the same or similar way or in extracts in another assignment.

\_____________________________

Personally signed

\newpage

# Zusammenfassung (German abstract) {-}

\newpage

\pagenumbering{roman}

# Table of contents {-}

\renewcommand{\contentsname}{}

\tableofcontents

\newpage

\renewcommand{\listfigurename}{}

# List of Figures {-}

\listoffigures

\newpage

\renewcommand{\listtablename}{}

# List of Tables {-}

\listoftables

\newpage

\pagenumbering{arabic}

```{r}
#| child = 'introduction.qmd'
```

\newpage

```{r}
#| child = 'methods.qmd'
```

\newpage

```{r}
#| child = 'results.qmd'
```

\newpage

```{r}
#| child = 'discussion.qmd'
```

\newpage

```{r}
#| child = 'conclusion.qmd'
```

\newpage

# Bibliography

<div id="refs"></div>

\newpage

\appendix

# Appendix

## Transparent Changes

\newpage

## Prisma Reporting Checklist

\newpage

## Blinded abstract example

![](../plots/abstracts/030.png)

\newpage

## Technical Details

### Session info

```{r session}
sessionInfo()
```

### Used Packages

```{r}
p_used <- unique(renv::dependencies(path = "../")$Package)
p_inst <- as.data.frame(installed.packages())
out <- p_inst[p_inst$Package %in% p_used, c("Package", "Version")]
rownames(out) <- NULL
out
```

### 2022-05-11 thesis/results.qmd
Add missing results part

@@
 ![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png){#fig-duration}
+
+## Experimental Comparison
+
+The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
+Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see @fig-comparison).
+
+![V̇O~2max~ varies by data processing strategy.
+Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review.
+Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
+Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
+Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png){#fig-comparison}
+
+Binned time averages systematically generate V̇O~2max~ values compared to moving averages.
+Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.
+Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
+
+Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals.
+This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see @fig-rr).
+
+![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png){#fig-rr}

### 2022-05-12 thesis/discussion.qmd
Move thesis from R Markdown to Quarto
Draft discussion

--- 2022-03-15 thesis/discussion.Rmd
@@
 # Discussion
+
+The collected data shows, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained.
+Many articles only incompletely report on their methods and use outdated strategies.
+These practices hinder the validity and reproducibility of V̇O~2max~ measurement.
+
+## Current state of data processing
+
+Despite calls to use moving averages [@robergs2003; @robergs2010], binned time averages remain the most commonly used data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies).
+These numbers are generally in line with the findings of the non-systematic search by @midgley2007 and the survey by @robergs2010.
+It is somewhat surprising, that practices have not changes in recent years despite the publication of recommendations, that highly discourage researchers from using binned averages [@robergs2010].
+Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison).
+The peak in oxygen uptake may fall in between two average intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
+Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
+Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain very common in scientific literature.
+
+Breath-based averages seem to be more common (≈ 8%) than reported previously (\< 1%) [@midgley2007], but less common than assessed in self-reporting (≈ 17%) [@robergs2010].
+The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for them [@robergs2003; @robergs2010].
+However the proportion of breath-based averages is much smaller than that of binned time averages, which have not been recommended in such a way.
+
+The length of the calculation intervals is highly diverse within the literature (see @fig-duration).
+This may reflect contradictory recommendation [@howley1995; @robergs2003].
+As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration.
+As different interval duration can heavily influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
+
+The exact reasons for exercise scientist ignoring most recommendations by using binned time averages remain unknown.
+Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~, but past publications on this issues have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
+Researchers may also use binned averages for traditional reasons.
+Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data appearently similar to those by a binned time average of breath-by-breath data.
+But comparability with older data should only be an issue when using data acquired by different measurement methods within on analysis.
+Current studies using new breath-by-breath data do not reasonably need to rely on outdated methods of data processing.
+
+A major source for choosing suboptimal processing strategies may be limitations by analysis software.
+My review shows that most researchers use the software of the metabolic cart's manufacturer to analyse the gas exchange data.
+These software may by default output binned time averages instead of raw breath-by-breath data.
+Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software.
+This is also an reasonable explanation why digital filtering has --- despite recommended by @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
+The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of easy-to-use software solutions.
+
+## Impact of different data processing
+
+The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such can influence the classification of individuals, evaluation of training success, and assessment of V̇O~2max~ attainment.
+In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison).
+The analysed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research.
+Some studies reported even higher mean differences up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison.
+The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies.
+While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in well-trained athletes.
+
+Binned time averages lead to systematically lower V̇O~2max~ values compared with moving averages, for the reasons explained above.
+While this general trend has been acknowledged previously [@martin-rincon2019], it has not been quantified.
+The data presented in this thesis suggest a ≈-1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
+This difference is well within the measurement error of most if not all metabolic carts, but is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g.
+in elite sports).
+
+Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison).
+This is hardly surprising, as the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths.
+For an athletic population, V̇O~2max~ values obtained by moving time and moving breath average can approximately used interchangeable.
+Given that less trained individual will likely display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.
+
+The exact impact of data processing strategies on the V̇O~2max~ is highly individual.
+Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019].
+On an individual level, data processing strategies may impact the V̇O~2max~ in different severity.
+For example, for 10% of the investigated athletes a binned time average of 5 seconds lead to a V̇O~2max~ \<3% higher than by a 30 second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
+Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classification.
+However on a individual level such work does not help.
+Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some to more than 10% in other individuals.
+Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported.
+While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.
+
+## Guidelines for data processing and reporting
+
+This thesis focussed on the occurrence and impact of different data processing strategies, but did not investigate their validity in the first place.
+However the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for data reporting.
+
+I highly disregard researchers from continuing to use binned time averages to determine V̇O~2max~.
+The main reason for such procedure is pure tradition, as it reduces the breath-by-breath data in an inappropriate way leading to a small, yet systematic, underestimation of V̇O~2max~.
+Moving time or moving breath averages are preferable to binned averages.
+In a athletic population they may be used equivalent, but in sedentary individuals breath-based averages will likely lead to lower V̇O~2max~ values than time-based averages with the same interval length parameter.
+While using a constant breath interval will lead to a similar degree of data smoothing regardless of the training status, using a constant time interval will lead to the same physiological time-frame for determining V̇O~2max~ (but with a different degree of data smoothing).
+From a data processing perspective, I prefer using breath-based moving averages, while keeping in mind, that this may underestimate true V̇O~2max~ in individuals with a low respiratory rate in the end of the exercise test.
+
+Digital filtering seems to be a promising method, as it markedly reduces the variability moving averages have due to single data outliers (see @fig-example).
+I therefore agree with @robergs2010, who recommends digital filters over any classical averaging procedure.
+The exact type of filter and values of filter parameters to be used have not yet been systematically investigated, and it is unclear whether it is possible at all to determine criteria for identifying an optimal filter.
+However, a Butterworth filter, as proposed by @robergs2010, seems to produce reasonably smoothed data.
+Note, that a single Butterworth filter comprises a time lag, a fact highlighted by @weir2004 that was not acknowledged by @robergs2010.
+To allow for a correct phasing of data, as well as for a correct V̇O~2max~ determination when the measurement is shortly terminated after reaching exhaustion, a zero-phase filter is needed.
+A zero-phase forward-backward Butterworth filter (as suggested by @weir2004) seems to produce reasonable smoothing when used with the parameters suggested by @robergs2010 (low-pass cut-off frequency: 0.04 Hz; third order; see @fig-example).
+It should be noted, that despite using the same parameters the degree of filtering varies from that by @robergs2010, as the filter is applied twice, changing the overall magnitude of parameters.
+Other filtering parameters may be as reasonable as this, but it is unclear how to objectively compare them.
+
+I recommend to use a zero-phase forward-backward Butterworth filter on the raw breath-by-breath data.
+The V̇O~2max~ is then defined as the highest single filtered data point.
+An filter order of 3 and a low cut-off frequency of 0.04 Hz appear to be reasonable parameters for each filter.
+While such data processing has advantages over traditional data processing forms, it requires specialized software.
+
+To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategy is crucial.
+The review presented here shows that almost half of the studies that measured V̇O~2max~ did not describe their processing strategy.
+Other aspects of the data processing, such as outlier filtering or rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting).
+@tbl-recommendation lists 7 item that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~.
+These items may be reported in form of a check-list, as an in-text enumeration or in sentence form.
+An example paragraph containing all the relevant information for the current study would be:
+
+> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America).
+The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro Package for R [@nolte2022].
+This strategy produces similar results as that recommended by @robergs2010, but does not produce a time lag."*
+
+| reporting item        | description                                                                                                           |
+|-----------------------|-----------------------------------------------------------------------------------------------------------------------|
+| Metabolic Cart        | State the exact device model and manufacturer                                                                         |
+| Measurement Mode      | State the measurement mode (e.g., mixing chamber, breath-by-breath,...)                                               |
+| Software              | State the name and version of the software used for data analysis                                                     |
+| Preprocessing         | State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis |
+| Processing Strategy   | State the exact data processing strategy used to determine the V̇O~2max~ (e.g., binned time averages)                  |
+| Processing Parameters | State the parameters used for the processing strategy (e.g., length of averaging interval)                            |
+| Rationale             | State the rationale for using the processing strategy (e.g., reference to recommendations)                            |
+
+: Recommendations for reporting data processing strategies to determine V̇O~2max~. {#tbl-recommendation}
+
+Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing.
+Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test.
+In cases where journals endorse word limits on articles, these reporting --- including the reporting on data processing strategies --- may be included in supplementary files.
+The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.
+
+The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
+But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies may vary between individuals and are potentially influenced by training status.
+Sharing of the raw metabolic exchange data is the solution to this problem.
+It allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
+Anonymization of raw gas exchange data files should not be a big deal for researchers, since most of these files are structured in a simple way and allow to easily remove any personal information (if this has not been done in the metabolic cart's system before).
+In terms of reproducibility of V̇O~2max~ determination, an even superior way to correct reporting and raw data sharing is to additionally share data analysis code.
+This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming, or at least code generating, environment.
+Luckily such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].
+
+## Limitations
+
+Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of the literature.
+The scoping review therefore relies on a random sample which not necessarily exactly captures trends of the whole literature.
+However, efforts were made --- such as random sampling and systematic article exclusions --- to ensure the sample to be representative.
+Notable, almost half of the studies did not report their data processing strategy at all.
+The data processing strategies used in the literature could only be described when studies reported them.
+
+Ambiguities in the reporting of the investigated studies may impact the analysis results.
+For example some studies using long binned averages (e.g.
+60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly.
+Moreover the exact definitions for building binned averages varied within the literature: While most studies defined the binning periods from the beginning of the exercise, some may have defined them from the endpoint.
+Additionally some studies reviewed did not define the maximum, but a predefined binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
+In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average.
+I did not seperately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.
+
+The experimental comparison of different data processing strategies was conducted on a standardized data set of exercise tests.
+This standardization in terms of training status, exercise protocol, and measurement device helps to highlight the impact of different data processing even in a relatively homogeneous data set.
+However the results may only partly be transferable to different settings, such as individuals with less training background.
+I did not conduct an formal analysis of the validity or reliability of different data processing strategies, so recommendations regarding their use rely on theoretical derivations and prior research.
+Since the reliability of different strategies appears to be similar [@midgley2007; @martin-rincon2019] and currently no accepted methods to quantify validity of V̇O~2max~ exist, the presented approach is sufficient to derive recommendations for data processing strategies to determine V̇O~2max~.

### 2022-05-12 thesis/introduction.qmd
Add further references to introduction

@@
 
-Calls to standardize data processing strategies of gas exchange data are frequent.
+Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998].
 @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~.
@@
 In absence of one commonly accepted method, data processing varies among the literature.
-In light of the influence on outcome variables, @myers1990 highlighted the need to report processing strategies in research articles.
+In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998].
 @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.

### 2022-05-13 thesis/discussion.qmd
Change layout

@@
 
-| reporting item        | description                                                                                                           |
-|-----------------------|-----------------------------------------------------------------------------------------------------------------------|
-| Metabolic Cart        | State the exact device model and manufacturer                                                                         |
-| Measurement Mode      | State the measurement mode (e.g., mixing chamber, breath-by-breath,...)                                               |
-| Software              | State the name and version of the software used for data analysis                                                     |
-| Preprocessing         | State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis |
-| Processing Strategy   | State the exact data processing strategy used to determine the V̇O~2max~ (e.g., binned time averages)                  |
-| Processing Parameters | State the parameters used for the processing strategy (e.g., length of averaging interval)                            |
-| Rationale             | State the rationale for using the processing strategy (e.g., reference to recommendations)                            |
+```{r}
+#| echo: false
+#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake"
+#| label: tbl-recommendation
 
-: Recommendations for reporting data processing strategies to determine V̇O~2max~. {#tbl-recommendation}
+rec_df <- data.frame(
+  item = c("Metabolic Cart", "Measurement Mode", "Software", "Preprocessing", "Processing Strategy", "Processing Parameters", "Rationale"),
+  desc = c(
+    "State the exact device model and manufacturer",
+    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…)",
+    "State the name and version of the software used for data analysis",
+    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis",
+    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages)",
+    "State the parameters used for the processing strategy (e.g., length of averaging interval)",
+    "State the rationale for using the processing strategy (e.g., reference to recommendations)")
+)
+
+out <- kableExtra::kbl(
+  rec_df,
+  col.names = c("Reporting Item", "Description"),
+  format = "latex",
+  booktabs = TRUE
+) |>
+   kableExtra::column_spec(
+    2, width = "11cm"
+  )
+
+knitr::asis_output(stringr::str_replace_all(out, "VO2max", "${\\\\dot{\\\\textsf{V}}\\\\textsf{O}_{\\\\textsf{2max}}}$"))
+```
 

### 2022-05-13 thesis/introduction.qmd
Change layout

@@
 
-### Maximum Oxygen Uptake: Definition and Relevance
-
 Performance in endurance sports is limited by the human physiology.
@@
 A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984].
-On a whole-body level the highest activity of oxidative phosphorylation can be approximated by measuring the maximum oxygen uptake (V̇O~2max~).
+The highest activity of oxidative phosphorylation on a cellular level can be approximated by measuring the maximum oxygen uptake (V̇O~2max~) on a whole-body level.
 
-V̇O~2max~ is defined as the highest rate a body can consume oxygen.
+V̇O~2max~ is defined as the highest rate a body can consume oxygen [@bassett2000].
 Determined by measuring gas exchange data, it is one of the most common assessed exercise parameters in sports science and medicine.
@@
 
-### Measuring the Maximum Oxygen Uptake
-
-Researchers measure V̇O~2max~ during exercise tests to exhaustion.
+Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion.
 In a laboratory setting such tests commonly take place on treadmills or cycling ergometer.
 Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value.
-Therefore most exercise protocols consist of a continuous or stepwise, fast increase in load.
-@buchfuhrer1983 and @yoon2007 suggested an optimal duration of 8-12 or 8-10 minutes for testing V̇O~2max~.
-While this narrow limits have been challenged, considerably shorter or longer protocols are advised against [@midgley2008].
+Therefore most exercise protocols consist of a continuous or stepwise increase with an optimal during around 8-12 minutes [@buchfuhrer1983 ; @yoon2007].
+While this narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008].
 Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.
 
-Early research in V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017].
+Early research on V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017].
 Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001].
@@
 
-### Peak vs. Maximum Oxygen Uptake
-
 Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018].
-To identify a 'true' maximum, researchers commonly evaluate a set of parameters measured during the ramp test --- the so called 'V̇O~2max~ criteria' [@howley1995].
+To identify a true maximum, researchers commonly evaluate a set of parameters measured during ramp tests --- the so called 'V̇O~2max~ criteria' [@howley1995].
 If these criteria are (partly) not fulfilled, researchers are advised to speak of a peak oxygen uptake instead of V̇O~2max~ [@howley1995].
 In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ (e.g.
-the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do by themselves heavily rely on the data processing strategy used [@astorino2009].
-I thus speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.
+the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do heavily rely on the data processing strategy used [@astorino2009].
+Thus, I speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.
 
@@
 Measured breath-by-breath data is noisy (see grey points in @fig-example).
-Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake.
+Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) lead to a highly fluctuating raw oxygen uptake.
 For interpretation the raw data requires some form of processing.
@@
 As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
-Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998; @astorino2009; @sell2021].
-@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that no optimal strategy exists in terms of reliability, but that consistency of strategies is key when comparing outcomes.
+Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021].
+@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
 Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
@@
 
-![Raw breath by breath data processed by different strategies during the final minutes of a ramp test to exhaustion.
+```{r}
+#| echo: false
+#| fig-cap: "Example of raw breath-by-breath data processed by different strategies during the final minutes of a ramp test to exhaustion.
 Grey points display oxygen uptake from the single breaths.
@@
 For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied.
-For more details on the used data processing strategies, see the methods section.](../plots/example.png){#fig-example}
+For more details on the used data processing strategies, see the methods section.
+Data Source: [@quittmann_unpubl_b]"
+#| fig-scap: "Example of raw breath-by-breath data processed by different strategies during the final minutes of a ramp test to exhaustion."
+#| label: fig-example
+#| out-width: '100%'
+
+knitr::include_graphics("../plots/example.png")
+```
 
@@
 @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~.
-Based on synthetic and experimental data, @robergs2003a opposed these conclusion, stating that shorter calculation intervals are less erroneous.
+Based on synthetic and experimental data, @robergs2003 opposed these conclusion, stating that shorter calculation intervals are less erroneous.
 They further argued to use breath-based moving averages instead of time averages.
@@
 In this thesis I will review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~.
-
 To date, the practices of data processing to determine V̇O~2max~ in the actual scientific work remain largely unknown.
@@
 Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare.
-Many studies extensively compared different averaging intervals, but not averaging types (e.g.
+Many studies compared different averaging intervals, but not averaging types (e.g.
 moving breath vs.
 binned time) [@astorino2009; @midgley2007].
-@martin-rincon2019 based their comparisons on a data set of sedentary or recreational trained athletes using two different metabolic carts.
+@martin-rincon2019 provided formula for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
 Therefore motivation and measurement devices may have interacted with the influence of processing strategies.
+
 Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies.
@@
 This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature.
-Secondly, I will compare different data processing strategies in relation to the most common applied on a set of 72 standardized treadmill tests in well-trained individuals.
+Secondly, I will compare different data processing strategies in relation to the most commonly applied on a set of 72 standardized treadmill tests in well-trained individuals.
 The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals.
 The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting.
-The results build a basis for providing new recommendations for data processing and its reporting in determining the V̇O~2max.~
+The results build a basis for providing new recommendations for data processing and its reporting for determining the V̇O~2max.~

### 2022-05-13 thesis/methods.qmd
Change layout

@@
 format: pdf
+crossref:
+  sec-prefix: appendix
 bibliography: ../references.bib
@@
 The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020].
-Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A.1).
+Any deviations from the preregistration are indicated in the 'Transparent Changes' document (@sec-tchanges).
 Major deviations will also be explicitly stated within the methods section.
@@
 I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022].
-The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a] .
-The attached packages and default settings are documented in Appendix A.4.
+The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a].
+The attached packages and default settings are documented in @sec-tech.
 
@@
 Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
-
-The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A.2 contains the corresponding reporting checklist.
+The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018a].
+@sec-checklist contains the corresponding reporting checklist.
 
@@
 
-| Source         | Search String                                                                                                                                                                                                            |
-|----------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
-| PubMed         | `((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))` |
-| Web of Science | `(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`                              |
+```{r}
+#| echo: false
+#| tbl-cap: "Search strings for the systematic scoping review."
+#| label: tbl-search
 
-: Search strings for the systematic scoping review. {#tbl-search}
+search_df <- data.frame(
+    s = c("PubMed", "Web of Science"),
+    B = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
+)
+
+kableExtra::kbl(
+  search_df,
+  format = "latex",
+  col.names = c("Source", "Search Strings"),
+  booktabs = TRUE
+) |> 
+  kableExtra::column_spec(
+    2, width = "12cm"
+  )
+```
 
@@
 After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
-@tbl-titlescan displays the exclusion terms for the automated title screening.
-
-| Search terms for exclusion by title                                               |
-|-----------------------------------------------------------------------------------|
-| review, correction, meta-analysis, comment, retraction, editorial, erratum, reply |
-
-: Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded. {#tbl-titlescan}
+All titles that contained one of the following words were excluded: review, correction, meta-analysis, comment, retraction, editorial, erratum, reply.
 
@@
 The abstracts from the articles included in the random sample were blinded for scanning.
-This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example).
+This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see @sec-blindabstract for an example).
 Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion.
@@
 
-|     | Criterium                                       | Details                                                                                                                                 |
-|-----|-------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|
-| A\* | not in English                                  | Full text only available in non-English language                                                                                        |
-| B\* | no primary research                             | research was no original investigation or only a reanalysis of data                                                                     |
-| C\* | research not in humans                          | research was conducted in animals                                                                                                       |
-| D\* | V̇O~2max~ only estimated                         | V̇O~2max~ was only approximated by means of a predictive equation                                                                        |
-| E   | no appropriate test protocol                    | Protocol for V̇O~2max~ testing did either not include exercise to voluntial exhaustion or was to long (\>20 min) for a reliable estimate |
-| F   | no information regarding the exclusion criteria | crucial information on V̇O~2max~ testing that allowed the evaluation of the other exclusion criteria were missing                        |
+```{r}
+#| echo: false
+#| tbl-cap: "Exclusion criteria for the screening process."
+#| label: tbl-exclusion
 
-: Exclusion criteria for the screening process.
-During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed.
-{#tbl-exclusion}
+excl_df <- data.frame(
+    id = c("A*", "B*", "C*", "D*", "E", "F"),
+    crit = c(
+      "not in English", "no primary research", "research not in humans", "VO2max only estimated",
+      "no appropriate test protocol", "no information regarding the exclusion criteria"),
+    info = c(
+      "Full text only available in non-English language",
+      "research was no original investigation or only a reanalysis of data", 
+      "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", 
+      "Protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
+      "no information regarding the exclusion criteria | crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
+      )
+)
 
-After the abstract screening I retrieved the full texts for all articles remaining in the review.
-The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria).
-Conflicts were resolved by discussion.
+out <- kableExtra::kbl(
+  excl_df,
+  format = "latex",
+  col.names = c("", "Criterion", "Details"),
+  booktabs = TRUE
+) |> 
+    kableExtra::column_spec(
+    2, width = "4cm"
+  ) |>
+    kableExtra::column_spec(
+    3, width = "9cm"
+  ) |>
+  kableExtra::footnote(
+    general = "*During the abstract screening only the criteria marked with an asterix were evaluated",
+    general_title = ""
+  )
 
+knitr::asis_output(stringr::str_replace_all(out, "VO2max", "${\\\\dot{\\\\textsf{V}}\\\\textsf{O}_{\\\\textsf{2max}}}$"))
+```
+
+After the abstract screening I retrieved the full-texts for all remaining articles.
+The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria).
+Conflicts were resolved by discussion between examiners.
 All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).
@@
 
-Gas exchange data were recorded using a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America).
+Gas exchange data were recorded using a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America).
 The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement.
@@
 
-#### Time based averaging
-
-Moving time based averages were calculated by initially linearly interpolating the breath-by-breath data to seconds.
+Moving time based averages were calculated by first linearly interpolating the breath-by-breath data to seconds.
 Subsequently a (center aligned) moving average was calculated over a defined timespan.
@@
 Note that some authors use a different procedure for determining their bins, starting by the end point of the measurement.
-However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software.
-
-#### Breath based averaging
+However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software (e.g.
+Cosmed Omnia).
 
@@
 As this functionality is implemented in the spiro package, I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.
-
-#### Butterworth filtering
 
@@
 Therefore I applied a zero-phase Butterworth filter by means of a forward-backward filtering.
-While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it can non be used online (i.e.
+While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it cannot be used online (i.e.
 in real time).
@@
 The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996].
-I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipi.signal.filtfilt()` function [@virtanen2020]).
+I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipy.signal.filtfilt()` function [@virtanen2020]).
 I used 3 as the order parameter and 0.04 as the cut-off frequency for each filter [@robergs2010].
@@
 I calculated the data for all integer parameter values within the range of the values found in the literature during the review.
-On a group level I calculated the median and 10%- and 90%-quantiles of each processing strategy.
+On a group level, I calculated the median and 10%- and 90%-quantiles of each processing strategy.
 

### 2022-05-13 thesis/results.qmd
Change layout

@@
 
-![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png){#fig-flow}
+```{r}
+#| echo: false
+#| fig-cap: "Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]"
+#| fig-scap: "Flow diagram for the systematic scoping review"
+#| label: fig-flow
+#| out-width: '90%'
+
+knitr::include_graphics("../plots/flow_diagram.png")
+```
 
@@
 #| label: tbl-reporting
-#| tbl-cap: "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing.
-*only examined within the subgroup of studies using breath-by-breath measurements"
+#| tbl-cap: "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing."
 load("../data/reporting.Rda")
@@
 
-knitr::kable(
+kableExtra::kbl(
   t,
   col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
-  align = "r"
+  align = "r",
+  booktabs = TRUE
+) |>
+  kableExtra::footnote(
+    general = "*only examined within the subgroup of studies using breath-by-breath measurements",
+    general_title = ""
   )
@@
 
-![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png){#fig-strategies}
+```{r}
+#| echo: false
+#| fig-cap: "Data strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
+#| label: fig-strategies
+#| out-width: '90%'
+
+knitr::include_graphics("../plots/strategies.png")
+```
 
@@
 
-![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png){#fig-duration}
+```{r}
+#| echo: false
+#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies."
+#| label: fig-duration
+#| out-width: '90%'
+
+knitr::include_graphics("../plots/duration_count.png")
+```
 
@@
 
-![V̇O~2max~ varies by data processing strategy.
-Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review.
+```{r}
+#| echo: false
+#| fig-cap: "V̇O~2max~ varies by data processing strategy.
+Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review.
 Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
 Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
-Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png){#fig-comparison}
+Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
+#| fig-scap: "Comparison of different data processing strategies"
+#| label: fig-comparison
+#| out-width: '90%'
+
+knitr::include_graphics("../plots/comparison.png")
+```
 
@@
 
-![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png){#fig-rr}
+```{r}
+#| echo: false
+#| fig-cap: "Respiratory rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72)."
+#| fig-scap: "Respiratory rates during the ramp tests"
+#| label: fig-rr
+#| out-width: '90%'
+
+knitr::include_graphics("../plots/rr.png")
+```

### 2022-05-13 thesis/thesis.qmd
Change layout

@@
 title: " "
-format: pdf
+format: 
+  pdf:
+    include-in-header: 
+      text: \usepackage{setspace}
 number-sections: true
@@
 bibliography: ../references.bib
-csl: '`r here::here("apa.csl")`'
+csl: '`r here::here("sports-med-new.csl")`'
 geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
@@
 
+\onehalfspacing
+
 ```{r}
@@
 
-# Bibliography
+# References {.unnumbered}
 
@@
 
-## Transparent Changes
+## Transparent Changes {#sec-tchanges}
 
@@
 
-## Prisma Reporting Checklist
+## Technical Details {#sec-tech}
 
-\newpage
-
-## Blinded abstract example
-
-![](../plots/abstracts/030.png)
-
-\newpage
-
-## Technical Details
-
-### Session info
+### Session Info
 
@@
 
-### Used Packages
+### Packages
 
 ```{r}
+#| message: false
+
 p_used <- unique(renv::dependencies(path = "../")$Package)
@@
 ```
+
+\newpage
+
+## Prisma Reporting Checklist {#sec-checklist}
+
+\newpage
+
+## Blinded Abstract Example {#sec-blindabstract}
+
+![](../plots/abstracts/030.png)

### 2022-05-15 thesis/discussion.qmd
Add information to thesis

@@
 These numbers are generally in line with the findings of the non-systematic search by @midgley2007 and the survey by @robergs2010.
-It is somewhat surprising, that practices have not changes in recent years despite the publication of recommendations, that highly discourage researchers from using binned averages [@robergs2010].
+It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that highly discourage researchers from using binned averages [@robergs2010].
 Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison).
-The peak in oxygen uptake may fall in between two average intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
+The peak in oxygen uptake may fall in between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
 Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
@@
 Breath-based averages seem to be more common (≈ 8%) than reported previously (\< 1%) [@midgley2007], but less common than assessed in self-reporting (≈ 17%) [@robergs2010].
-The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for them [@robergs2003; @robergs2010].
+The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010].
 However the proportion of breath-based averages is much smaller than that of binned time averages, which have not been recommended in such a way.
@@
 As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration.
-As different interval duration can heavily influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
+As different interval durations can heavily influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
 
-The exact reasons for exercise scientist ignoring most recommendations by using binned time averages remain unknown.
-Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~, but past publications on this issues have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
+The exact reasons for exercise scientist to ignore most recommendations by using binned time averages remain unknown.
+Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~, but past publications on this issue have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
 Researchers may also use binned averages for traditional reasons.
-Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data appearently similar to those by a binned time average of breath-by-breath data.
+Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data apparently similar to those by a binned time average of breath-by-breath data.
 But comparability with older data should only be an issue when using data acquired by different measurement methods within on analysis.
@@
 
-The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such can influence the classification of individuals, evaluation of training success, and assessment of V̇O~2max~ attainment.
+The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such can influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment.
 In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison).
@@
 The data presented in this thesis suggest a ≈-1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
-This difference is well within the measurement error of most if not all metabolic carts, but is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g.
+This difference is well within the measurement error of most if not all metabolic carts, but it is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g.
 in elite sports).
@@
 On an individual level, data processing strategies may impact the V̇O~2max~ in different severity.
-For example, for 10% of the investigated athletes a binned time average of 5 seconds lead to a V̇O~2max~ \<3% higher than by a 30 second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
+For example, for 10% of the investigated athletes a binned time average of 5 seconds lead to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
 Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classification.
-However on a individual level such work does not help.
-Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some to more than 10% in other individuals.
+However on an individual level such work does not help.
+Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others.
 Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported.
@@
 Note, that a single Butterworth filter comprises a time lag, a fact highlighted by @weir2004 that was not acknowledged by @robergs2010.
-To allow for a correct phasing of data, as well as for a correct V̇O~2max~ determination when the measurement is shortly terminated after reaching exhaustion, a zero-phase filter is needed.
+To allow for a correct phasing of data, as well as for a correct V̇O~2max~ determination when the measurement is terminated shortly after reaching exhaustion, a zero-phase filter is needed.
 A zero-phase forward-backward Butterworth filter (as suggested by @weir2004) seems to produce reasonable smoothing when used with the parameters suggested by @robergs2010 (low-pass cut-off frequency: 0.04 Hz; third order; see @fig-example).
@@
 
-> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America).
+> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America).
 The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro Package for R [@nolte2022].
@@
 The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
-But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies may vary between individuals and are potentially influenced by training status.
+But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
 Sharing of the raw metabolic exchange data is the solution to this problem.
@@
 Anonymization of raw gas exchange data files should not be a big deal for researchers, since most of these files are structured in a simple way and allow to easily remove any personal information (if this has not been done in the metabolic cart's system before).
-In terms of reproducibility of V̇O~2max~ determination, an even superior way to correct reporting and raw data sharing is to additionally share data analysis code.
-This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming, or at least code generating, environment.
+In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code.
+This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment.
 Luckily such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].
@@
 Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of the literature.
-The scoping review therefore relies on a random sample which not necessarily exactly captures trends of the whole literature.
+The scoping review therefore relies on a random sample which not necessarily captures exact trends of the literature.
 However, efforts were made --- such as random sampling and systematic article exclusions --- to ensure the sample to be representative.
@@
 Ambiguities in the reporting of the investigated studies may impact the analysis results.
-For example some studies using long binned averages (e.g.
-60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly.
+For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly.
 Moreover the exact definitions for building binned averages varied within the literature: While most studies defined the binning periods from the beginning of the exercise, some may have defined them from the endpoint.
@@
 In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average.
-I did not seperately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.
+I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.
 
@@
 This standardization in terms of training status, exercise protocol, and measurement device helps to highlight the impact of different data processing even in a relatively homogeneous data set.
-However the results may only partly be transferable to different settings, such as individuals with less training background.
-I did not conduct an formal analysis of the validity or reliability of different data processing strategies, so recommendations regarding their use rely on theoretical derivations and prior research.
+However the results may only partly transfer to different settings, such as individuals with less training background.
+I did not conduct a formal analysis of the validity or reliability of different data processing strategies, so recommendations regarding their use rely on theoretical derivations and prior research.
 Since the reliability of different strategies appears to be similar [@midgley2007; @martin-rincon2019] and currently no accepted methods to quantify validity of V̇O~2max~ exist, the presented approach is sufficient to derive recommendations for data processing strategies to determine V̇O~2max~.

### 2022-05-15 thesis/methods.qmd
Add information to thesis

@@
 
-I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022].
+I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022].
 The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a].
@@
 
-The spiro Package Version XXX for R [@nolte2022] processed the raw gas exchange data.
+The spiro Package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data.
 It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data.

### 2022-05-15 thesis/results.qmd
Add information to thesis

@@
 Reporting practices of the methodology of gas exchange measures differed widely across the literature (see @tbl-reporting).
-More than half (51.8%) of the articles did not report any information regarding their data processing strategy.
-One in twenty articles (5.4%) provided a rationale for their used strategy.
+Almost half (44.2%) of the articles did not report any information regarding their data processing strategy.
+One in twenty articles (5.8%) provided a rationale for their used strategy.
 Only a single article [@maturana2021] reported information regarding all the investigated criteria.

### 2022-05-17 scripts/review.Rmd
Create additional data summary

@@
 ```
+
+#### Create data of all sampled studies
+
+```{r}
+load("../data/screen/sample.Rda")
+abstract <- read.csv("../data/screen/abstract_screening.csv")
+fulltext <- read.csv("../data/screen/fulltext_screening.csv")
+fullsample <- sampled_data[, c("sid", "title", "authors", "year", "doi")]
+get_status <- function(sid) {
+  if (!(sid %in% fulltext$sid)) {
+    status <- "excluded_abstract"
+    reason <- abstract$decision[sid]
+  } else {
+    if (is.na(fulltext$decision[sid])) {
+      status <- "included"
+      reason <- NA
+    } else {
+      status <- "excluded_fulltext"
+      reason <- fulltext$decision[sid]
+    }
+  }
+  data.frame(status = status, reason = reason)
+}
+fullsample <- cbind(
+  fullsample, 
+  purrr::map_dfr(seq_len(nrow(fullsample)), get_status)
+)
+
+fullsample$reason <- factor(
+  fullsample$reason, 
+  levels = c("t", "l", "r", "h", "m", "c", "e", "d", "o"),
+  labels = c("report not retrieved", "not in English", "no primary research", 
+             "reasearch not in humans", "VO2 only estimated", "no appropriate test protocol (VO2 not continously measured)", 
+             "no appropriate test protocol (no exhaustion)", "no appropriate test protocol (longer than 20 minutes)", "no information regarding the exclusion criteria")
+)
+
+write.csv(fullsample, file = "../data/fullsample.csv", na = "", row.names = FALSE)
+```

### 2022-05-17 thesis/methods.qmd
Create additional data summary

@@
 
-I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022].
+I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022].
 The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a].
@@
 
-To determine the influence the most common data processing strategies have on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from ramp tests in running.
+To determine the influence of the most common data processing strategies on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from ramp tests in running.
 
@@
 Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
-The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
+The data were from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
 The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies).
@@
 
-Moving time based averages were calculated by first linearly interpolating the breath-by-breath data to seconds.
+Moving time-based averages were calculated by first linearly interpolating the breath-by-breath data to seconds.
 Subsequently a (center aligned) moving average was calculated over a defined timespan.
@@
 Binned time averages were calculated using a custom function (available in the [analysis script](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md) on GitHub).
-Breath-by-breath data was initially interpolated to full seconds and then binned into consecutive interval of a constant length.
+Breath-by-breath data were initially interpolated to full seconds and then binned into consecutive intervals of constant lengths.
 The average of each interval was aligned to its center.
@@
 Butterworth filters are a class of recursive digital filters.
-@robergs2010 and @weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data.
-However @robergs2010 missed to account for the time lag introduced by Butterworth filters.
-Therefore I applied a zero-phase Butterworth filter by means of a forward-backward filtering.
+Robergs et al.
+@robergs2010 and Weir et al.
+@weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data.
+However Robergs et al.
+@robergs2010 missed to account for the time lag introduced by Butterworth filters.
+Therefore I applied a zero-phase Butterworth filter by means of forward-backward filtering.
 While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it cannot be used online (i.e.
@@
 I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipy.signal.filtfilt()` function [@virtanen2020]).
-I used 3 as the order parameter and 0.04 as the cut-off frequency for each filter [@robergs2010].
-Note that due to the forward-backward approach both the overall order and overall cut-off frequency are different from these values.
+I used 3 as the order parameter and 0.04 as the low-pass cut-off frequency for each filter [@robergs2010].
+Note that due to the forward-backward approach both the filter's overall order and overall cut-off frequency are different from these values.
 The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.

### 2022-05-17 thesis/introduction.qmd
Adjust to new citation style

@@
 Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value.
-Therefore most exercise protocols consist of a continuous or stepwise increase with an optimal during around 8-12 minutes [@buchfuhrer1983 ; @yoon2007].
-While this narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008].
+Therefore most exercise protocols consist of a continuous or stepwise increase in load with an optimal duration of 8-12 minutes [@buchfuhrer1983 ; @yoon2007].
+While these narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008].
 Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.
 
-Early research on V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017].
+Early research on V̇O~2max~ used bags to collect expired air for later analysis --- the so called Douglas bag method [@shephard2017].
 Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001].
@@
 Different data processing strategies influence measured parameters of gas exchange.
-As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
+As soon as when the first automated systems for gas exchange measurement were available, Matthews et al.
+@matthews1987 reported that different processing strategies lead to different V̇O~2max~ values.
 Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021].
-@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
-Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
+@midgley2007 Midgley et al.
+found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
+Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 Martin-Rincon et al.
+provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
 There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
@@
 Differences in V̇O~2max~ due to data processing can cause serious implications in practice.
-They hinder the comparability of data from studies within meta-analysis [@martin-rincon2019] or the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing.
-Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion of V̇O~2max~ [@astorino2009].
-Crucially, in situation where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].
+They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analysis [@martin-rincon2019].
+Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009].
+Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].
 
 Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998].
+Howley et al.
 @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~.
-Based on synthetic and experimental data, @robergs2003 opposed these conclusion, stating that shorter calculation intervals are less erroneous.
+Based on synthetic and experimental data, Robergs & Burnett @robergs2003 opposed this conclusion, stating that shorter calculation intervals are less erroneous.
 They further argued to use breath-based moving averages instead of time averages.
 In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003].
-In a classic opinion piece, @robergs2010 argued for using digital filtering (i.e.
-a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004.
-However, @robergs2010 acknowledged the lack of accessible software implementing such procedures.
+In a classic opinion piece, Robergs et al.
+@robergs2010 argued for using digital filtering (i.e.
+a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by Weir et al.
+@weir2004.
+However, @robergs2010 Robergs et al.
+acknowledged the lack of accessible software implementing such procedures.
 
@@
 In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998].
-@midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.
+@midgley2007 Midgley et al.
+were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.
 They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively.
 One third of the studies did not describe their processing method at all.
-While providing interesting first insights into reporting and processing practices, the search by @midgley2007 was not systematic and its methods were not described in detail.
+While providing interesting first insights into reporting and processing practices, the search by Midgley et al.
+@midgley2007 was not systematic and its methods were not described in detail.
+Robergs et al.
 @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*.
-They provided the latter one with a total of 75 respondents, which reported a great variety in data processing strategies.
+They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies.
 Most researchers reported the use of binned time averages over 30 or 60 seconds.
@@
 binned time) [@astorino2009; @midgley2007].
-@martin-rincon2019 provided formula for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
-Therefore motivation and measurement devices may have interacted with the influence of processing strategies.
+@martin-rincon2019 Martin-Rincon et al.
+provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
+Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.
 

### 2022-05-17 thesis/results.qmd
Adjust to new citation style

@@
 Out of the random sample (n = 500), 242 articles were included in the final analysis.
+All sampled studies and their inclusion/exclusion status and reason are available on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/fullsample.csv).
 
@@
 This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5).
-For the automated outlier detection authors removed single data points differing from a local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval.
+For the automated outlier detection authors removed single data points differing from a not further defined local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval.
 When reported, the software used for data processing varied among studies showing a total of more than 15 reported programs (for 30 studies that reported this parameter).
 
-The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
+The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
@@
 
-Binned time averages systematically generate V̇O~2max~ values compared to moving averages.
+Binned time averages systematically generate lower V̇O~2max~ values compared to moving averages.
 Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.

### 2022-05-20 scripts/comparison.Rmd
Update comparison script

@@
 # Get data files
+files <- list.files("../data/ramptests", full.names = TRUE)
 
-files <- list.files("../data/ramptests", full.names = TRUE)
+# ZAN raw data only saves imprecise body mass data in the raw data file.
+info <- read.csv("../data/participants.csv")
 ```
@@
 ```{r calc, eval = FALSE}
-# ZAN raw data only saves imprecise body mass data in the raw data file.
-info <- read.csv("../data/participants.csv")
-
 # level 1: iterate over each data file
@@
 
+## Calculate zero phase digital filter
+
+```{r}
+# Function to get VO2max with digital filtering
+get_fzmax <- function(i) {
+  dta <- spiro(files[i], weight = info$bodymass[i])
+  spiro_max(dta, smooth = "0.04fz3")$VO2_rel
+}
+
+# iterate over all tests
+fzmax <- purrr::map_dbl(seq_along(files), get_fzmax)
+
+# normalize to reference procedure
+fzmax_norm <- fzmax / res$tb_30
+
+# get median and quantile values
+median(fzmax_norm)
+quant(fzmax_norm, 0.1, 0.9)
+```
+
 ## Plot: Example of different strategies

### 2022-05-20 thesis/results.qmd
Add sentence for comparison of filtered data

@@
   t,
-  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
+  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Rationale"),
   align = "r",
@@
 This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see @fig-rr).
+Using a zero-phase forward-backward filter (third order, 0.04 Hz cut-off frequency) lead to V̇O~2max~ values 0.4% \[-0.2%; +1.4%\] (median, 10th and 90th quantile) higher than the 30-second binned time average approach.
 

### 2022-05-20 thesis/discussion.qmd
Recreate references

@@
 Despite calls to use moving averages [@robergs2003; @robergs2010], binned time averages remain the most commonly used data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies).
-These numbers are generally in line with the findings of the non-systematic search by @midgley2007 and the survey by @robergs2010.
-It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that highly discourage researchers from using binned averages [@robergs2010].
+These numbers are generally in line with the findings of the non-systematic search by Midgley et al.
+@midgley2007 and the survey by Robergs et al.
+@robergs2010.
+It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that discourage researchers from using binned averages [@robergs2010].
 Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison).
@@
 Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
-Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain very common in scientific literature.
+Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain extremely common in the scientific literature.
 
@@
 The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010].
-However the proportion of breath-based averages is much smaller than that of binned time averages, which have not been recommended in such a way.
+However the proportion of breath-based averages is still much smaller than that of binned time averages, which have not been recommended in such a way.
 
-The length of the calculation intervals is highly diverse within the literature (see @fig-duration).
-This may reflect contradictory recommendation [@howley1995; @robergs2003].
+The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration).
+This may reflect contradictory recommendations [@howley1995; @robergs2003].
 As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration.
@@
 The exact reasons for exercise scientist to ignore most recommendations by using binned time averages remain unknown.
-Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~, but past publications on this issue have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
+Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~.
+But past publications on this issue have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
 Researchers may also use binned averages for traditional reasons.
 Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data apparently similar to those by a binned time average of breath-by-breath data.
-But comparability with older data should only be an issue when using data acquired by different measurement methods within on analysis.
+But comparability with older data should only be an issue when using data acquired by different measurement methods within one analysis.
 Current studies using new breath-by-breath data do not reasonably need to rely on outdated methods of data processing.
@@
 Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software.
-This is also an reasonable explanation why digital filtering has --- despite recommended by @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
+This may also explain why digital filtering has --- despite recommended by Robergs et al.
+@robergs2010 --- has not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
 The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of easy-to-use software solutions.
@@
 In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison).
-The analysed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research.
-Some studies reported even higher mean differences up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison.
-The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies.
+The analysed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019].
+Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison.
+The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such is not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies.
 While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in well-trained athletes.
@@
 Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison).
-This is hardly surprising, as the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths.
-For an athletic population, V̇O~2max~ values obtained by moving time and moving breath average can approximately used interchangeable.
-Given that less trained individual will likely display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.
+This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths.
+For an athletic population, V̇O~2max~ values obtained by moving time and moving breath average can approximately used interchangeably.
+Given that less trained individual display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.
 
@@
 On an individual level, data processing strategies may impact the V̇O~2max~ in different severity.
-For example, for 10% of the investigated athletes a binned time average of 5 seconds lead to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
-Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classification.
-However on an individual level such work does not help.
+For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
+Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications.
+However on an individual level these can only be applied with a high margin of error.
 Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others.
@@
 This thesis focussed on the occurrence and impact of different data processing strategies, but did not investigate their validity in the first place.
-However the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for data reporting.
+However, the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for data reporting.
 
@@
 Moving time or moving breath averages are preferable to binned averages.
-In a athletic population they may be used equivalent, but in sedentary individuals breath-based averages will likely lead to lower V̇O~2max~ values than time-based averages with the same interval length parameter.
+In an athletic population they may be used equivalent, but in sedentary individuals breath-based averages will lead to lower V̇O~2max~ values than time-based averages with the same interval length parameter.
 While using a constant breath interval will lead to a similar degree of data smoothing regardless of the training status, using a constant time interval will lead to the same physiological time-frame for determining V̇O~2max~ (but with a different degree of data smoothing).
@@
 Digital filtering seems to be a promising method, as it markedly reduces the variability moving averages have due to single data outliers (see @fig-example).
-I therefore agree with @robergs2010, who recommends digital filters over any classical averaging procedure.
+I therefore agree with Robergs et al.
+@robergs2010, who recommend digital filters over any classical averaging procedure.
 The exact type of filter and values of filter parameters to be used have not yet been systematically investigated, and it is unclear whether it is possible at all to determine criteria for identifying an optimal filter.
-However, a Butterworth filter, as proposed by @robergs2010, seems to produce reasonably smoothed data.
-Note, that a single Butterworth filter comprises a time lag, a fact highlighted by @weir2004 that was not acknowledged by @robergs2010.
+However, a Butterworth filter, as proposed by Robergs et al.
+@robergs2010, seems to produce reasonably smoothed data.
+Note, that a single Butterworth filter comprises a time lag, a fact highlighted by Weir et al.
+@weir2004 that was not acknowledged by Robergs et al.
+@robergs2010.
 To allow for a correct phasing of data, as well as for a correct V̇O~2max~ determination when the measurement is terminated shortly after reaching exhaustion, a zero-phase filter is needed.
-A zero-phase forward-backward Butterworth filter (as suggested by @weir2004) seems to produce reasonable smoothing when used with the parameters suggested by @robergs2010 (low-pass cut-off frequency: 0.04 Hz; third order; see @fig-example).
-It should be noted, that despite using the same parameters the degree of filtering varies from that by @robergs2010, as the filter is applied twice, changing the overall magnitude of parameters.
-Other filtering parameters may be as reasonable as this, but it is unclear how to objectively compare them.
+A zero-phase forward-backward Butterworth filter (as suggested by Weir et al.
+@weir2004) seems to produce reasonable smoothing when used with the parameters suggested by Robergs et al.
+@robergs2010 (low-pass cut-off frequency: 0.04 Hz; third order; see @fig-example).
+It should be noted, that despite using the same parameters the degree of filtering varies from that by Robergs et al.
+@robergs2010, as the filter is applied twice, which changes the overall magnitude of parameters.
+Other filtering parameters may be as reasonable as this, but it is unclear how to objectively investigate this issue.
 
@@
 The V̇O~2max~ is then defined as the highest single filtered data point.
-An filter order of 3 and a low cut-off frequency of 0.04 Hz appear to be reasonable parameters for each filter.
+A filter order of 3 and a low cut-off frequency of 0.04 Hz appear to be reasonable parameters for each filter.
 While such data processing has advantages over traditional data processing forms, it requires specialized software.
@@
 @tbl-recommendation lists 7 item that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~.
-These items may be reported in form of a check-list, as an in-text enumeration or in sentence form.
-An example paragraph containing all the relevant information for the current study would be:
+These items may be reported in form of a checklist, as an in-text enumeration or in sentence form.
+An example paragraph containing all the relevant information for the original data presented in this thesis [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:
 
@@
 The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro Package for R [@nolte2022].
-This strategy produces similar results as that recommended by @robergs2010, but does not produce a time lag."*
+This strategy produces similar results as that recommended by Robergs et al.
+@robergs2010, but does not produce a time lag."*
 
@@
 But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
-Sharing of the raw metabolic exchange data is the solution to this problem.
+Sharing of the raw metabolic gas exchange data is the solution to this problem.
 It allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
-Anonymization of raw gas exchange data files should not be a big deal for researchers, since most of these files are structured in a simple way and allow to easily remove any personal information (if this has not been done in the metabolic cart's system before).
+Anonymization of raw gas exchange data files should not be a hinder sharing, since most of these files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before).
 In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code.
@@
 For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly.
-Moreover the exact definitions for building binned averages varied within the literature: While most studies defined the binning periods from the beginning of the exercise, some may have defined them from the endpoint.
-Additionally some studies reviewed did not define the maximum, but a predefined binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
-In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average.
+Moreover the exact definitions for building binned averages varies within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint.
+Additionally some studies reviewed did not define the maximum, but a preset binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
+In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing.
 I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.
+
+This work treated each breath as the single data processing unit of cardiopulmonary exercise testing.
+However, metabolic carts sample gas fraction and gas flow data at a much higher frequency (e.g., 50 Hz).
+The data for each breath is subsequently calculated from these raw signals by the means of an algorithm.
+Different algorithms to generate the breath-by-breath data can lead to different outcomes [@koschate2019], and as such may also influence the V̇O~2max~.
+Hence documenting and reporting of the breath-by-breath algorithm is advisable.
+Yet many metabolic carts do not describe their default algorithm and limit access to the raw data signal.
 
@@
 I did not conduct a formal analysis of the validity or reliability of different data processing strategies, so recommendations regarding their use rely on theoretical derivations and prior research.
-Since the reliability of different strategies appears to be similar [@midgley2007; @martin-rincon2019] and currently no accepted methods to quantify validity of V̇O~2max~ exist, the presented approach is sufficient to derive recommendations for data processing strategies to determine V̇O~2max~.
+Since the reliability of different strategies appears to be similar [@midgley2007; @martin-rincon2019] and currently no accepted methods to quantify the validity of V̇O~2max~ exist, the presented approach is sufficient to derive recommendations for data processing strategies to determine V̇O~2max~.

### 2022-05-20 thesis/introduction.qmd
Recreate references

@@
 Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021].
-@midgley2007 Midgley et al.
-found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
-Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 Martin-Rincon et al.
-provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
+Midgley et al.
+@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
+Based on own data in sedentary and moderately trained individuals, Martin-Rincon et al.
+@martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
 There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
@@
 For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied.
-For more details on the used data processing strategies, see the methods section.
-Data Source: [@quittmann_unpubl_b]"
+For more details on the used data processing strategies, see the @sec-processing.
+Data from [@quittmann_unpubl_b]"
 #| fig-scap: "Example of raw breath-by-breath data processed by different strategies during the final minutes of a ramp test to exhaustion."
@@
 Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009].
-Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].
+Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].
 
@@
 @weir2004.
-However, @robergs2010 Robergs et al.
-acknowledged the lack of accessible software implementing such procedures.
+However, Robergs et al.
+@robergs2010 acknowledged the lack of accessible software implementing such procedures.
 
@@
 In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998].
-@midgley2007 Midgley et al.
-were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.
+Midgley et al.
+@midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals.
 They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively.

### 2022-05-20 thesis/methods.qmd
Recreate references

@@
 
-The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020].
+The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker].
 Any deviations from the preregistration are indicated in the 'Transparent Changes' document (@sec-tchanges).
@@
 I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022].
-The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a].
+The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022].
 The attached packages and default settings are documented in @sec-tech.
@@
 Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
-The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018a].
+The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018].
 @sec-checklist contains the corresponding reporting checklist.
@@
 
-### Data Processing
+### Data Processing {#sec-processing}
 

### 2022-05-21 thesis/conclusion.qmd
Write conclusion section

@@
 # Conclusion
+
+The determination of V̇O~2max~ from raw breath-by-breath data requires some form of data post-processing.
+As shown by the scoping review, researchers use a variety of strategies for this data processing.
+I demonstrated on experimental data, that different processing strategies systematically bias the V̇O~2max~ assessment, which can ultimately lead to misclassification of individuals or groups and incorrect evaluation of interventions.
+Despite contradictory recommendations, binned time average remain by far the most common data processing strategy.
+The calculation intervals for time averages range from 5 to 60 seconds in the reviewed literature, which heavily influences the V̇O~2max~ value obtained.
+Almost all research did only incompletely report on their data processing to determine V̇O~2max~, and as such can be regarded non-reproducible.
+To ensure valid and reproducible V̇O~2max~ assessments, researchers should at minimum report their processing strategy based on a set of guidelines I developed in this thesis.
+Despite calls for standardization of data processing, to date no evidence for an optimal strategy exists, but digital filtering (e.g., zero-phase Butterworth filter) appears to generate a more reasonable data smoothing compared to traditional (moving) averages.
+I encourage researchers to share their raw gas exchange data and use programming software for data analysis, to minimize the bias due to data processing when determining the V̇O~2max~.

### 2022-05-21 thesis/conclusion.qmd
Improve conclusion

@@
 As shown by the scoping review, researchers use a variety of strategies for this data processing.
-I demonstrated on experimental data, that different processing strategies systematically bias the V̇O~2max~ assessment, which can ultimately lead to misclassification of individuals or groups and incorrect evaluation of interventions.
+I demonstrated on experimental data, that different processing strategies systematically bias the V̇O~2max~ assessment, which can ultimately lead to misclassification of individuals or groups, or to incorrect evaluation of interventions.
 Despite contradictory recommendations, binned time average remain by far the most common data processing strategy.

### 2022-05-21 thesis/abstract.qmd
Add German abstract to thesis

---
title: ''
format: pdf
bibliography: ../references.bib
---

# Zusammenfassung (German abstract) {.unnumbered}

Die maximale Sauerstoffaufnahme (V̇O~2max~) ist die wichtigste physiologische Kenngröße der Ausdauerleistungsfähigkeit. Zur Bestimmung der V̇O~2max~ werden während eines Rampentests die Atemgase kontinuierlich gemessen. Die üblicherweise im breath-by-breath-Verfahren erhobenen Daten erfordern eine Weiterverarbeitung um die V̇O~2max~ zu bestimmen. Unterschiedliche Verfahren der Datenverarbeitung können die gemessenen V̇O~2max~ beeinflussen. Ziel der vorliegenden Arbeit ist es, anhand eines scoping review zu bestimmen, welche Datenverarbeitungsstrategien in der wissenschaftlichen Literatur genutzt werden, um diese im Anschluss anhand experimenteller Daten zu vergleichen. Für den Review wurde eine Zufallsstichprobe von 500 Artikeln untersucht; der experimentelle Vergleich erfolgte anhand von 72 standardisierten Rampentests auf dem Laufband. Die im Review untersuchten Studien benutzen eine Vielzahl von verschiedenen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben werden. Die unterschiedlichen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, mit deutlich höheren individuellen Abweichungen. Die Ergebnisse verdeutlichen den Bedarf einer Standardisierung und korrekter Darstellung der benutzten Datenverarbeitungsstrategien zur Bestimmung der V̇O~2max~. Diese Arbeit bietet hierfür Empfehlungen zur Verwendung und Vergleichbarkeit verschiedener Strategien, sowie Richtlinien zur Beschreibung dieser im Rahmen wissenschaftlicher Studien.

### 2022-05-23 scripts/comparison.Rmd
Change legend in example plot

@@
 p_ex <- ggplot(exm_tidy, aes(x = time)) +
-  geom_point(data = exm_tidy[exm_tidy$method == "raw", ], aes(y = value), alpha = 0.1) +
-  geom_line(data = exm_tidy[exm_tidy$method %in% c("Moving Breath", "Moving Time", "Digital Filter"), ], aes(y = value, color = method), size = 0.7) +
-  geom_point(data = exm_tidy[exm_tidy$method == "Binned Time", ], aes(y = value, fill = method, colour = "Binned Time"), size = 1.4) +
+  geom_point(
+    data = exm_tidy[exm_tidy$method == "raw", ], 
+    aes(y = value, alpha = method, colour = "Raw Data")
+  ) +
+  geom_line(
+    data = exm_tidy[exm_tidy$method %in% c("Moving Breath", "Moving Time", "Digital Filter"), ], 
+    aes(y = value, color = method), 
+    size = 0.7
+  ) +
+  geom_point(
+    data = exm_tidy[exm_tidy$method == "Binned Time", ], 
+    aes(y = value, fill = method, colour = "Binned Time"), 
+    size = 1.4
+  ) +
   scale_x_continuous(
@@
     name = "Processing Strategy",
-    values = c(met.brewer("Java", 3)[3], "blue", met.brewer("Java", 3)[c(1,2)]),
-    labels = c("Binned Time", "Digital Filter","Moving Time","Moving Breath"),
+    values = c(met.brewer("Java", 3)[3], "blue", met.brewer("Java", 3)[c(1,2)], "black"),
+    labels = c("Raw Data", "Binned Time", "Digital Filter","Moving Time","Moving Breath"),
     guide = guide_legend(
@@
       override.aes = list(
-        linetype = c(rep.int("solid", 3), "blank"),
-        shape = c(rep.int(NA, 3), 19),
-        color = c(met.brewer("Java", 3)[c(1,2)], "blue", met.brewer("Java", 3)[3])
+        linetype = c(rep.int("solid", 3), "blank", "blank"),
+        shape = c(rep.int(NA, 3), 19, 19),
+        color = c(met.brewer("Java", 3)[c(1,2)], "blue", met.brewer("Java", 3)[3], "black"),
+        alpha = c(rep.int(1, 4), 0.1)
       )
@@
     guide = "none"
+  ) +  scale_alpha_manual(
+    values = 0.1,
+    guide = "none"
   ) +
@@
   theme(
-    legend.position = c(0.88, 0.6),
+    legend.position = c(0.88, 0.63),
   )

### 2022-05-23 thesis/abstract.qmd
Improve minor issues in thesis

@@
 Zur Bestimmung der V̇O~2max~ werden während eines Rampentests die Atemgase kontinuierlich gemessen.
-Die üblicherweise im breath-by-breath-Verfahren erhobenen Daten erfordern eine Weiterverarbeitung um die V̇O~2max~ zu bestimmen.
-Unterschiedliche Verfahren der Datenverarbeitung können die gemessenen V̇O~2max~ beeinflussen.
+Die üblicherweise im breath-by-breath-Verfahren erhobenen Werte erfordern eine Weiterverarbeitung um die V̇O~2max~ zu bestimmen.
+Unterschiedliche Verfahren der Datenverarbeitung können die gemessene V̇O~2max~ beeinflussen.
 Ziel der vorliegenden Arbeit ist es, anhand eines scoping review zu bestimmen, welche Datenverarbeitungsstrategien in der wissenschaftlichen Literatur genutzt werden, um diese im Anschluss anhand experimenteller Daten zu vergleichen.
 Für den Review wurde eine Zufallsstichprobe von 500 Artikeln untersucht; der experimentelle Vergleich erfolgte anhand von 72 standardisierten Rampentests auf dem Laufband.
-Die im Review untersuchten Studien benutzen eine Vielzahl von verschiedenen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben werden.
-Die unterschiedlichen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, mit deutlich höheren individuellen Abweichungen.
-Die Ergebnisse verdeutlichen den Bedarf einer Standardisierung und korrekter Darstellung der benutzten Datenverarbeitungsstrategien zur Bestimmung der V̇O~2max~.
+Die im Review untersuchten Studien benutzen eine Vielzahl von unterschiedlichen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben wurden.
+Die verschiedenen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, bei teils deutlich größeren individuellen Abweichungen.
+Die Ergebnisse verdeutlichen den Bedarf einer Standardisierung und korrekten Darstellung der benutzten Datenverarbeitungsstrategien zur Bestimmung der V̇O~2max~.
 Diese Arbeit bietet hierfür Empfehlungen zur Verwendung und Vergleichbarkeit verschiedener Strategien, sowie Richtlinien zur Beschreibung dieser im Rahmen wissenschaftlicher Studien.

### 2022-05-23 thesis/conclusion.qmd
Improve minor issues in thesis

@@
 I demonstrated on experimental data, that different processing strategies systematically bias the V̇O~2max~ assessment, which can ultimately lead to misclassification of individuals or groups, or to incorrect evaluation of interventions.
-Despite contradictory recommendations, binned time average remain by far the most common data processing strategy.
+Despite contradictory recommendations, binned time averages remain by far the most common data processing strategy.
 The calculation intervals for time averages range from 5 to 60 seconds in the reviewed literature, which heavily influences the V̇O~2max~ value obtained.
-Almost all research did only incompletely report on their data processing to determine V̇O~2max~, and as such can be regarded non-reproducible.
+Almost all research did only incompletely report on their data processing to determine V̇O~2max~, and as such can be regarded as non-reproducible.
 To ensure valid and reproducible V̇O~2max~ assessments, researchers should at minimum report their processing strategy based on a set of guidelines I developed in this thesis.

### 2022-05-23 thesis/discussion.qmd
Improve minor issues in thesis

@@
 
-The collected data shows, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained.
+This thesis aimed to map current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data.
+The results show, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained.
 Many articles only incompletely report on their methods and use outdated strategies.
@@
 Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison).
-The peak in oxygen uptake may fall in between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
+The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
 Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
@@
 
-Breath-based averages seem to be more common (≈ 8%) than reported previously (\< 1%) [@midgley2007], but less common than assessed in self-reporting (≈ 17%) [@robergs2010].
+Breath-based averages seem to be more common (\~8%) than reported previously (\<1%) [@midgley2007], but less common than assessed in self-reporting (\~17%) [@robergs2010].
 The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010].
@@
 Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~.
-But past publications on this issue have been widely cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
+But past publications on this issue have been frequently cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field.
 Researchers may also use binned averages for traditional reasons.
@@
 This may also explain why digital filtering has --- despite recommended by Robergs et al.
-@robergs2010 --- has not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
-The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of easy-to-use software solutions.
+@robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
+The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.
 
@@
 
-The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such can influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment.
+The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment.
 In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison).
@@
 The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such is not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies.
-While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in well-trained athletes.
+While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
 
@@
 While this general trend has been acknowledged previously [@martin-rincon2019], it has not been quantified.
-The data presented in this thesis suggest a ≈-1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
+The data presented in this thesis suggest a \~1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
 This difference is well within the measurement error of most if not all metabolic carts, but it is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g.
@@
 @robergs2010, who recommend digital filters over any classical averaging procedure.
-The exact type of filter and values of filter parameters to be used have not yet been systematically investigated, and it is unclear whether it is possible at all to determine criteria for identifying an optimal filter.
+The exact type of filter and values of filter parameters to be used have not yet been systematically investigated, and it is unclear whether it is possible to determine criteria for identifying an optimal filter.
 However, a Butterworth filter, as proposed by Robergs et al.
@@
 @robergs2010, but does not produce a time lag."*
+
+Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing.
+Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test.
+In cases where journals endorse word limits on articles, these reporting --- including the reporting on data processing strategies --- may be included in supplementary files.
+The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.
+
+The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
+But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
+Sharing of the raw metabolic gas exchange data is the solution to this problem.
+It allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
+Anonymization of raw gas exchange data files should not be a hinder for sharing, since most of these files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before).
+In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code.
+This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment.
+Luckily such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].
 
@@
 
-Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing.
-Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test.
-In cases where journals endorse word limits on articles, these reporting --- including the reporting on data processing strategies --- may be included in supplementary files.
-The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.
-
-The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
-But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
-Sharing of the raw metabolic gas exchange data is the solution to this problem.
-It allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
-Anonymization of raw gas exchange data files should not be a hinder sharing, since most of these files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before).
-In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code.
-This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment.
-Luckily such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].
-
 ## Limitations

### 2022-05-23 thesis/introduction.qmd
Improve minor issues in thesis

@@
 V̇O~2max~ is defined as the highest rate a body can consume oxygen [@bassett2000].
-Determined by measuring gas exchange data, it is one of the most common assessed exercise parameters in sports science and medicine.
+Determined by measuring gas exchange data, it is one of the most commonly assessed exercise parameters in sports science and medicine.
 V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological parameter for predicting endurance performance [@bassett2000].
@@
 While these narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008].
-Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.
+Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] affect the V̇O~2max~ determination.
 
@@
 Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning.
-Practices in reporting and processing may have changed more than a decade after the publication of such numbers.
+Practices in reporting and processing may have changed more than a decade after the publication of these numbers.
 
@@
 binned time) [@astorino2009; @midgley2007].
-@martin-rincon2019 Martin-Rincon et al.
-provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
+Martin-Rincon et al.
+@martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
 Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.

### 2022-05-23 thesis/methods.qmd
Improve minor issues in thesis

@@
 
+The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI).
+Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step.
+After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
+All titles that contained one of the following words were excluded: *'review', 'correction', 'meta-analysis', 'comment', 'retraction', 'editorial', 'erratum', 'reply'.*
+
 ```{r}
@@
 
-The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI).
-Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step.
-After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles.
-All titles that contained one of the following words were excluded: review, correction, meta-analysis, comment, retraction, editorial, erratum, reply.
-
 In accordance with the preregistration I drew a random sample from the search results.
@@
 When the screeners disagreed in their assessments, they resolved the conflict by discussion.
+
+After the abstract screening I retrieved the full-texts for all remaining articles.
+The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria).
+Conflicts were resolved by discussion between examiners.
+All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).
+
+### Data Extraction
+
+I retrieved data from all articles remaining after the abstract and full-text screening. Extraction included the following data:
+
+-   metabolic cart used
+-   measurement type (breath-by-breath, mixing chamber, ...)
+-   type of outcome for V̇O~2max~ (primary, secondary, other)
+-   data preprocessing (e.g., filtering)
+-   data processing software
+-   interpolation procedure
+-   data processing/determination of V̇O~2max~:
+    -   type (time average, breath average, digital filtering)
+    -   alignment (rolling, binned, ...)
+    -   interval (in seconds or breaths, parameters for filtering)
+-   reference for the used data processing strategy
+
+The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.
+All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).
 
@@
 
-After the abstract screening I retrieved the full-texts for all remaining articles.
-The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria).
-Conflicts were resolved by discussion between examiners.
-All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).
-
-### Data Extraction
-
-I retrieved data from all articles remaining after the abstract and full-text screening. Extraction included the following data:
-
--   metabolic cart used
--   measurement type (breath-by-breath, mixing chamber, ...)
--   type of outcome for V̇O~2max~ (primary, secondary, other)
--   data preprocessing (e.g., filtering)
--   data processing software
--   interpolation procedure
--   data processing/determination of V̇O~2max~:
-    -   type (time average, breath average, digital filtering)
-    -   alignment (rolling, binned, ...)
-    -   interval (in seconds or breaths, parameters for filtering)
--   reference for the used data processing strategy
-
-The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.
-All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).
-
 ### Data Synthesis
@@
 The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies).
-The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using an identical exercise protocol.
-Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for ten minutes at a velocity of 2.8 m·s^-1^ as a warm-up.
+The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b], respectively, while using an identical exercise protocol.
+Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up.
 After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds.
@@
 
-To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure.
+To compare different processing methods within and between individuals, I chose to express the V̇O~2max~ normalized to a reference procedure.
 The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review.

### 2022-05-23 thesis/results.qmd
Improve minor issues in thesis

@@
 Almost half (44.2%) of the articles did not report any information regarding their data processing strategy.
-One in twenty articles (5.8%) provided a rationale for their used strategy.
+About one in twenty articles (5.8%) provided a rationale for their used strategy.
 Only a single article [@maturana2021] reported information regarding all the investigated criteria.
@@
 
-The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
+The average V̇O~2max~ as determined by a binned 30-second average was 62.2 ± 6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
 Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see @fig-comparison).
@@
 Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.
-Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
+Notably on an individual level these increases may be lower (\~2%) or much higher (\>10%) than the median value.
 

### 2022-05-31 transparent_changes.Rmd
Update transparent changes

@@
 
-This document includes all deviation and modifications of code and methods in the final project compared to its [preregistration](https://osf.io/3am4s).
+This document includes all deviations and modifications of code and methods in the final project compared to its [preregistration](https://osf.io/3am4s).
 
@@
 Due to a miscalculation, the preregistration provided an incorrect number of exercise test (n = 76).
-The correct number of exercise tests is n = 72, with 44 from @quittmann_unpubl_a (one test only partly included in the original work due to missing other data) and 28 from @quittmann_unpubl_b (three test only partly included in the original work due to missing other data).
+The correct number of exercise tests is n = 72, with 44 from @quittmann_unpubl_a (one test only partly included in the original work due to missing other data) and 28 from @quittmann_unpubl_b (three tests only partly included in the original work due to missing other data).
 
@@
 
-`outcome`: Which type of outcome VO2max is. Either `primary`, `secondary` or `other`
+`outcome`: Which type of outcome V̇O~2max~ is. Either `primary`, `secondary` or `other`
 
@@
 Temporary unblinding only applied to the primary researcher.
-This concerns the abstract with the Sampling ID (sid): 50, 238, 288, 416, 488, 490, 500
+This concerns the abstracts with the sampling id (sid): 50, 238, 288, 416, 488, 490, 500
 -   Manual retrieval of abstracts for articles, as the automatically collected abstract contained html-tags that could not be removed for later abstract plots.
@@
 Temporary unblinding only applied to the primary researcher.
-This concerns the abstract with the Sampling ID (sid): 344, 356
--   Unblinding during screening to assess the implications of title given in squared brackets. This concerns the abstract with the Sampling ID (sid): 262, 303
--   Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the Sampling ID (sid): 275
+This concerns the abstracts with the sampling id (sid): 344, 356
+-   Unblinding during screening to assess the implications of title given in squared brackets. This concerns the abstracts with the sampling id (sid): 262, 303
+-   Consulting of online abstract due to incomplete abstract plot. This concerns the abstract with the sampling id (sid): 275
 

### 2022-05-31 thesis/abstract.qmd
Finish thesis work

@@
 Für den Review wurde eine Zufallsstichprobe von 500 Artikeln untersucht; der experimentelle Vergleich erfolgte anhand von 72 standardisierten Rampentests auf dem Laufband.
-Die im Review untersuchten Studien benutzen eine Vielzahl von unterschiedlichen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben wurden.
+Die im Review untersuchten Studien benutzen eine Vielzahl von unterschiedlichen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben werden.
 Die verschiedenen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, bei teils deutlich größeren individuellen Abweichungen.

### 2022-05-31 thesis/discussion.qmd
Finish thesis work

@@
 
-This thesis aimed to map current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data.
-The results show, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained.
+This thesis aimed to review current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data.
+The results show, that current research uses a variety of processing strategies to determine V̇O~2max~, which directly influences the values obtained.
+Identical raw breath-by-breath data can result in different V̇O~2max~ values when processed differently.
 Many articles only incompletely report on their methods and use outdated strategies.
@@
 
-## Current state of data processing
+## Current State of Data Processing
 
 Despite calls to use moving averages [@robergs2003; @robergs2010], binned time averages remain the most commonly used data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies).
-These numbers are generally in line with the findings of the non-systematic search by Midgley et al.
+The found numbers are generally in line with the findings of the non-systematic search by Midgley et al.
 @midgley2007 and the survey by Robergs et al.
@@
 The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example).
-Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
+Binned time averages revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
 Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain extremely common in the scientific literature.
@@
 As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration.
-As different interval durations can heavily influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
+As different interval durations influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
 
-The exact reasons for exercise scientist to ignore most recommendations by using binned time averages remain unknown.
+The exact reasons for exercise scientists to ignore most recommendations by using binned time averages remain unknown.
 Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~.
@@
 But comparability with older data should only be an issue when using data acquired by different measurement methods within one analysis.
-Current studies using new breath-by-breath data do not reasonably need to rely on outdated methods of data processing.
+Current studies using newly aquired breath-by-breath data do not reasonably need to rely on outdated methods of data processing.
 
 A major source for choosing suboptimal processing strategies may be limitations by analysis software.
-My review shows that most researchers use the software of the metabolic cart's manufacturer to analyse the gas exchange data.
+My review shows that most researchers use the software of the metabolic cart's manufacturer to analyze the gas exchange data.
 These software may by default output binned time averages instead of raw breath-by-breath data.
@@
 
-## Impact of different data processing
+## Impact of Different Data Processing Strategies
 
@@
 In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison).
-The analysed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019].
+The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019].
 Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison.
 The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such is not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies.
-While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
+While previous research was often conducted in sedentary or recreationally trained individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
 
@@
 This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths.
-For an athletic population, V̇O~2max~ values obtained by moving time and moving breath average can approximately used interchangeably.
-Given that less trained individual display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.
+For an athletic population, V̇O~2max~ values obtained by moving time and moving breath averages can approximately be used interchangeably.
+Given that less trained individuals display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.
 
@@
 Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications.
-However on an individual level these can only be applied with a high margin of error.
+However, on an individual level these equations can only be applied with a high margin of error.
 Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others.
@@
 
-## Guidelines for data processing and reporting
+## Guidelines for Data Processing and Reporting
 
 This thesis focussed on the occurrence and impact of different data processing strategies, but did not investigate their validity in the first place.
-However, the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for data reporting.
+However, the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for reporting.
 
@@
 Moving time or moving breath averages are preferable to binned averages.
-In an athletic population they may be used equivalent, but in sedentary individuals breath-based averages will lead to lower V̇O~2max~ values than time-based averages with the same interval length parameter.
+In an athletic population they may be used equivalent, but in sedentary individuals breath-based averages will lead to lower V̇O~2max~ values than time-based averages with the same interval length.
 While using a constant breath interval will lead to a similar degree of data smoothing regardless of the training status, using a constant time interval will lead to the same physiological time-frame for determining V̇O~2max~ (but with a different degree of data smoothing).
@@
 
-To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategy is crucial.
+To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial.
 The review presented here shows that almost half of the studies that measured V̇O~2max~ did not describe their processing strategy.
 Other aspects of the data processing, such as outlier filtering or rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting).
-@tbl-recommendation lists 7 item that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~.
+@tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~.
 These items may be reported in form of a checklist, as an in-text enumeration or in sentence form.
@@
 > *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America).
-The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro Package for R [@nolte2022].
+The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro package for R [@nolte2022].
 This strategy produces similar results as that recommended by Robergs et al.
@@
 #| echo: false
-#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake"
+#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake."
 #| label: tbl-recommendation
@@
   desc = c(
-    "State the exact device model and manufacturer",
-    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…)",
-    "State the name and version of the software used for data analysis",
-    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis",
-    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages)",
-    "State the parameters used for the processing strategy (e.g., length of averaging interval)",
-    "State the rationale for using the processing strategy (e.g., reference to recommendations)")
+    "State the exact device model and manufacturer.",
+    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…).",
+    "State the name and version of the software used for data analysis.",
+    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis.",
+    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages).",
+    "State the parameters used for the processing strategy (e.g., length of averaging interval).",
+    "State the rationale for using the processing strategy (e.g., reference to recommendations).")
 )
@@
 For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly.
-Moreover the exact definitions for building binned averages varies within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint.
-Additionally some studies reviewed did not define the maximum, but a preset binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
+Moreover the exact definitions for building binned averages vary within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint.
+Additionally some studies reviewed did not define the maximum, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
 In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing.
-I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.
+I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguities in their reporting.
 

### 2022-05-31 thesis/introduction.qmd
Finish thesis work

@@
 
-Performance in endurance sports is limited by the human physiology.
+Performance in endurance sports is mainly limited by the human physiology.
 Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation.
@@
 Differences in V̇O~2max~ due to data processing can cause serious implications in practice.
-They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analysis [@martin-rincon2019].
+They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analyses [@martin-rincon2019].
 Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009].
@@
 
-Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998].
+Calls to standardize processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998].
 Howley et al.
@@
 In absence of one commonly accepted method, data processing varies among the literature.
-In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998].
+In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research [@myers1990; @astorino2009; @scheadler2017; @johnson1998].
 Midgley et al.
@@
 Martin-Rincon et al.
-@martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts.
+@martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts.
 Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.
@@
 Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies.
-No research has yet compared a variety of systematically derived strategies among a group of well-trained individuals using a standardized measurement set-up.
+No research has yet compared a variety of systematically derived strategies among a group of trained individuals using a standardized measurement set-up.
 
 This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature.
-Secondly, I will compare different data processing strategies in relation to the most commonly applied on a set of 72 standardized treadmill tests in well-trained individuals.
+Secondly, I will compare the most common data processing strategies on a set of 72 standardized treadmill tests in well-trained individuals.
 The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals.
 The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting.
-The results build a basis for providing new recommendations for data processing and its reporting for determining the V̇O~2max.~
+The results build a basis for providing new recommendations for data processing to determine the V̇O~2max~ and its reporting in research articles.

### 2022-05-31 thesis/methods.qmd
Finish thesis work

@@
 format: pdf
-crossref:
-  sec-prefix: appendix
 bibliography: ../references.bib
@@
 The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker].
-Any deviations from the preregistration are indicated in the 'Transparent Changes' document (@sec-tchanges).
+Any deviations from the preregistration are indicated in the 'Transparent Changes' document ([Appendix @sec-tchanges]).
 Major deviations will also be explicitly stated within the methods section.
@@
 The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022].
-The attached packages and default settings are documented in @sec-tech.
+The attached packages and default settings are documented in [Appendix @sec-tech].
 
@@
 The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018].
-@sec-checklist contains the corresponding reporting checklist.
+[Appendix @sec-checklist] contains the corresponding reporting checklist.
 
@@
 
-The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science]().
+The article search was conducted on 16th March 2022 using [PubMed](https://pubmed.ncbi.nlm.nih.gov/) and [Web of Science](https://www.webofscience.com/wos/woscc/basic-search).
 The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords.
@@
 The abstracts from the articles included in the random sample were blinded for scanning.
-This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see @sec-blindabstract for an example).
-Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion.
-When the screeners disagreed in their assessments, they resolved the conflict by discussion.
+This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see [Appendix @sec-blindabstract] for an example).
+Two researchers independently scanned the blinded abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion.
+When the screeners disagreed in their assessment, they resolved the conflict by discussion.
 
@@
 -   metabolic cart used
--   measurement type (breath-by-breath, mixing chamber, ...)
+-   measurement type (breath-by-breath, mixing chamber)
 -   type of outcome for V̇O~2max~ (primary, secondary, other)
@@
 -   data processing/determination of V̇O~2max~:
-    -   type (time average, breath average, digital filtering)
-    -   alignment (rolling, binned, ...)
+    -   type (time average, breath average, digital filtering, ...)
+    -   alignment (moving, binned, ...)
     -   interval (in seconds or breaths, parameters for filtering)
--   reference for the used data processing strategy
+-   rationale for the used data processing strategy (e.g. a reference)
 
-The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.
+The criteria 'type of outcome' and 'rationale' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.
 All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).
@@
     info = c(
-      "Full text only available in non-English language",
+      "full text only available in non-English language",
       "research was no original investigation or only a reanalysis of data", 
       "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", 
-      "Protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
-      "no information regarding the exclusion criteria | crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
+      "protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
+      "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
       )
@@
 I calculated the relative and absolute frequency for the reporting of the extracted items.
-Similarly, I counted the use of different data strategies for processing data in all articles that reported measuring breath-by-breath.
+Similarly, I counted the use of different strategies for processing data in all articles that reportedly measured breath-by-breath.
 Total interval duration of averaging procedures were derived from the reported parameters.
@@
 
-A total of N = 72 exercise tests were analysed for this study.
+A total of N = 72 exercise tests were analyzed for this study.
 Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
@@
 
-The spiro Package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data.
+The spiro package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data.
 It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data.
@@
 Breath based moving averages were calculated on the raw data.
-As this functionality is implemented in the spiro package, I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.
+As this functionality is implemented in the spiro package [@nolte2022], I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.
 
@@
 @robergs2010 and Weir et al.
-@weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data.
+@weir2004 argued that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data.
 However Robergs et al.
@@
 
-### Comparison of methods
+### Comparison of Methods
 
 To compare different processing methods within and between individuals, I chose to express the V̇O~2max~ normalized to a reference procedure.
-The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review.
+The reference procedure was chosen as being the most commonly applied in current literature as determined by the scoping review.
 Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method.

### 2022-05-31 thesis/results.qmd
Finish thesis work

@@
 #| fig-cap: "Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]"
-#| fig-scap: "Flow diagram for the systematic scoping review"
+#| fig-scap: "Flow diagram for the systematic scoping review."
 #| label: fig-flow
@@
 #| echo: false
-#| fig-cap: "Data strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
+#| fig-cap: "Strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
 #| label: fig-strategies
@@
 The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
-30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
+30-second intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
 
@@
 Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review.
-Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile.
+Solid lines display the median, the shaded areas mark the interval between 10th and 90th percentile.
 Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
 Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
-#| fig-scap: "Comparison of different data processing strategies"
+#| fig-scap: "Comparison of different data processing strategies."
 #| label: fig-comparison
@@
 This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see @fig-rr).
-Using a zero-phase forward-backward filter (third order, 0.04 Hz cut-off frequency) lead to V̇O~2max~ values 0.4% \[-0.2%; +1.4%\] (median, 10th and 90th quantile) higher than the 30-second binned time average approach.
+Using a zero-phase forward-backward filter (third order, 0.04 Hz cut-off frequency) leads to V̇O~2max~ values 0.4% \[-0.2%; +1.4%\] (median, 10%- and 90%-quantile) higher than the 30-second binned time average approach.
 
@@
 #| fig-cap: "Respiratory rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72)."
-#| fig-scap: "Respiratory rates during the ramp tests"
+#| fig-scap: "Respiratory rates during the ramp tests."
 #| label: fig-rr

### 2022-05-31 thesis/thesis.qmd
Include checklist in thesis
Add German abstract to thesis
Improve minor issues in thesis
Finish thesis work

@@
       text: \usepackage{setspace}
+    title-meta: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
+    author-meta: 'Simon Nolte'
+    subject: 'Released under a CC-BC-NC-ND license'
 number-sections: true
@@
 
-Herewith I affirm in lieu of an oath that I have authored this Master thesis independently and did not use any other sources and tools than indicated.
+Herewith I affirm in lieu of an oath that I have authored this Bachelor thesis independently and did not use any other sources and tools than indicated.
 All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named.
@@
 
-# Zusammenfassung (German abstract) {-}
+\onehalfspacing
+
+```{r}
+#| child = 'abstract.qmd'
+```
 
 \newpage
+
+\singlespacing
 
@@
 
-\newpage
+\vspace{2cm}
 
@@
 
+```{r}
+#| child = '../transparent_changes.Rmd'
+```
+
 \newpage
@@
 
+![](checklist.pdf)
+
+```{r}
+#| echo: false
+#| out-extra: "page=2"
+
+knitr::include_graphics("checklist.pdf")
+```
+
 \newpage

### 2022-10-09 paper.qmd
Create first paper draft

---
title: "Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison with Guidelines for Reporting"
format: docx
bibliography: references.bib
csl: sports-med-new.csl
always_allow_html: true
---

# Abstract

**Background** Gas exchange data from maximum oxygen uptake (V̇O~2max~) testing needs post-processing. Different processing strategies may lead to varying V̇O~2max~ values in individuals. The exact processing strategies used in the literature have not yet been systematically investigated. Previous research provided comparisons of methods only on a group level.

**Methods** Out of a random sample, we investigated 242 recently published articles that measured V̇O~2max~ in an appropriate setting. We extracted the reported methods of data processing and their rationale. We compared the most common processing strategies on a data set of 72 standardized exercise tests in trained athletes.

**Results** Most researchers use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages. Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the strategy used. The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% with considerable variation on an individual level.

**Conclusion** Researchers should report their data processing strategy used to determine V̇O~2max~. We provide a reporting checklist of seven items that can function as a template for reporting.

## Key Points

-   Despite calls to use moving averages or digital filter, binned averages remain the most common data processing strategy to determine V̇O~2max~.
-   Different processing strategies lead to varying V̇O~2max~ values, which can hardly be predicted on an individual level.
-   Researchers should at minimum report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.

# Introduction

The maximum oxygen uptake (V̇O~2max~) is one of the most commonly assessed physiological parameters in sports and exercise science [@bassett2000]. The V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological predictor of endurance performance [@bassett2000]. Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. Among the factors that influence the V̇O~2max~ determination are the exercise test protocol [@buchfuhrer1983 ; @yoon2007; @midgley2008], exercise modality [@myers1991], motivation [@midgley2018], and biological variability [@katch1982]. The measuring method (e.g. mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].

Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018]. To identify a true maximum, researchers commonly evaluate a set of parameters measured during ramp tests --- the so called 'V̇O~2max~ criteria' [@howley1995]. We will not distinguish between peak and maximum oxygen uptake in this article, as the criteria for V̇O~2max~ (e.g. the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do heavily rely on the data processing strategy used [@astorino2009]. Thus, we speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.

Measured gas exchange data is noisy. Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) lead to a highly fluctuating raw oxygen uptake on a breath-by-breath basis. For interpretation the raw data requires some form of processing. Different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021]. Midgley et al. @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, inferring that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, Martin-Rincon et al. @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

Researchers have proposed a variety of calculation intervals and computational methods to process oxygen uptake data [@howley1995; @robergs2003; @ats/accp2003; @weir2004; @robergs2010] and calls to standardize processing strategies are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. Midgley et al. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting the use of binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by Midgley et al. @midgley2007 was not systematic and its methods were not described in detail. Robergs et al. @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning [@robergs2010]. Practices in reporting and processing may have changed more than a decade after the publication of these numbers.

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. Martin-Rincon et al. @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts. Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies. Most comparisons only report mean differences between strategies [@astorino2009; @midgley2007]. No research has yet compared a variety of systematically derived strategies among a group of trained individuals using a standardized measurement set-up.

Differences in V̇O~2max~ due to data processing can have serious implications in practice. They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analyses [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

This paper aims to review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~. The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals. The review allows to assess the implementation of data processing routines and to find malpractices in reporting. The results build a basis for providing recommendations for the reporting of data processing strategies to determine the V̇O~2max~.

# Methods

This work was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker]. Any deviations from the preregistration are indicated in a 'Transparent Changes' document \[LINK\]. Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing). We conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022].

## Systematic Scoping Review

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, we randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans. The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018], see LINK for the CHECKLIST.

### Search & Screening

The article search was conducted on 16th March 2022 using [PubMed](https://pubmed.ncbi.nlm.nih.gov/) and [Web of Science](https://www.webofscience.com/wos/woscc/basic-search). The search included articles published from 2017 to the date of the search referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords. APPENDIX \[LINK\] shows the exact search terms used.

The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step. This was followed by an automated title scanning to exclude results that were likely no original research articles. All titles that contained one of the following words were excluded: *'review', 'correction', 'meta-analysis', 'comment', 'retraction', 'editorial', 'erratum', 'reply'.*

In accordance with the preregistration we drew a random sample from the search results. The goal of this process was to give an unbiased estimate of the current state of scientific V̇O~2max~ testing. Based on the procedure described in the preregistration, the sample included a total of 500 articles.

The abstracts from the articles included in the random sample were blinded for scanning, by removing any authors identities and journal information. Two of the authors (SN and OJQ) independently scanned the blinded abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion [LINK]. When the screeners disagreed in their assessment, they resolved the conflict by discussion.

After the abstract screening we retrieved the full-texts for the remaining articles. The full-texts were again independently scanned by two authors (SN and OJQ) to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria\[LINK\]). Conflicts were resolved by discussion between the two examiners.

### Data Extraction

We retrieved data from all articles remaining after the abstract and full-text screening. Extraction included the following data:

-   metabolic cart used
-   measurement type (breath-by-breath, mixing chamber)
-   type of outcome for V̇O~2max~ (primary, secondary, other)
-   data preprocessing (e.g., filtering)
-   data processing software
-   interpolation procedure
-   data processing/determination of V̇O~2max~:
    -   type (time average, breath average, digital filtering, ...)
    -   alignment (moving, binned, ...)
    -   interval (in seconds or breaths, parameters for filtering)
-   rationale for the used data processing strategy (e.g. a reference)

The criteria 'type of outcome' and 'rationale' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration.

### Data Synthesis

The extracted data is presented in a purely descriptive way. We calculated the relative and absolute frequency for the reporting of the extracted items. Similarly, we counted the use of different strategies for processing data in all articles that reportedly measured breath-by-breath. Total interval duration of averaging procedures were derived from the reported parameters.

## Experimental Comparison

To determine the influence of the most common data processing strategies on the determination of V̇O~2max~, we compared them on a set of already collected gas exchange data from ramp tests in running.

### Data Source

A total of N = 72 exercise tests were analyzed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data were from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b], respectively, while using identical exercise protocols and test equipment. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

Gas exchange data were recorded using a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America). The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement. The measured breath-by-breath data is available on [GitHub](https://github.com/smnnlt/vo2max-processing/tree/master/data/ramptests).

### Data Processing

The spiro package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data. The software includes various algorithms to calculate V̇O~2max~ with user-defined parameters. Moving time-based averages were calculated by first linearly interpolating the breath-by-breath data to full seconds. Subsequently a (center aligned) moving average was calculated over the defined timespan.

For binned time-averages, the breath-by-breath data were initially interpolated to full seconds and then binned into consecutive intervals of constant lengths. The average of each interval was aligned to its center. Incomplete intervals (i.e. the last seconds of measurement) were not considered in the analysis. Note that some authors use a different procedure for determining their bins, starting by the end point of the measurement. However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software (e.g. Cosmed Omnia). Breath based moving averages were calculated on the raw data.

### Comparison of Methods

To compare different processing methods within and between individuals, we chose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was the most commonly applied one in current literature as determined by our scoping review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. We calculated the data for all integer parameter values within the range of the calculation intervals found in the literature during the review. On a group level, we calculated the median and 10%- and 90%-quantiles of each processing strategy.

# Results

## Systematic Scoping Review

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in @fig-flow). Out of the random sample (n = 500), 242 articles were included in the final analysis.

```{r}
#| echo: false
#| fig-cap: "Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]"
#| fig-scap: "Flow diagram for the systematic scoping review."
#| label: fig-flow
#| out-width: '90%'

knitr::include_graphics("./plots/flow_diagram.png")
```

Reporting practices of the methodology of gas exchange measures differed widely across the literature (see @tbl-reporting). Almost half (44.2%) of the articles did not report any information regarding their data processing strategy. About one in twenty articles (5.8%) provided a rationale for their used strategy. Only a single article [@maturana2021] reported information regarding all the investigated criteria.

```{r}
#| echo: false
#| label: tbl-reporting
#| tbl-cap: "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing."
load("./data/reporting.Rda")
t <- purrr::map_dfr(reporting, scales::percent, accuracy = 0.1)

t[["pre"]] <- paste0(t[["pre"]], "*")
t[["software"]] <- paste0(t[["software"]], "*")

kableExtra::kbl(
  t,
  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Rationale"),
  align = "r",
  booktabs = TRUE
) |>
  kableExtra::footnote(
    general = "*only examined within the subgroup of studies using breath-by-breath measurements",
    general_title = ""
  )
```

Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~. Moving time averages, or breath-based averages were uncommon (see @fig-strategies). No study used methods of digital filtering to determine V̇O~2max~.

```{r}
#| echo: false
#| fig-cap: "Strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
#| label: fig-strategies
#| out-width: '90%'

knitr::include_graphics("./plots/strategies.png")
```

For preprocessing, some authors reported the use of a (linear) interpolation for the breath-by-breath data to seconds (n = 7; 4.3%). A minority of researchers reported the use of data filtering strategies to remove outliers. This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5). For the automated outlier detection authors removed single data points differing from a not further defined local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval. When reported, the software used for data processing varied among studies showing a total of more than 15 reported programs (for 30 studies that reported this parameter).

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30-second intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

```{r}
#| echo: false
#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies."
#| label: fig-duration
#| out-width: '90%'

knitr::include_graphics("./plots/duration_count.png")
```

## Experimental Comparison

The average V̇O~2max~ as determined by a binned 30-second average was 62.2 ± 6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation). Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see @fig-comparison).

```{r}
#| echo: false
#| fig-cap: "V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review. Solid lines display the median, the shaded areas mark the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
#| fig-scap: "Comparison of different data processing strategies."
#| label: fig-comparison
#| out-width: '90%'

knitr::include_graphics("./plots/comparison.png")
```

Binned time averages systematically generate lower V̇O~2max~ values compared to moving averages. Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values. Notably on an individual level these increases may be lower (\~2%) or much higher (\>10%) than the median value.

Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals, as most of the trained athletes reached respiratory rates around 60 min^-1^ in the final minutes of the exercise test (see SUPPL FIG LINK).

# Discussion

We aimed to review current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data. Our results show, that current research uses a variety of processing strategies to determine V̇O~2max~, which directly influences the values obtained. Identical raw breath-by-breath data can result in different V̇O~2max~ values when processed differently. Many articles only incompletely report on their methods, which hinders the reproducibility of V̇O~2max~ measurement.

## Current State of Data Processing

Despite calls to use moving averages or digital filters [@robergs2003; @robergs2010], binned time averages remain the most common data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies). Our numbers are generally in line with the findings of the non-systematic search by Midgley et al. @midgley2007 and the survey by Robergs et al. @robergs2010. It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that discourage researchers from using binned averages [@robergs2010]. Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison). The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~. Binned time averages revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data. Despite these arguments speaking against the use of binned time averages, our review demonstrates that they remain extremely common in the scientific literature.

Breath-based averages seem to be more common (\~8%) than reported previously (\<1%) [@midgley2007], but less common than assessed in self-reporting (\~17%) [@robergs2010]. The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010]. The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration). This may reflect contradictory recommendations [@howley1995; @robergs2003]. There is no consensus on using shorter or longer calculation intervals, and an is unclear whether an optimal interval duration even exists. As different interval durations influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.

A major source for choosing binned averages for data processing may be limitations by analysis software. Our review shows that most researchers use the software of the metabolic cart's manufacturer to analyze the gas exchange data. These software may by default output binned time averages instead of raw breath-by-breath data. Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software. This may also explain why digital filtering has --- despite recommended by Robergs et al. @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations. The common practices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.

## Impact of Different Data Processing Strategies

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @fig-duration); so there is no reason to compare it to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.

Binned time averages lead to systematically lower V̇O~2max~ values compared with moving averages, for the reasons explained above. While this general trend has been acknowledged previously [@martin-rincon2019], it has not been quantified. Our data suggest a \~1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length. This difference is well within the measurement error of most if not all metabolic carts, but it is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g. in elite sports).

Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison). This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see SUPPL LINK), resulting in equivalent time- and breath-based interval lengths. For an athletic population, V̇O~2max~ values obtained by moving time and moving breath averages can approximately be used interchangeably. Given that less trained individuals display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will likely not translate to a sedentary population.

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. On an individual level, data processing strategies may impact the V̇O~2max~ in different severity. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications. However, on an individual level these equations can only be applied with a high margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.

## Guidelines for Reporting

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. Our review demonstrates that almost half of the studies measuring V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or the rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:

> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America). The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro package for R version 0.1.0 [@nolte2022]. This strategy produces similar results as that recommended by Robergs et al. @robergs2010, but does not include a time lag."*

Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing. Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test. In cases where journals endorse word limits on articles, this reporting --- including the reporting on data processing strategies --- may be included in supplementary files. The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.

The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies. But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status. Sharing of the raw gas exchange data can solve this challenge, as it allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy. Most raw gas exchange data files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before). In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code. This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment. Such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].

```{r}
#| echo: false
#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake."
#| label: tbl-recommendation

rec_df <- data.frame(
  item = c("Metabolic Cart", "Measurement Mode", "Software", "Preprocessing", "Processing Strategy", "Processing Parameters", "Rationale"),
  desc = c(
    "State the exact device model and manufacturer.",
    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…).",
    "State the name and version of the software used for data analysis.",
    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis.",
    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages).",
    "State the parameters used for the processing strategy (e.g., length of averaging interval).",
    "State the rationale for using the processing strategy (e.g., reference to recommendations).")
)

out <- kableExtra::kbl(
  rec_df,
  col.names = c("Reporting Item", "Description"),
  # format = "latex",
  booktabs = TRUE
)

# knitr::asis_output(stringr::str_replace_all(out, "VO2max", "${\\\\dot{\\\\textsf{V}}\\\\textsf{O}_{\\\\textsf{2max}}}$"))

out
```

## Limitations

Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of the literature. The scoping review therefore relies on a random sample which not necessarily captures exact trends of the literature. However, efforts were made --- such as random sampling and systematic article exclusions --- to ensure the sample to be representative. Notable, almost half of the studies did not report their data processing strategy at all. The data processing strategies used in the literature could only be investigated when studies reported them.

Ambiguities in the reporting of the investigated studies may impact the analysis results. For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g., 4x15 second), without describing this correctly. Moreover the exact definitions for building binned averages vary within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint. Additionally some studies reviewed did not define the maximum bin, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value). In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing. We did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguities in their reporting.

This work treated each breath as the single data processing unit of cardiopulmonary exercise testing. However, metabolic carts sample gas fraction and gas flow data at a much higher frequency (e.g., 50 Hz). The data for each breath is subsequently calculated from these raw signals by the means of an algorithm. Different algorithms to generate the breath-by-breath data can lead to different outcomes [@koschate2019], and as such may also influence the V̇O~2max~. Hence documenting and reporting of the breath-by-breath algorithm is advisable. Yet many metabolic carts do not describe their default algorithm and limit access to the raw data signal.

The experimental comparison of different data processing strategies was conducted on a standardized data set of exercise tests. This standardization in terms of training status, exercise protocol, and measurement device helps to highlight the impact of different data processing even in a relatively homogeneous data set. However the results may only partly transfer to different settings, such as individuals with less training background.

# References

### 2022-11-04 paper.qmd
Add changes by RR

@@
 
-**Background** Gas exchange data from maximum oxygen uptake (V̇O~2max~) testing needs post-processing.
-Different processing strategies may lead to varying V̇O~2max~ values in individuals.
-The exact processing strategies used in the literature have not yet been systematically investigated.
-Previous research provided comparisons of methods only on a group level.
+**Background** Gas exchange data from maximum oxygen uptake (V̇O~2max~) testing typically requires post-processing.
+Different processing strategies may lead to varying V̇O~2max~ values affecting their interpretation.
+However, the exact processing strategies used in the literature have not been systematically investigated yet.
+Previous research investigated differences across methods at the group level only.
 
-**Methods** Out of a random sample, we investigated 242 recently published articles that measured V̇O~2max~ in an appropriate setting.
-We extracted the reported methods of data processing and their rationale.
+**Methods** Out of a random sample, we investigated 242 recently published articles which measured V̇O~2max~ in an appropriate setting(Was ist das?).
+Reported data processing methods and their rationale were extracted.
 We compared the most common processing strategies on a data set of 72 standardized exercise tests in trained athletes.
 
-**Results** Most researchers use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages.
-Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the strategy used.
-The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% with considerable variation on an individual level.
+**Results** Most studies use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages.
+Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen.
+The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% [RANGE_min, RANGE_max] with considerable variation at the individual level.
 
@@
 -   Despite calls to use moving averages or digital filter, binned averages remain the most common data processing strategy to determine V̇O~2max~.
--   Different processing strategies lead to varying V̇O~2max~ values, which can hardly be predicted on an individual level.
+-   Different processing strategies lead to varying V̇O~2max~ values, which can hardly be predicted on an individual level (Verstehe ich nicht).
 -   Researchers should at minimum report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.
@@
 The maximum oxygen uptake (V̇O~2max~) is one of the most commonly assessed physiological parameters in sports and exercise science [@bassett2000].
-The V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological predictor of endurance performance [@bassett2000].
-Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~
+V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded as on of the most relevant physiological predictor of endurance performance [@bassett2000].
+Accordingly, many training programs target V̇O~2max~ as the performance criteria and evaluate for example the effectiveness of treatment by measuring changes in V̇O~2max.~
 
@@
 mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994].
+(HIER EIN SATZ WARUM DIE DATEN ÜBERHAUPT PROZESSIERT WERDEN MÜSSEN).
 But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].
 
-Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018].
+Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in previous research [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018].
 To identify a true maximum, researchers commonly evaluate a set of parameters measured during ramp tests --- the so called 'V̇O~2max~ criteria' [@howley1995].
@@
 
-Measured gas exchange data is noisy.
-Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) lead to a highly fluctuating raw oxygen uptake on a breath-by-breath basis.
-For interpretation the raw data requires some form of processing.
-Different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021].
+Measured gas exchange data is inherently noisy.
+Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) result in large fluctuation of raw oxygen uptake data on a breath-by-breath basis.
+To interpret the results the raw data therefore requires some form of processing to obtain a more stable estimate.
+However, different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021].
 Midgley et al.
-@midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, inferring that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes.
-Based on own data in sedentary and moderately trained individuals, Martin-Rincon et al.
+@midgley2007 found differences in estimated VO~2max~ values between data processing strategies.
+As no difference regarding the reliability of the different approaches was found, no optimal strategy exists per se.
+Accordingly, when comparing outcomes within and across studies obviously a consistent processing strategy should be used.
+Based on own(???) data in sedentary and moderately trained individuals, Martin-Rincon et al.
 @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
-There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
+The results showed ...
+.
+Furhtermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
 
@@
 While providing interesting first insights into reporting and processing practices, the search by Midgley et al.
-@midgley2007 was not systematic and its methods were not described in detail.
+@midgley2007 was not systematic and its methods were not described in detail (DER SATZ IST GEFHÄHRLICH UND SOLLTE POSITIVER WAS DER ARTIKEL ÜBER DIE ERGEBNISSE VON MIDGLEY HINAUS NEUES BRINGT).
+
 Robergs et al.
-@robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*.
-They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies.
+@robergs2010 argued that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*.
+They chose the latter approach with a total of 75 respondents, who reported a great variety in data processing strategies.
 Most researchers reported the use of binned time averages over 30 or 60 seconds.
-Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning [@robergs2010].
-Practices in reporting and processing may have changed more than a decade after the publication of these numbers.
+Surprinsingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010].
+The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by @midgley2007 and @robergs2003.
 
 Selected data processing strategies have been extensively compared in the literature.
-Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare.
+Due to the absence of a systematic mapping of current practices, these studies however lacked the reasoning of which strategies to compare.
 Many studies compared different averaging intervals, but not averaging types (e.g.
@@
 
-Differences in V̇O~2max~ due to data processing can have serious implications in practice.
-They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analyses [@martin-rincon2019].
+Differences in estimated V̇O~2max~ vaules due to variation in data processing can have serious implications in practice.
+For example, the assessment of longitudinal data from athletes which  participated in diagnostics using differing data processing approaches becomes problematic or the pooling of results across studies in meta-analyses [@martin-rincon2019].
 Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009].
@@
 
-This paper aims to review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~.
+This paper aims to review the usage and reporting of different data processing strategies in the scientific literature and investigates their influence on V̇O~2max~.
 The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals.
-The review allows to assess the implementation of data processing routines and to find malpractices in reporting.
+The review allows to assess the implementation of data processing routines and to identify problematic reporting strategies.
 The results build a basis for providing recommendations for the reporting of data processing strategies to determine the V̇O~2max~.
@@
 
-This work was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker].
+The present work was preregistered before project start with the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker].
 Any deviations from the preregistration are indicated in a 'Transparent Changes' document \[LINK\].
 Major deviations will also be explicitly stated within the methods section.
-All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).
-We conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022].
+All data and code of this research project can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).
+All analyses were performed using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022].
 
@@
 The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature.
-Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, we randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords.
-Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.
+Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, we randomly sampled 500 articles which referred to V̇O~2max~ or similar keywords.
+Data on processing strategies were extracted from all sampled articles which directly measured V̇O~2max~ using an appropriate testing procedure in human subject.
 The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018], see LINK for the CHECKLIST.
@@
 Moving time averages, or breath-based averages were uncommon (see @fig-strategies).
-No study used methods of digital filtering to determine V̇O~2max~.
+No study used digital filtering methods to determine V̇O~2max~.
+HIER FÄNDE ICH EINE TABELLE BESSER.
+ZIEMLICH VIEL PLATZ FÜR WENIG INFO.
 
@@
 For preprocessing, some authors reported the use of a (linear) interpolation for the breath-by-breath data to seconds (n = 7; 4.3%).
-A minority of researchers reported the use of data filtering strategies to remove outliers.
+Few studies reported the use of data filtering strategies to remove outliers.
 This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5).
@@
 The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
-30-second intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.
+30-second intervals were most common to define V̇O~2max~, while some authors also often employed shorter (10-20 s) and longer (60 s) periods.
+(HIER GLAUBE ICH AUCH LIEBER TABELLE)
 
@@
 The average V̇O~2max~ as determined by a binned 30-second average was 62.2 ± 6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation).
-Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see @fig-comparison).
+Applying different data processing strategies for V̇O~2max~ determination lead to different outcome values (see @fig-comparison).
 
@@
 
-Binned time averages systematically generate lower V̇O~2max~ values compared to moving averages.
-Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values.
-Notably on an individual level these increases may be lower (\~2%) or much higher (\>10%) than the median value.
+(SOLLTE DIE BESCHRIFTUNG AUF DER X-ACHSE NICHT DAUER SEIN)
+
+Binned time averages systematically generated lower V̇O~2max~ values compared to moving averages.
+Decreasing the calculation interval to 5 seconds lead to a 3-4% median increase of V̇O~2max~ values.
+Notably at the individual level these increases may be lower (\~2%) or much higher (\>10%) than the median value.
 
@@
 Identical raw breath-by-breath data can result in different V̇O~2max~ values when processed differently.
-Many articles only incompletely report on their methods, which hinders the reproducibility of V̇O~2max~ measurement.
+Many articles provide only incomplete reports about their chosen methods, which hinders reproducibility of V̇O~2max~ measurement.
 
@@
 Despite calls to use moving averages or digital filters [@robergs2003; @robergs2010], binned time averages remain the most common data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies).
-Our numbers are generally in line with the findings of the non-systematic search by Midgley et al.
+The present findings are generally in agreement with the findings of the non-systematic search by Midgley et al.
 @midgley2007 and the survey by Robergs et al.
 @robergs2010.
-It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that discourage researchers from using binned averages [@robergs2010].
+It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations discouraging researchers from using binned averages [@robergs2010].
 Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison).
 The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~.
-Binned time averages revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data.
-Despite these arguments speaking against the use of binned time averages, our review demonstrates that they remain extremely common in the scientific literature.
+Binned time averages undermine the most important argument in favour of measuring breath-by-breath: the high temporal resolution of data.
+Despite these arguments speaking against the use of binned time averages, the present review demonstrates that they remain extremely common in the scientific literature.
 
@@
 This may reflect contradictory recommendations [@howley1995; @robergs2003].
-There is no consensus on using shorter or longer calculation intervals, and an is unclear whether an optimal interval duration even exists.
-As different interval durations influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
+There is no consensus on using shorter or longer calculation intervals, and it is unclear whether an optimal interval duration even exists.
+As different interval durations influence V̇O~2max~ by more than 5% with respect to median levels (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.
 
-A major source for choosing binned averages for data processing may be limitations by analysis software.
-Our review shows that most researchers use the software of the metabolic cart's manufacturer to analyze the gas exchange data.
+One possible reason for the prevalence of binned averages data processing approaches may be due to limitations in analysis software used.
+The results show that most researchers use the vendor software of the metabolic cart's manufacturer.
 These software may by default output binned time averages instead of raw breath-by-breath data.
@@
 
+(HIER WÄRE NOCH ZUSÄTZLICHE INFO WAS TYPISCHE UNTERSCHIEDE IN INTERVENTIONSSTUDIEN SIND UM DIE VARIATION ZWISCHEN DEN METHODEN BESSER EINSCHÄTZEN ZU KÖNNEN).
+
 Binned time averages lead to systematically lower V̇O~2max~ values compared with moving averages, for the reasons explained above.
 While this general trend has been acknowledged previously [@martin-rincon2019], it has not been quantified.
-Our data suggest a \~1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
+The present data suggest a \~1% lower median V̇O~2max~ when using binned averages compared with moving averages of the same calculation interval length.
 This difference is well within the measurement error of most if not all metabolic carts, but it is systematic and as such may bias the evaluation in scenarios where small changes in V̇O~2max~ are important (e.g.
@@
 
-Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison).
+Moving time and moving breath averages lead to almost identical V̇O~2max~ values with respect to median values (@fig-comparison).
 This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see SUPPL LINK), resulting in equivalent time- and breath-based interval lengths.
 For an athletic population, V̇O~2max~ values obtained by moving time and moving breath averages can approximately be used interchangeably.
-Given that less trained individuals display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will likely not translate to a sedentary population.
+Given that less trained individuals display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will likely not translate to sedentary populations.
 
@@
 Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019].
-On an individual level, data processing strategies may impact the V̇O~2max~ in different severity.
+At the individual level, data processing strategies may impact V̇O~2max~ values with varying magnitudes.
 For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
-Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications.
-However, on an individual level these equations can only be applied with a high margin of error.
+Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparizon of group results for meta-analyses or group classifications.
+However, at the individual level these equations can only be applied with a large margin of error.
 Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others.
@@
 
-The results of this thesis suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
-But on an individual level and for a precise comparison, reporting is not enough, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
+The results of the present work suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
+But on an individual level and for a precise comparison, reporting may not be sufficient, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
 Sharing of the raw gas exchange data can solve this challenge, as it allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
@@
 
-Ambiguities in the reporting of the investigated studies may impact the analysis results.
+Ambiguities in the reporting of the investigated studies may impact analysis results.
 For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g., 4x15 second), without describing this correctly.

### 2022-11-17 paper.qmd
Adjust to feedback by RR

@@
 
-**Methods** Out of a random sample, we investigated 242 recently published articles which measured V̇O~2max~ in an appropriate setting(Was ist das?).
+**Methods** Out of a random sample, we investigated 242 recently published articles which measured V̇O~2max~ during ramp tests.
 Reported data processing methods and their rationale were extracted.
@@
 Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen.
-The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% [RANGE_min, RANGE_max] with considerable variation at the individual level.
+The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% (range 0-7%) with considerable variation at the individual level.
 
@@
 -   Despite calls to use moving averages or digital filter, binned averages remain the most common data processing strategy to determine V̇O~2max~.
--   Different processing strategies lead to varying V̇O~2max~ values, which can hardly be predicted on an individual level (Verstehe ich nicht).
+-   Different processing strategies lead to varying V̇O~2max~ values at the median as well as at the individual level.
 -   Researchers should at minimum report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.
@@
 mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994].
-(HIER EIN SATZ WARUM DIE DATEN ÜBERHAUPT PROZESSIERT WERDEN MÜSSEN).
-But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].
+Due to its inherent variation, the collected raw data data can hardly be interpreted without processing.
+The same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].
 
@@
 Accordingly, when comparing outcomes within and across studies obviously a consistent processing strategy should be used.
-Based on own(???) data in sedentary and moderately trained individuals, Martin-Rincon et al.
+Based on data in sedentary and moderately trained individuals, Martin-Rincon et al.
 @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
-The results showed ...
-.
-Furhtermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
+The results showed that reliability of V̇O~2max~ was higher in trained individuals, but not influenced by the data processing.
+Furthermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
 
@@
 One third of the studies did not describe their processing method at all.
-While providing interesting first insights into reporting and processing practices, the search by Midgley et al.
-@midgley2007 was not systematic and its methods were not described in detail (DER SATZ IST GEFHÄHRLICH UND SOLLTE POSITIVER WAS DER ARTIKEL ÜBER DIE ERGEBNISSE VON MIDGLEY HINAUS NEUES BRINGT).
+In our article we will provide a systematic analysis of the reporting and usage of data processing strategies in recent literature.
 
@@
 Most researchers reported the use of binned time averages over 30 or 60 seconds.
-Surprinsingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010].
+Surprisingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010].
 The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by @midgley2007 and @robergs2003.
@@
 
-Differences in estimated V̇O~2max~ vaules due to variation in data processing can have serious implications in practice.
+Differences in estimated V̇O~2max~ values due to variation in data processing can have serious implications in practice.
 For example, the assessment of longitudinal data from athletes which participated in diagnostics using differing data processing approaches becomes problematic or the pooling of results across studies in meta-analyses [@martin-rincon2019].
@@
 Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
-The data were from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b].
+The data were from previous research on the metabolic profile of endurance runners [@quittmann_accepted; @schwarz2022].
 The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies).
-The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b], respectively, while using identical exercise protocols and test equipment.
+The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_accepted] and March to October 2021 [@schwarz2022], respectively, while using identical exercise protocols and test equipment.
 Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up.
@@
 Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~.
-Moving time averages, or breath-based averages were uncommon (see @fig-strategies).
+Moving time averages, or breath-based averages were uncommon (see @tbl-strategies).
 No study used digital filtering methods to determine V̇O~2max~.
-HIER FÄNDE ICH EINE TABELLE BESSER.
-ZIEMLICH VIEL PLATZ FÜR WENIG INFO.
 
@@
 #| echo: false
-#| fig-cap: "Strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
-#| label: fig-strategies
-#| out-width: '90%'
+#| label: tbl-strategies
+#| tbl-cap: "Strategies for processing breath-by-breath data in the reviewed literature (n = 88)"
 
-knitr::include_graphics("./plots/strategies.png")
+load("./data/bbb_strategies.Rda")
+
+e_sort <- ext_count[order(ext_count$n, decreasing = TRUE), ]
+
+knitr::kable(
+  e_sort[, c("proc_combination", "perc")],
+  col.names = c("Processing Strategy", "Usage"),
+  align = "r",
+  row.names = FALSE
+)
 ```
@@
 30-second intervals were most common to define V̇O~2max~, while some authors also often employed shorter (10-20 s) and longer (60 s) periods.
-(HIER GLAUBE ICH AUCH LIEBER TABELLE)
 
@@
 #| echo: false
-#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies."
+#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. ROBERT SCHLÄGT VOR HIER LIEBER EINE TABELLE ZU NEHMEN. ICH BIN FÜR DIE GRAFIK. WAS MEINST DU?."
 #| label: fig-duration
@@
 Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
-Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
+Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%.
+XLABEL GEÄNDERT IN 'DURATION' STATT 'PARAMETER'.
+OKAY?"
 #| fig-scap: "Comparison of different data processing strategies."
@@
 
-knitr::include_graphics("./plots/comparison.png")
+knitr::include_graphics("./plots/comparison_xlabel.png")
 ```
-
-(SOLLTE DIE BESCHRIFTUNG AUF DER X-ACHSE NICHT DAUER SEIN)
 
@@
 
-Despite calls to use moving averages or digital filters [@robergs2003; @robergs2010], binned time averages remain the most common data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies).
+Despite calls to use moving averages or digital filters [@robergs2003; @robergs2010], binned time averages remain the most common data processing strategy to determine V̇O~2max~ in the reviewed literature (see @tbl-strategies).
 The present findings are generally in agreement with the findings of the non-systematic search by Midgley et al.
@@
 The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~.
-Binned time averages undermine the most important argument in favour of measuring breath-by-breath: the high temporal resolution of data.
+Binned time averages undermine the most important argument in favor of measuring breath-by-breath: the high temporal resolution of data.
 Despite these arguments speaking against the use of binned time averages, the present review demonstrates that they remain extremely common in the scientific literature.
@@
 While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
-
-(HIER WÄRE NOCH ZUSÄTZLICHE INFO WAS TYPISCHE UNTERSCHIEDE IN INTERVENTIONSSTUDIEN SIND UM DIE VARIATION ZWISCHEN DEN METHODEN BESSER EINSCHÄTZEN ZU KÖNNEN).
+Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies can bias the evaluation of their success.
 
@@
 For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
-Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparizon of group results for meta-analyses or group classifications.
+Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications.
 However, at the individual level these equations can only be applied with a large margin of error.
@@
 > *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America).
-The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro package for R version 0.1.0 [@nolte2022].
+The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro package for R version 0.0.4 [@nolte2022].
 This strategy produces similar results as that recommended by Robergs et al.

### 2022-11-17 scripts/comparison.Rmd
Adjust to feedback by RR

@@
 
-p_comp <- ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
+p_comp_pre <- ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
   geom_hline(
@@
   ) +
-  scale_x_continuous(
-    name = "Parameter",
-    breaks = c(15, 30, 45, 60)
-  ) +
   scale_y_continuous(
@@
     values = met.brewer("Java", 3)
+  )
+
+p_comp <- p_comp_pre +
+  scale_x_continuous(
+    name = "Parameter",
+    breaks = c(15, 30, 45, 60)
   ) +
@@
 knitr::include_graphics("../plots/comparison.png")
+
+# create plot with 'Duration' as x label
+p_comp_duration <- p_comp_pre +
+  scale_x_continuous(
+    name = "Duration",
+    breaks = c(15, 30, 45, 60)
+  ) +
+  theme_classic() +
+  theme(
+    legend.position = c(0.8,0.8)
+  )
+# ggsave("../plots/comparison_xlabel.png", plot = p_comp_duration, width = 6, height = 4, dpi = 600, bg = "white")
 ```

### 2022-11-17 scripts/review.Rmd
Adjust to feedback by RR

@@
 percs <- scales::percent(ext_count$n / sum(ext_count$n))
+ext_count$perc <- percs
 ext_count$label <- paste0(ext_count$proc_combination, " (", percs, ")")
 ext_count$subgroup <- c(1, 1, 2, 3, 4)
+
+# save(ext_count, file = "../data/bbb_strategies.Rda")
 

### 2022-11-25 paper.qmd
Adjust to feedback by OJQ

@@
 
-**Results** Most studies use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages.
-Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen.
+**Results** Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen.
+Most studies use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages.
 The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% (range 0-7%) with considerable variation at the individual level.
@@
 Among the factors that influence the V̇O~2max~ determination are the exercise test protocol [@buchfuhrer1983 ; @yoon2007; @midgley2008], exercise modality [@myers1991], motivation [@midgley2018], and biological variability [@katch1982].
-The measuring method (e.g.
+Even the measuring method (e.g.
 mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994].
-Due to its inherent variation, the collected raw data data can hardly be interpreted without processing.
-The same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].
+Due to its inherent variation, the collected raw data can hardly be interpreted without processing.
+The same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009], which can have serious implications in practice [@johnson1998].
 
@@
 Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) result in large fluctuation of raw oxygen uptake data on a breath-by-breath basis.
-To interpret the results the raw data therefore requires some form of processing to obtain a more stable estimate.
+Therefore, the raw data require some form of processing to obtain a more stable estimate.
 However, different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021].
 Midgley et al.
-@midgley2007 found differences in estimated VO~2max~ values between data processing strategies.
+@midgley2007 found differences in estimated V̇O~2max~ values between data processing strategies.
 As no difference regarding the reliability of the different approaches was found, no optimal strategy exists per se.
@@
 @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths.
-The results showed that reliability of V̇O~2max~ was higher in trained individuals, but not influenced by the data processing.
+The results showed that the reliability of V̇O~2max~ was higher in trained individuals, but not influenced by data processing.
 Furthermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].
@@
 One third of the studies did not describe their processing method at all.
-In our article we will provide a systematic analysis of the reporting and usage of data processing strategies in recent literature.
 
@@
 Surprisingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010].
-The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by @midgley2007 and @robergs2003.
+The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by Midgley et al.
+[@midgley2007] and Robergs et al.
+[@robergs2003].
 
@@
 @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts.
-Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.
+Therefore in their work motivation and measurement devices may have interacted with the influence of processing strategies.
 Most comparisons only report mean differences between strategies [@astorino2009; @midgley2007].
@@
 Differences in estimated V̇O~2max~ values due to variation in data processing can have serious implications in practice.
-For example, the assessment of longitudinal data from athletes which participated in diagnostics using differing data processing approaches becomes problematic or the pooling of results across studies in meta-analyses [@martin-rincon2019].
+For example, the assessment of longitudinal data from athletes which participated in diagnostics using differing data processing approaches becomes problematic.
+The same applies to the pooling of results across studies in meta-analyses [@martin-rincon2019].
 Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009].
@@
 #| echo: false
-#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. ROBERT SCHLÄGT VOR HIER LIEBER EINE TABELLE ZU NEHMEN. ICH BIN FÜR DIE GRAFIK. WAS MEINST DU?."
+#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. STILL UP TO DEBATE WHETHER RATHER PLOT (VERTICAL OR HORIZONTAL BARS?) OR TABLE."
 #| label: fig-duration
@@
 Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages.
-Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%.
-XLABEL GEÄNDERT IN 'DURATION' STATT 'PARAMETER'.
-OKAY?"
+Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
 #| fig-scap: "Comparison of different data processing strategies."
@@
 
-knitr::include_graphics("./plots/comparison_xlabel.png")
+# keeping 'Parameter' as x label, as 'Duration' may be improper
+# when describing breath-based averages.
+knitr::include_graphics("./plots/comparison.png")
 ```
@@
 The results show that most researchers use the vendor software of the metabolic cart's manufacturer.
-These software may by default output binned time averages instead of raw breath-by-breath data.
+These software products may by default output binned time averages instead of raw breath-by-breath data.
 Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software.
@@
 @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations.
-The common practices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.
+The common practices of data processing are likely to be attributed to both: a lack of awareness and adequate software solutions.
 
@@
 While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
-Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies can bias the evaluation of their success.
+Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies is in the same range and can bias the evaluation of their success.
 
@@
 Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019].
-At the individual level, data processing strategies may impact V̇O~2max~ values with varying magnitudes.
-For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison).
+Data processing strategies may impact V̇O~2max~ values with varying magnitudes at the individual level.
+For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% of the investigated athletes the V̇O~2max~ was \>6% higher (see @fig-comparison).
 Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications.
@@
 Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported.
-While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.
+While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, the raw data from each test is required on the individual level.
 
@@
 These items may be reported in form of a checklist, as an in-text enumeration or in sentence form.
-An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:
+An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_accepted; @schwarz2022] would be:
 
@@
 Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing.
-Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test.
+Further aspects to be reported include, but are not limited to: the study population, exercise protocol, device calibration, and criteria to terminate the test.
 In cases where journals endorse word limits on articles, this reporting --- including the reporting on data processing strategies --- may be included in supplementary files.
@@
 The results of the present work suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies.
-But on an individual level and for a precise comparison, reporting may not be sufficient, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
+However, on an individual level and for a precise comparison, reporting may not be sufficient, as differences between data processing strategies vary between individuals and are potentially influenced by training status.
 Sharing of the raw gas exchange data can solve this challenge, as it allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy.
 Most raw gas exchange data files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before).
-In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code.
-This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment.
+In terms of reproducibility of V̇O~2max~ determination, sharing the raw data as well as the data analysis code seems to be an even superior way.
+This requires the data analysis to take place in a programming (or at least code generating) environment.
 Such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].
@@
 
-Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of the literature.
+Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of all articles.
 The scoping review therefore relies on a random sample which not necessarily captures exact trends of the literature.
@@
 Ambiguities in the reporting of the investigated studies may impact analysis results.
-For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g., 4x15 second), without describing this correctly.
-Moreover the exact definitions for building binned averages vary within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint.
-Additionally some studies reviewed did not define the maximum bin, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
+For example some studies using long binned averages (e.g.
+60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g.
+4×15 second), without describing this correctly.
+Moreover, the exact definitions for building binned averages vary within the literature.
+While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint.
+Additionally some of the included studies did not define the maximum bin, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value).
 In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing.
-We did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguities in their reporting.
+We did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to investigate precisely due to ambiguities in their reporting.
 
@@
 
+DISCUSS IF WE SHOULD CHANGE THE ORDER OF RECOMMENDATIONS/GUIDELINES AND LIMITATIONS.
+
 # References

### 2023-07-04 scripts/comparison.Rmd
Add plots
revision and congress version

--- 2022-05-23 scripts/comparison.Rmd
@@
 library(MetBrewer)
+library(patchwork)
+library(lmerTest)
+library(broom.mixed)
+library(multcomp)
+library(ggdist)
 
@@
 
-## Normalize results
+## Plot: Comparison of various strategies
+
+### Normalize results
 
@@
 
-## Turn into tidy format
+### Turn into tidy format
 
@@
 
-## Plot: Comparison of processing strategies
+### Create comparison plot
 
@@
 
-p_comp <- ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type)) +
+p_comp_pre <- ggplot(res_tidy, aes(x = parameter, y = perdiff, colour = type, fill = type, linetype = type)) +
   geom_hline(
@@
   ) +
+  scale_color_manual(
+    name = "Processing Strategy",
+    values = met.brewer("Java", 3)
+  ) +
+  scale_fill_manual(
+    name = "Processing Strategy",
+    values = met.brewer("Java", 3)
+  ) +
+  scale_linetype_discrete(
+    name = "Processing Strategy"
+  )
+
+p_comp <- p_comp_pre +
   scale_x_continuous(
@@
   ) +
-  scale_color_manual(
-    name = "Processing Strategy",
-    values = met.brewer("Java", 3)
-  ) +
-  scale_fill_manual(
-    name = "Processing Strategy",
-    values = met.brewer("Java", 3)
-  ) +
   theme_classic() +
@@
 
-knitr::include_graphics("../plots/comparison.png")
+# save plot as EPS
+# ggsave("../plots/comparison.pdf", plot = p_comp, width = 6, height = 4, dpi = 1200, bg = "white")
+
+# create plot with different labels
+
+p_comp_duration <- p_comp_pre +
+  scale_x_continuous(
+    name = "calculation interval length (seconds or breaths)",
+    breaks = c(15, 30, 45, 60)
+  ) +
+  scale_y_continuous(
+    name = expression(paste(dot(V), O[2][max], " (% to reference)")),
+    labels = scales::percent,
+    limits = c(-0.053,0.1)
+  ) +
+  theme_classic() +
+  theme(
+    legend.position = c(0.8,0.8)
+  )
+# ggsave("../plots/comparison_xlabel.png", plot = p_comp_duration, width = 6, height = 4, dpi = 600, bg = "white")
+# ggsave("../plots/comparison_xlabel.pdf", plot = p_comp_duration, width = 6, height = 4, dpi = 1200, bg = "white")
+
+knitr::include_graphics("../plots/comparison_xlabel.png")
 ```
 
-## Plot: Respiratory Rate over time
+## Plot: Respiratory rate and VO2 over time
 
@@
   # Zero-lag Butterworth filter with default parameters
-  rr <- spiro_smooth(dat, smooth = "fz", columns = "RR")[, 1]
+  rr <- spiro_smooth(dat, smooth = "f", columns = "RR")[, 1]
+  vo2 <- spiro_smooth(dat, smooth = "f", columns = "VO2")[, 1] / attr(dat, "info")$weight
   time <- attr(dat, "raw")$time
@@
     time = time,
-    value = rr,
+    value_rr = rr,
+    value_vo2 = vo2,
     load = load,
@@
 # using a scatter plot instead
-p_rr <- ggplot(rr_all, aes(x = time, y = value, colour = is_end, alpha = is_end)) +
+p_rr <- ggplot(rr_all, aes(x = time, y = value_rr, colour = is_end, alpha = is_end)) +
+  geom_point(size = 0.5, show.legend = FALSE) +
+  scale_colour_manual(values = c("black", met.brewer("Peru2")[2])) +
+  scale_alpha_manual(values = c(0.05, 0.3)) +
+  scale_x_continuous(
+    name = NULL,
+    limits = c(60,950),
+    breaks = c(180, 360, 540, 720, 900),
+    labels = c(2, 5, 8, 12, 15),
+    expand = c(0,0)
+  ) +
+  scale_y_continuous(
+    name = expression(paste("Respiratory Rate ", (min^-1))),
+    breaks = c(20,40,60,80)
+  ) +
+  theme_classic()
+
+p_vo2 <- ggplot(rr_all, aes(x = time, y = value_vo2, colour = is_end, alpha = is_end, group = id)) +
   geom_point(size = 0.5, show.legend = FALSE) +
@@
   scale_y_continuous(
-    name = expression(paste("Respiratory Rate ", (min^-1))),
-    breaks = c(20,40,60,80)
+    name = expression(paste(dot(V), O[2], " uptake ", (ml%.%min^-1%.%kg^-1)))
   ) +
@@
 
-# ggsave(filename = "../plots/rr.png", plot = p_rr, width = 6, height = 4, dpi = 600, bg = "white")
+p_all <- p_rr / p_vo2 + plot_annotation(tag_levels = 'A')
 
-knitr::include_graphics("../plots/rr.png")
-```
+# ggsave(filename = "../plots/timecourse.png", plot = p_all, width = 6, height = 7, dpi = 600, bg = "white")
+# ggsave(filename = "../plots/timecourse.pdf", plot = p_all, width = 6, height = 7, dpi = 1200, bg = "white")
 
-## Calculate zero phase digital filter
+p_all_vert <- (p_rr | p_vo2) + plot_annotation(tag_levels = 'A')
+# ggsave(filename = "../plots/timecourse_vert.png", plot = p_all_vert, width = 8, height = 3, dpi = 600, bg = "white")
 
-```{r}
-# Function to get VO2max with digital filtering
-get_fzmax <- function(i) {
-  dta <- spiro(files[i], weight = info$bodymass[i])
-  spiro_max(dta, smooth = "0.04fz3")$VO2_rel
-}
-
-# iterate over all tests
-fzmax <- purrr::map_dbl(seq_along(files), get_fzmax)
-
-# normalize to reference procedure
-fzmax_norm <- fzmax / res$tb_30
-
-# get median and quantile values
-median(fzmax_norm)
-quant(fzmax_norm, 0.1, 0.9)
+knitr::include_graphics("../plots/timecourse.png")
 ```
@@
 knitr::include_graphics("../plots/example.png")
+
+## ECSS presentation plot
+p2_raw <- ggplot(exm_tidy, aes(x = time)) +
+  geom_point(
+    data = exm_tidy[exm_tidy$method == "raw", ], 
+    aes(y = value, alpha = method, colour = "Raw Data"),
+    show.legend = FALSE
+  ) +
+  scale_x_continuous(
+    name = "Time (min)",
+    limits = c(500,820),
+    breaks = c(540, 630, 720),
+    labels = c(8, 10, 12),
+    expand = c(0,0)
+  ) +
+  scale_y_continuous(
+    name = expression(paste(dot(V), O[2], " ", (ml%.%min^-1%.%kg^-1))),
+    limits = c(45,75)
+  ) +
+  scale_colour_manual(values = "black") +
+  scale_fill_manual(
+    values = "white",
+    guide = "none"
+  ) +  scale_alpha_manual(
+    values = 0.3,
+    guide = "none"
+  ) +
+  labs(title = "Raw data") +
+  theme_classic(15) +
+  theme(
+    legend.position = c(0.88, 0.63),
+  )
+
+# ggsave(filename = "../plots/example2_raw.png", plot = p2_raw, width = 6, height = 4, dpi = 1200, bg = "white")
+
+fz2 <- spiro_smooth(exm_data, "fz", "VO2")[,1]
+mb2 <- spiro_smooth(exm_data, "7b", "VO2")[,1]
+mt2 <- spiro_smooth(exm_data, 60, "VO2")[,1]
+tb2 <- rep(max_time_binned(exm_data, 30, max = FALSE) * attr(exm_data, "info")[["weight"]], each = 30)
+
+exm_tidy2 <- data.frame(
+  method = c(rep.int("raw", length(raw_vo2)), rep.int("fz", length(fz2)), rep.int("mb", length(mb2)), rep.int("mt", length(mt2)), rep.int("tb", length(tb2))),
+  value = c(raw_vo2, fz2, mb2, mt2, tb2),
+  time = c(rep.int(raw_time, 3), seq_along(mt2), seq_along(tb2))
+)
+
+exm_tidy2$method <- factor(
+  exm_tidy2$method, 
+  levels = c("raw", "mb", "mt", "tb", "fz"), 
+  labels = c("raw","Moving Breath","Moving Time","Binned Time", "Digital Filter"),
+  ordered = TRUE
+)
+exm_tidy2$value <- exm_tidy2$value / bodymass
+
+exm_tidy_cut <- exm_tidy2[exm_tidy2$time >= 500,]
+exm_tidy_cut_copy <- exm_tidy_cut[exm_tidy_cut$method == "raw", ]
+colnames(exm_tidy_cut_copy)[3] <- "timecopy"
+
+create_gif <- function(type, title, color, savename) {
+  out <- ggplot(exm_tidy_cut, aes(x = time)) +
+  geom_point(
+    data = exm_tidy_cut_copy, 
+    aes(x = timecopy, y = value, alpha = method),
+    color = "black",
+    show.legend = FALSE,
+    size = 1.5
+  ) +
+  geom_line(
+    data = exm_tidy_cut[exm_tidy_cut$method %in% c(type), ], 
+    aes(y = value, color = method), 
+    size = 1.3,
+    show.legend = FALSE,
+    color = color
+  ) +
+  scale_x_continuous(
+    name = "Time (min)",
+    limits = c(500,820),
+    breaks = c(540, 630, 720),
+    labels = c(8, 10, 12),
+    expand = c(0,0)
+  ) +
+  scale_y_continuous(
+    name = expression(paste(dot(V), O[2], " ", (ml%.%min^-1%.%kg^-1))),
+    limits = c(45,75)
+  ) +
+  scale_alpha_manual(
+    values = 0.3,
+    guide = "none"
+  ) +
+  labs(
+    title = title
+  ) +
+  theme_classic(15) +
+  theme(
+    legend.position = c(0.88, 0.63),
+    plot.title = element_text(colour = color)
+  ) +
+  gganimate::transition_reveal(time)
+  gganimate::anim_save(
+    paste0("../plots/", savename, ".gif"), out, fps = 25,
+    res = 300, width = 6.5, height = 4.5, units = "in" # passed to grDevices::png()
+  ) 
+}
+
+#create_gif("Moving Breath", "7-breath moving (Robergs & Burnett 2003)", "darkgreen", "anim1")
+#create_gif("Moving Time", "60-s moving (Howley et al. 1995)", "purple", "anim2")
+#create_gif("Digital Filter", "Digital Filter (Robergs et al. 2010)", "orange", "anim3")
+#create_gif("Binned Time", "30-s binned", "turquoise3", "anim4")
+
+# ECSS all methods suppl. plot
+
+allplot <- ggplot(exm_tidy_cut, aes(x = time)) +
+  geom_point(
+    data = exm_tidy_cut_copy, 
+    aes(x = timecopy, y = value, alpha = method),
+    color = "black",
+    show.legend = FALSE,
+    size = 1.5
+  ) +
+  geom_line(
+    data = exm_tidy_cut[exm_tidy_cut$method %in% c("Moving Breath", "Moving Time", "Digital Filter", "Binned Time"), ], 
+    aes(y = value, color = method), 
+    size = 1.3
+  ) +
+  scale_x_continuous(
+    name = "Time (min)",
+    limits = c(500,800),
+    breaks = c(540, 630, 720),
+    labels = c(8, 10, 12),
+    expand = c(0,0)
+  ) +
+  scale_y_continuous(
+    name = expression(paste(dot(V), O[2], " ", (ml%.%min^-1%.%kg^-1))),
+    limits = c(45,75)
+  ) +
+  scale_alpha_manual(
+    values = 0.3,
+    guide = "none"
+  ) +
+  scale_color_manual(
+    name = "Method",
+    values = c("darkgreen", "purple", "orange", "turquoise2")
+  ) +
+  theme_classic(15)
+
+#ggsave("../plots/allplot.png", plot = allplot, dpi = 300, width = 7.5, height = 4.5, bg = "white")
 ```
+
+## Calculate zero phase digital filter
+
+```{r}
+# Function to get VO2max with digital filtering
+get_fzmax <- function(i) {
+  dta <- spiro(files[i], weight = info$bodymass[i])
+  spiro_max(dta, smooth = "0.04fz3")$VO2_rel
+}
+
+get_fmax <- function(i) {
+  dta <- spiro(files[i], weight = info$bodymass[i])
+  spiro_max(dta, smooth = "0.04f3")$VO2_rel
+}
+
+# iterate over all tests
+fzmax <- purrr::map_dbl(seq_along(files), get_fzmax)
+fmax <- purrr::map_dbl(seq_along(files), get_fmax)
+
+# normalize to reference procedure
+fzmax_norm <- fzmax / res$tb_30
+
+# get median and quantile values
+median(fzmax_norm)
+#quant(fzmax_norm, 0.1, 0.9)
+mean(fmax)
+sd(fmax)
+```
+
+## Statistical comparison of selected strategies
+
+```{r}
+res <- read.csv("../data/results.csv")
+
+slct <- data.frame(
+  id = seq_len(nrow(res)),
+  bm_07 = res$bm_7,
+  bm_15 = res$bm_15,
+  tb_30 = res$tb_30,
+  filt = fmax,
+  tm_30 = res$tm_30,
+  tb_60 = res$tb_60
+)
+
+slct_long <- tidyr::pivot_longer(slct, cols = -id)
+
+slct_long$name <- factor(
+  slct_long$name, 
+  levels = names(sort(apply(slct[,-1],2,mean))),
+  labels = c("(f) 60-s binned", "(e) 30-s binned","(d) 30-s moving","(c) digital filter","(b) 15-breath moving","(a) 7-breath moving"))
+
+p_selcomp <- ggplot(slct_long, aes(x = value, y = name)) +
+  geom_line(aes(group = id), orientation = "y", colour = "grey80") +
+  stat_dotsinterval(
+    alpha = 0.7, 
+    .width = c(0.5, 0.8), 
+    binwidth = 0.5, dotsize = 1,
+    point_interval = "mean_qi"
+    ) +
+  scale_x_continuous(
+    name = expression(paste(dot(V), O[2][max], " ", (ml%.%min^-1%.%kg^-1))),
+    expand = c(0.02,0.02),
+    breaks = seq(50,75,5),
+    labels = c("50","","60","","70","")
+  ) +
+  scale_y_discrete(
+    name = NULL,
+    expand = c(0.05,0.1)
+  ) +
+  theme_classic(base_size = 14) +
+  theme(
+    axis.line.y = element_blank(),
+    axis.text.y = element_text(colour = "black")
+  )
+
+mod <- lmer(value ~ name + (1 | id), slct_long)
+anova(mod)
+
+summary(glht(mod, linfct = mcp(name = "Tukey")), test = adjusted(type = "single-step"))
+
+# ggsave(filename = "../plots/sel_comp.pdf", width = 8, height = 5, dpi = 600, bg = "white")
+# ggsave(filename = "../plots/sel_comp.png", width = 8, height = 5, dpi = 600, bg = "white")
+
+p_selcomp
+```

### 2023-07-27 scripts/comparison.Rmd
second revision work

@@
     breaks = c(180, 360, 540, 720, 900),
-    labels = c(2, 5, 8, 12, 15),
+    labels = c(2, 5, 8, 11, 14),
     expand = c(0,0)
@@
 
-p_all <- p_rr / p_vo2 + plot_annotation(tag_levels = 'A')
+p_all <- p_rr / p_vo2 + plot_annotation(tag_levels = 'a')
 
@@
 
-p_all_vert <- (p_rr | p_vo2) + plot_annotation(tag_levels = 'A')
+p_all_vert <- (p_rr | p_vo2) + plot_annotation(tag_levels = 'a')
 # ggsave(filename = "../plots/timecourse_vert.png", plot = p_all_vert, width = 8, height = 3, dpi = 600, bg = "white")
@@
 
+# plot with individual data
 p_selcomp <- ggplot(slct_long, aes(x = value, y = name)) +
@@
     alpha = 0.7, 
-    .width = c(0.5, 0.8), 
+    .width = c(0.33, 0.66), 
     binwidth = 0.5, dotsize = 1,
@@
 p_selcomp
+
+# calculate place for significance symbol
+get_mean_plus <- function(i) {
+  mean(slct_long$value[slct_long$name == slct_long$name[i]], 0.9) + 6.5
+}
+slct_long$place <- map_dbl(seq_len(nrow(slct_long)), get_mean_plus)
+
+get_symbol <- function(i) {
+  ifelse(
+    slct_long$name[i] %in% c("(d) 30-s moving", "(c) digital filter"),"#","*"
+  )
+}
+get_vjust <- function(i) {
+  ifelse(
+    slct_long$name[i] %in% c("(d) 30-s moving", "(c) digital filter"), 0.3 ,0.6
+  )
+}
+slct_long$symbol <- map_chr(seq_len(nrow(slct_long)), get_symbol)
+slct_long$vjust <- map_dbl(seq_len(nrow(slct_long)), get_vjust)
+# plot without individual data
+p_selcomp2 <- ggplot(slct_long, aes(x = value, y = name)) +
+  stat_pointinterval(
+    alpha = 0.7, 
+    .width = c(0.33, 0.66), 
+    point_interval = "mean_qi"
+  ) +
+  geom_text(
+    aes(x = place, label = symbol, vjust = vjust), 
+    show.legend = FALSE
+  ) +
+  scale_x_continuous(
+    name = expression(paste(dot(V), O[2][max], " ", (ml%.%min^-1%.%kg^-1))),
+    expand = c(0.02,0.02),
+    breaks = seq(52.5,72.5,2.5),
+    labels = c("","55","","60","","65","","70","")
+  ) +
+  scale_y_discrete(
+    name = NULL,
+    expand = c(0.05,0.1)
+  ) +
+  theme_classic(base_size = 14) +
+  theme(
+    axis.line.y = element_blank(),
+    axis.text.y = element_text(colour = "black")
+  )
+
+p_selcomp2
+
+# ggsave(filename = "../plots/sel_comp2.pdf", width = 7, height = 3, dpi = 600, bg = "white")
+# ggsave(filename = "../plots/sel_comp2.png", width = 7, height = 3.5, dpi = 600, bg = "white")
+
 ```
+
+## Compare impact of binning preprocessing on VO2max
+
+```{r}
+# normal time binned
+max_time_binned <- function(data, parameter, max = TRUE) {
+  nbin <- ceiling(nrow(data) / parameter)
+  ibins <- seq(from = 1, by = parameter, length.out = nbin)
+  binmean <- function(i, length, data) {
+    values <- data$VO2_rel[i:(i + length - 1)]
+    values[is.na(values)] <- 0
+    mean(values)
+  }
+  binmeans <- purrr::map_dbl(.x = ibins, .f = binmean, length = parameter, data = data)
+  if (max) max(binmeans) else binmeans
+}
+
+max_time_binned_late <- function(data, parameter, max = TRUE) {
+  raw <- attr(data, "raw")
+  last_load <- round(raw$time[max(which(raw$load != 0))])
+  nbin <- ceiling(last_load / parameter)
+  ibins <- seq(to = last_load, by = parameter, length.out = nbin)
+  binmean <- function(i, length, data) {
+    values <- data$VO2_rel[i:(i + length - 1)]
+    values[is.na(values)] <- 0
+    mean(values)
+  }
+  binmeans <- purrr::map_dbl(.x = ibins, .f = binmean, length = parameter, data = data)
+  if (max) max(binmeans) else binmeans
+}
+
+max_time_binned_noninterp <- function(data, parameter, max = TRUE) {
+  raw <- attr(data, "raw")
+  nbin <- ceiling(nrow(data) / parameter)
+  ibins <- seq(from = 0, by = parameter, length.out = nbin)
+  binmean <- function(i, length, data) {
+    raw <- attr(data, "raw")
+    values <- raw$VO2[raw$time >= ibins[i] & raw$time <= (ibins[i]+length)] / attr(data, "info")$weight
+    values[is.na(values)] <- 0
+    mean(values)
+  }
+  binmeans <- purrr::map_dbl(.x = seq_along(ibins), .f = binmean, length = parameter, data = data)
+  if (max) max(binmeans) else binmeans
+}
+
+get_vo2max <- function(n, parameter = 30) {
+  data <- spiro(files[n], weight = info$bodymass[n])
+  data.frame(
+    begin = max_time_binned(data = data, parameter = parameter),
+    end = max_time_binned_late(data = data, parameter = parameter),
+    nonint = max_time_binned_noninterp(data = data, parameter = parameter)
+  )
+}
+
+o <- purrr::map_dfr(seq_along(files), get_vo2max)
+o$dif_align <- o$begin - o$end
+o$dif_interp <- o$begin - o$nonint
+apply(o, 2, mean)
+apply(o, 2, sd)
+
+```
+
+## Plot for article: Visualizing different strategies in one raw data set
+
+```{r example, warning = FALSE}
+plot_data <- function(id, show_legend = TRUE, show_ylabel = TRUE) {
+id_info <- read.csv("../data/participants.csv")
+bodymass <- id_info$bodymass[id_info$p == id]
+
+exm_data <- spiro(files[id])
+raw_vo2 <- attr(exm_data, "raw")$VO2
+raw_time <- attr(exm_data, "raw")$time
+
+fz2 <- spiro_smooth(exm_data, "fz", "VO2")[,1]
+mb2 <- spiro_smooth(exm_data, "7b", "VO2")[,1]
+mt2 <- spiro_smooth(exm_data, 30, "VO2")[,1]
+tb2 <- rep(max_time_binned(exm_data, 60, max = FALSE) * attr(exm_data, "info")[["weight"]], each = 60)
+
+exm_tidy2 <- data.frame(
+  method = c(rep.int("raw", length(raw_vo2)), rep.int("fz", length(fz2)), rep.int("mb", length(mb2)), rep.int("mt", length(mt2)), rep.int("tb", length(tb2))),
+  value = c(raw_vo2, fz2, mb2, mt2, tb2),
+  time = c(rep.int(raw_time, 3), seq_along(mt2), seq_along(tb2))
+)
+
+exm_tidy2$method <- factor(
+  exm_tidy2$method, 
+  levels = c("raw", "mb", "mt", "tb", "fz"), 
+  labels = c("raw","7-breath moving","30-s moving","60-s binned", "digital filter"),
+  ordered = TRUE
+)
+exm_tidy2$value <- exm_tidy2$value / bodymass
+
+exm_tidy_cut <- exm_tidy2[exm_tidy2$time >= 500,]
+exm_tidy_cut_copy <- exm_tidy_cut[exm_tidy_cut$method == "raw", ]
+colnames(exm_tidy_cut_copy)[3] <- "timecopy"
+
+allplot <- ggplot(exm_tidy_cut, aes(x = time)) +
+  geom_point(
+    data = exm_tidy_cut_copy, 
+    aes(x = timecopy, y = value, alpha = method),
+    color = "black",
+    alpha = 0.15,
+    show.legend = FALSE,
+    size = 1.5
+  ) +
+  geom_line(
+    data = exm_tidy_cut[exm_tidy_cut$method %in% c("7-breath moving", "30-s moving", "digital filter", "60-s binned"), ], 
+    aes(y = value, color = method, linetype = method), 
+    size = 0.9,
+    show.legend = show_legend
+  ) +
+  scale_x_continuous(
+    name = "Time (min)",
+    limits = c(520,800),
+    breaks = c(540, 660, 780),
+    labels = c(8, 10, 12),
+    expand = c(0,0)
+  ) +
+    scale_y_continuous(
+      name = ifelse(show_ylabel, expression(paste(dot(V), O[2], " ", (ml%.%min^-1%.%kg^-1))), ""),
+      limits = c(45,80)
+    ) +
+  scale_color_manual(
+    name = "Method",
+    values = c("darkgreen", "purple", "orange", "turquoise2")
+  ) +
+  scale_linetype_manual(name = "Method", values = c("solid", "dashed", "longdash", "dotted")) +
+  theme_classic(12) +
+  theme(legend.position = "top")
+
+allplot
+}
+
+guide_area() / (plot_data(53, show_legend = FALSE) | plot_data(61, show_ylabel = FALSE)) + plot_layout(guides = 'collect', heights = c(1,6)) + plot_annotation(tag_levels = 'a') & theme(legend.position = "top")
+
+#ggsave("../plots/allplot_article.png", dpi = 300, width = 7, height = 4, bg = "white")
+#ggsave("../plots/allplot_article.pdf", dpi = 300, width = 7, height = 4, bg = "white")
+```
+
+### Difference between filter and 30-s moving average
+```{r}
+res <- read.csv("../data/results.csv")
+ba <- data.frame(
+  f = fmax,
+  m = res$tm_30
+)
+
+ba$dif <- ba$f - ba$m
+
+mean(ba$dif)
+sd(ba$dif)
+```

### 2023-08-31 paper.qmd

@@
 -   Different processing strategies lead to varying V̇O~2max~ values at the median as well as at the individual level.
--   Researchers should at minimum report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.
+-   Researchers are advised to report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.
 
@@
 Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests.
-The data were from previous research on the metabolic profile of endurance runners [@quittmann_accepted; @schwarz2022].
+The data were from previous research on the metabolic profile of endurance runners [@quittmann2022; @schwarz2022].
 The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies).
-The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_accepted] and March to October 2021 [@schwarz2022], respectively, while using identical exercise protocols and test equipment.
+The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann2022] and March to October 2021 [@schwarz2022], respectively, while using identical exercise protocols and test equipment.
 Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up.
@@
 
-The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration).
+The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @tbl-duration).
 30-second intervals were most common to define V̇O~2max~, while some authors also often employed shorter (10-20 s) and longer (60 s) periods.
@@
 #| echo: false
-#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. STILL UP TO DEBATE WHETHER RATHER PLOT (VERTICAL OR HORIZONTAL BARS?) OR TABLE."
-#| label: fig-duration
-#| out-width: '90%'
+#| tbl-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies."
+#| label: tbl-duration
 
-knitr::include_graphics("./plots/duration_count.png")
+load("./data/intervals.Rda")
+
+i_df <- as.data.frame(table(i_data$param_duration))
+
+knitr::kable(
+  i_df,
+  col.names = c("Interval duration (s)", "Count"),
+  align = "r"
+)
 ```
@@
 The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010].
-The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration).
+The length of the calculation interval for averaging is highly diverse within the literature (see @tbl-duration).
 This may reflect contradictory recommendations [@howley1995; @robergs2003].
@@
 Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison.
-The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @fig-duration); so there is no reason to compare it to other strategies.
-While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
+The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @tbl-duration); so there is no reason to compare it to other strategies.
+While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.
+
 Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies is in the same range and can bias the evaluation of their success.
+Together with biological [@katch1982] and technical [@winkert2021] variability, data processing is just one of several sources of variation in the process of V̇O~2max~ determination, but it is one that can easily be controlled without the need for multiple testing.
 
@@
 Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported.
-While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, the raw data from each test is required on the individual level.
+While the comparison of V̇O~2max~ from different processing strategies require their reporting for a sufficient analysis on a group level, the raw data from each test is required on the individual level.
 
@@
 These items may be reported in form of a checklist, as an in-text enumeration or in sentence form.
-An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_accepted; @schwarz2022] would be:
+An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann2022; @schwarz2022] would be:
 
@@
 
-DISCUSS IF WE SHOULD CHANGE THE ORDER OF RECOMMENDATIONS/GUIDELINES AND LIMITATIONS.
+# Summary
+
+Despite calls for standardization, researchers use a variety of data processing strategies to determine the V̇O~2max~ from raw gas exchange data.
+Based on current reporting practices we developed a checklist that can serve as a guideline for reporting data processing methods for V̇O~2max~ determination.
+We advise authors to follow the reporting guidelines and ideally share anonymized raw data to improve the reproducibility of research in exercise physiology.
 

### 2023-08-31 scripts/review.Rmd
Adjust to next feedback round

@@
 # Show data processing type for breath-by-breath measurements only
+ext$total_comb <- paste(ext$proc_combination, ext$parameters, sep = "_")
+table(ext$total_comb[ext$type == "bbb"], useNA = "ifany")
+
+## Show processing type X interval combinations
+
 table(ext$proc_combination[ext$type == "bbb"], useNA = "ifany")
@@
 knitr::include_graphics("../plots/duration_count.png")
+
+# save data for table generation as alternative to a plot
+i_data <- ext[ext$ptype == "time" & !is.na(ext$ptype), ]
+save(i_data, file = "../data/intervals.Rda")
 ```
